# Additive phenotype models {#sec-additive}

The first chapter simulated one additive trait with default settings. This
chapter opens the additive layer up: how the causal markers are chosen, how
their effects are generated, how the layer is scaled to its requested share of
the phenotypic variance, and what the per-marker `var_explained` diagnostic does
and does not mean. It also shows the options of `simulate_phenotype()` that
apply to every later chapter: replications, a phenotype mean, a subset of
individuals, entry means, and the one-call form.

[]{#additive-setup}

```{r}
#| label: additive-setup
#| include: false
source("_setup.R")
source("_data.R")
```

## Worked example {#sec-additive-example}

### Load the package and the maize genotypes

[]{#additive-library}

```{r}
#| label: additive-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)

data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
```

The data are the inbred maize panel used in [A first simulation](first-simulation.qmd).
The genotype-input chapter explains the format and its quality checks.

### A single additive layer and its causal markers

`simulate_phenotype()` fixes the data, the seed, and the global settings;
`additive()` adds one additive component. `prop` is the share of the
phenotypic variance that the layer receives, and `n_qtn` is the number of
causal markers, which the package calls QTNs (quantitative trait nucleotides).

[]{#additive-model}

```{r}
#| label: additive-model
sim <- simulate_phenotype(geno, seed = 11) |>
  additive(prop = 0.6, n_qtn = 10)
print(sim)

qt <- qtn_table(sim)
qt |>
  select(snp, chr, pos, maf, effect, var_explained)
```

`qtn_table()` lists the causal markers in the order they were drawn, together
with the minor-allele frequency (`maf`), the effect assigned to each marker,
and `var_explained`. The effects follow a geometric series: the first marker's
effect is 0.5, the second 0.25, and so on. The markers are a simple random
sample of the markers that vary in the panel.

### Effect series: geometric, custom base, or explicit effects

The `effect` argument changes the series. A single value is the base of the
geometric series; a vector of length `n_qtn` is used as given. Equal effects
give a trait where no single marker dominates.

[]{#additive-effects}

```{r}
#| label: additive-effects
base_09 <- simulate_phenotype(geno, seed = 11) |>
  additive(prop = 0.6, n_qtn = 10, effect = 0.9)

equal <- simulate_phenotype(geno, seed = 11) |>
  additive(prop = 0.6, n_qtn = 10, effect = rep(1, 10))

tibble(
  marker = seq_len(10),
  default = qtn_table(sim)$effect,
  base_0.9 = qtn_table(base_09)$effect,
  equal = qtn_table(equal)$effect
)
```

The three simulations share the same seed and therefore the same ten markers;
only the effects differ. Because the layer is rescaled to `prop` afterwards,
the absolute size of the effects does not matter. Only their relative sizes do.
Multiplying every effect by the same constant leaves the phenotype unchanged.

`phase = "repulsion"` alternates the sign of successive effects in draw order.
It is a bookkeeping convenience, not a phase read from the haplotypes: for
unlinked markers it merely relabels which allele increases the trait. The
[multi-trait chapter](multitrait.qmd) shows the haplotype-derived phase of the
`"ld"` architecture.

[]{#additive-phase}

```{r}
#| label: additive-phase
repulsion <- simulate_phenotype(geno, seed = 11) |>
  additive(prop = 0.6, n_qtn = 10, phase = "repulsion")

tibble(
  coupling = qtn_table(sim)$effect,
  repulsion = qtn_table(repulsion)$effect
)
```

### Fix the causal markers yourself

`qtn` accepts marker names or column indices. This is the way to simulate a
trait on loci chosen for a reason, for example genes from a known pathway or
markers at a chosen allele frequency. `n_qtn` is then taken from the length
of `qtn`.

[]{#additive-fixed-qtn}

```{r}
#| label: additive-fixed-qtn
chosen <- geno$snp[c(10, 2000, 5000)]
fixed <- simulate_phenotype(geno, seed = 11) |>
  additive(prop = 0.6, qtn = chosen)

qtn_table(fixed) |>
  select(snp, chr, maf, effect, var_explained)
```

### The one-call form

When `h2` and `n_qtn` are both given to `simulate_phenotype()`, the package
builds the implied model directly. For a single additive layer this is the
same simulation as the pipe, and the phenotypes are identical.

[]{#additive-one-call}

```{r}
#| label: additive-one-call
one_call <- simulate_phenotype(geno, h2 = 0.6, n_qtn = 10, seed = 11)

identical(phenotypes_wide(one_call), phenotypes_wide(sim))
```

### Replications with the same or with new architectures

`n_reps` requests simulation replications. By default every replication keeps
the same causal markers and effects and draws new residuals. With
`vary_qtn = TRUE` each replication draws its own markers and effects, so the
replications are different genetic architectures. `qtn_table(rep = )` and
`genetic_values(rep = )` retrieve a particular replication.

[]{#additive-reps}

```{r}
#| label: additive-reps
same_qtn <- simulate_phenotype(geno, seed = 11, n_reps = 3) |>
  additive(prop = 0.6, n_qtn = 10)
new_qtn <- simulate_phenotype(geno, seed = 11, n_reps = 3, vary_qtn = TRUE) |>
  additive(prop = 0.6, n_qtn = 10)

tibble(
  replication = 1:3,
  same_qtn = sapply(1:3, function(r) qtn_table(same_qtn, rep = r)$snp[1]),
  new_qtn = sapply(1:3, function(r) qtn_table(new_qtn, rep = r)$snp[1])
)

phenotypes_long(same_qtn) |>
  count(rep)
```

The first table shows the first causal marker of each replication: the same
marker under the default, three different markers with `vary_qtn = TRUE`.
The second confirms that each replication phenotypes every individual.

### Mean, subset of individuals, and entry means

`mean` shifts the phenotype without touching the genetic values.
`individuals` restricts the simulation to a subset; allele frequencies are
recomputed on that subset. `reps` is different from `n_reps`: it represents
each phenotype as the average of `reps` independent records, which divides
the residual by the square root of `reps`. The notation chapter
([conventions](conventions.qmd)) distinguishes the two kinds of replication.

[]{#additive-mean-subset}

```{r}
#| label: additive-mean-subset
subset_sim <- simulate_phenotype(geno, seed = 11, mean = 100, individuals = 1:100) |>
  additive(prop = 0.6, n_qtn = 10)

entry_means <- simulate_phenotype(geno, seed = 11, reps = 4) |>
  additive(prop = 0.6, n_qtn = 10)

tibble(
  simulation = c("mean = 100, 100 individuals", "reps = 4, all individuals"),
  individuals = c(nrow(phenotypes_long(subset_sim)), nrow(phenotypes_long(entry_means))),
  phenotype_mean = round(c(mean(phenotypes_long(subset_sim)$value),
                           mean(phenotypes_long(entry_means)$value)), 3),
  realized_ratio = c(
    var(genetic_values(subset_sim)[, 1]) / var(phenotypes_wide(subset_sim)$Trait_1),
    var(genetic_values(entry_means)[, 1]) / var(phenotypes_wide(entry_means)$Trait_1)
  )
)
```

For the entry-mean simulation the realized ratio is far above the requested
0.6. That is expected: averaging four records shrinks the residual variance
to a quarter, so the entry-mean heritability is larger than the single-record
value. The `print()` method reports both. `prop` and `h2` always refer to the
single-record scale.

### Variance explained per marker

`var_explained` is the marginal share of the realized phenotypic variance
attributable to one marker: the variance of that marker's scaled contribution
divided by the phenotypic variance. The shares need not add up to the layer's
`prop`, because contributions of different markers covary when the markers are
in linkage disequilibrium or, in a finite sample, by chance.

[]{#additive-var-explained}

```{r}
#| label: additive-var-explained
tibble(
  quantity = c("Requested additive share", "Sum of var_explained",
               "Realized genetic / phenotypic variance"),
  value = c(
    0.6,
    sum(qtn_table(sim)$var_explained),
    var(genetic_values(sim)[, 1]) / var(phenotypes_wide(sim)$Trait_1)
  )
)
```

### Equal effects on an outbred animal population

The maize panel was filtered to minor-allele frequencies above 0.4, so every
marker is nearly as variable as a marker can be. The heterogeneous-stock mice
(see the [genotype-input chapter](input.qmd)) span the full frequency range.
Giving ten mouse markers the same effect isolates the role of allele
frequency: differences in `var_explained` then come from the genotypes alone.

[]{#additive-mice}

```{r}
#| label: additive-mice
mice <- book_mice()

mice_sim <- simulate_phenotype(mice, seed = 11) |>
  additive(prop = 0.6, n_qtn = 10, effect = rep(1, 10))

qtn_table(mice_sim) |>
  mutate(hw_variance = 2 * maf * (1 - maf)) |>
  select(snp, maf, effect, var_explained, hw_variance) |>
  arrange(desc(var_explained))
```

The markers with the highest minor-allele frequency explain the largest
shares, and the ordering broadly follows `hw_variance`, the variance of the
genotype code expected under Hardy-Weinberg proportions. Departures from
that ordering reflect the realized genotype counts, which need not be in
Hardy-Weinberg proportions in this structured population. Fewer individuals carry a rare
allele, so a rare marker's contribution has a smaller variance even when its
effect is the same. The theory section gives the formula.

[]{#additive-plot}

```{r}
#| label: additive-plot
#| fig-height: 6
#| fig-cap: "Diagnostic plots for the mouse simulation. The effect panel shows the geometric series; the variance panel shows the requested and realized shares."
plot(mice_sim)
```

## Assumptions and theory

### The additive score

Write $x_{ij}$ for the coded genotype of individual $i$ at causal marker $j$,
with values `-1`, `0`, `1`, and $b_j$ for the marker's effect. The raw additive
score is the weighted sum

$$
c_i = \sum_{j=1}^{m} x_{ij}\, b_j ,
$$ {#eq-additive-score}

and the layer's genetic value is the centered and rescaled score
$g_i = \sqrt{\pi}\,(c_i - \bar c)/s_c$, as derived in @sec-scaling-derivation.
Here $\pi$ is `prop` and $s_c$ is the sample standard deviation of the scores.
Both steps are package definitions, implemented in `.component_raw()` and
`.genetic_matrix()`. The sample variance of $g$ equals $\pi$ exactly, whatever
the effects and whatever the allele frequencies.

The coding counts one allele and is symmetric around the heterozygote; the
model is linear in the number of copies, so it has no dominance. This is the
v1 convention of the original publication (Fernandes and Lipka, 2020), not
Fisher's average-effect decomposition. For a single additive layer on a
population in Hardy-Weinberg proportions the requested `prop` is the
narrow-sense heritability of the simulated trait, because the whole genetic
value is then additive.

### Geometric effect series

The default effect of the $k$-th drawn marker is

$$
b_k = \beta^{\,k}, \qquad k = 1, \dots, m, \qquad \beta = 0.5 ,
$$ {#eq-geometric-series}

implemented in `.effect_series()`. The base $\beta$ is the scalar `effect`;
a base near one gives nearly equal effects and a small base concentrates the
trait on the first few markers. With $\beta = 0.5$ the tenth marker has an
effect about a thousand times smaller than the first and is causal in name
only. An explicit vector replaces the series. Because of the rescaling in the
previous section, only ratios of effects matter.

Under `phase = "repulsion"` the series is multiplied by $(-1)^{k+1}$ in draw
order. The package documents this as positional alternation, not a phase
derived from linkage.

### Why the variance explained depends on allele frequency

For a marker in Hardy-Weinberg proportions with allele frequency $p$ of the
counted allele, the variance of the `-1`, `0`, `1` genotype code is $2p(1-p)$.
This is the variance of a binomial count with two trials, shifted by a constant.
The variance of the marker's contribution $x_{ij} b_j$ is therefore
$2p(1-p)\,b_j^2$ in expectation. Two markers with the same effect can explain
different shares of variance if their frequencies differ, and a rare allele
with a large effect can explain less than a common allele with a smaller one.

In the maize panel every marker has a minor-allele frequency above 0.4, so
$2p(1-p)$ is close to its maximum of 0.5 and the ranking of `var_explained`
follows the ranking of effects. In the mouse example the effects are equal, so
the ranking follows $2p(1-p)$ instead.

The package computes `var_explained` from the realized columns rather than
from this expectation. Its definition (package convention, `.qtn_var()`) is
the variance of the marker's scaled column divided by the realized phenotypic
variance, so it inherits the finite-sample frequencies and any departure from
Hardy-Weinberg proportions. Inbred lines are an extreme departure: their
genotype codes take almost only the values `-1` and `1`, and the code variance
at frequency $p$ is close to $4p(1-p)$ rather than $2p(1-p)$. The rescaling to
`prop` absorbs this for the layer as a whole.

### Why the shares do not add up

Expanding the variance of the sum in @eq-additive-score gives

$$
\operatorname{Var}(c) = \sum_j b_j^2 \operatorname{Var}(x_j)
  + 2 \sum_{j<k} b_j b_k \operatorname{Cov}(x_j, x_k).
$$ {#eq-additive-sum-variance}

Marginal shares report only the first sum. The covariance terms are nonzero
when causal markers are in linkage disequilibrium, and in a finite sample they
are nonzero by chance even for unlinked markers. The table in the worked
example therefore shows a sum of `var_explained` that differs from 0.6. This is
an algebraic consequence of the definitions, not a simulation error. For the
same reason the sum can exceed `prop` when the covariances are negative, for
example under `phase = "repulsion"` with linked markers.

### Seeds and reproducibility

With a non-missing `seed`, each layer draws its markers and effects under a
sub-seed derived from the seed, the layer type, and the number of earlier
layers of the same type (package scheme, `.layer_seed()`). Adding a dominance
layer after an additive layer does not change the additive draw, which is why
the chapters that follow can build on the simulations of this one. Inserting a
second additive layer before an existing one does change the second one's
draw. The caller's random-number state is left untouched.

## Implementation and practical limits

The additive layer is pure R: marker sampling, effect generation, scaling, and
residual draws all stay in R so that seeded results are reproducible across
platforms. The internal functions named above are listed in the
[developer index](developer-index.qmd).

- `n_qtn` must not exceed the number of markers that vary in the data.
  Monomorphic markers and markers heterozygous in every individual are never
  candidates.
- A marker fixed with `qtn` that does not vary is accepted with a warning and
  contributes nothing.
- `var_explained` is a diagnostic of one simulated sample. It is not an
  estimate and does not predict what an association study would detect.
- `h2` in `simulate_phenotype()` is the budget for additive, dominance, and
  epistatic layers together. With a single additive layer the layer may omit
  `prop` and take the whole budget.
- `reps` scales only the residual; the genetic value is unchanged. It does not
  model repeated measurements with a shared permanent environment.

::: {.callout-tip title="Choosing the number of causal markers"}
A geometric series with base 0.5 makes the trait effectively oligogenic
however large `n_qtn` is. For a polygenic trait raise the base toward one or
give an explicit effect vector, and check the resulting `var_explained`
column before trusting the architecture.
:::
