# Binary phenotypes with human genotypes {#sec-binary}

Suppose you have human genotypes and want to simulate a disease that affects
about one person in ten, with half of the variation in disease risk explained
by 20 causal variants. The observed phenotype is a case/control label, but the
genetic model needs a continuous quantity to act on. This chapter simulates the
disease under the liability-threshold model, checks the realized prevalence
against the requested one, separates the liability scale from the observed
scale, and shows what a case/control sample does and does not preserve.

[]{#binary-setup}

```{r}
#| label: binary-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-binary-example}

### Load the package and the human genotypes

[]{#binary-library}

```{r}
#| label: binary-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

The genotypes are the HapMap Phase II subset distributed with the SNPRelate
package: 279 individuals from four populations (CEU, YRI, HCB, JPT) and about
nine thousand SNPs after linkage-disequilibrium pruning. The
[genotype input chapter](input.qmd) explains how this file is read, why 60
markers with missing or duplicated identifiers are dropped, and why missing
calls are imputed to the heterozygote. Here we reuse those steps through the
book's shared loader.

[]{#binary-genotypes}

```{r}
#| label: binary-genotypes
source("_data.R")
human <- book_human()
samples <- book_human_samples()

tibble(markers = nrow(human), individuals = ncol(human) - 5)
count(samples, population)
```

Markers occupy rows, as in the [first chapter](first-simulation.qmd). The
population labels come with the file and are used below only to split the
simulated cases by group.

### Simulate the disease

We request an additive genetic share of 0.5 on 20 causal markers and then
convert the continuous phenotype into two categories with expected proportions
0.9 and 0.1. The second category is the disease.

[]{#binary-model}

```{r}
#| label: binary-model
disease <- simulate_phenotype(human, seed = 2024) |>
  additive(prop = 0.5, n_qtn = 20) |>
  liability_threshold(prop = c(0.9, 0.1))
print(disease)
```

The printout keeps the variance partition and the realized genetic ratio on the
liability scale, and says so. `liability_threshold()` changes only the phenotype
column: the values are now category labels `1` (unaffected) and `2` (affected).

[]{#binary-tables}

```{r}
#| label: binary-tables
status <- phenotypes_long(disease) |>
  mutate(case = value == 2)

count(status, value, case)
```

The continuous phenotype that produced these labels is retained as the
liability. It is stored in the simulation object, with the same columns as the
phenotype table, and is what `genetic_values()` and `coheritability()` continue
to use.

[]{#binary-liability}

```{r}
#| label: binary-liability
liability <- disease$liability |>
  rename(liability = value)

genetic <- genetic_values(disease) |>
  as_tibble(rownames = "id") |>
  rename(genetic_value = Trait_1)

values <- status |>
  left_join(liability, by = c("id", "trait", "rep")) |>
  left_join(genetic, by = "id")

values |>
  group_by(case) |>
  summarise(
    n = n(),
    lowest_liability = min(liability),
    highest_liability = max(liability),
    mean_genetic_value = mean(genetic_value)
  )
```

Every case has a higher liability than every control, because the label is a
deterministic cut of the liability. The cases also have a higher mean genetic
value, but the two groups overlap in genetic value: the residual half of the
liability moves some individuals with low genetic risk above the threshold
and some with high genetic risk below it.

### Where the threshold is placed

The package standardizes the liability within the simulated population and
places the cut point at the normal quantile of the cumulative proportion. For
two categories with `prop = c(0.9, 0.1)` the cut is the sample mean plus the
sample standard deviation times the 0.9 quantile of the standard normal
distribution. We reproduce it and compare it with the boundary between the
observed groups.

[]{#binary-threshold}

```{r}
#| label: binary-threshold
cut_point <- mean(values$liability) + sd(values$liability) * qnorm(0.9)

tibble(
  quantity = c(
    "Threshold (package rule)", "Highest control liability",
    "Lowest case liability", "Requested prevalence", "Realized prevalence"
  ),
  value = c(
    cut_point,
    max(values$liability[!values$case]),
    min(values$liability[values$case]),
    0.1,
    mean(values$case)
  )
)
```

The threshold lies between the highest control and the lowest case. The
realized prevalence is the fraction of liabilities above that cut and is not
forced to equal 0.1. A sample-quantile rule would have produced exactly 28
cases in 279 individuals; the normal-quantile rule does not. We return to this
distinction in the theory section.

### Realized prevalence across replications

To see how much the realized prevalence varies, we repeat the residual draw
200 times with `n_reps`. The causal markers and effects stay the same across
replications; only the residual, and therefore the liability and the labels,
change.

[]{#binary-replications}

```{r}
#| label: binary-replications
replicated <- simulate_phenotype(human, seed = 2024, n_reps = 200) |>
  additive(prop = 0.5, n_qtn = 20) |>
  liability_threshold(prop = c(0.9, 0.1))

prevalence <- phenotypes_long(replicated) |>
  group_by(rep) |>
  summarise(prevalence = mean(value == 2))

tibble(
  quantity = c("Mean prevalence", "Lowest", "Highest", "Standard deviation"),
  parameter = c(0.1, NA, NA, NA),
  simulated = c(
    mean(prevalence$prevalence), min(prevalence$prevalence),
    max(prevalence$prevalence), sd(prevalence$prevalence)
  )
)
```

The mean over replications is close to the requested value, and individual
replications scatter around it. The spread is smaller than the binomial
standard deviation for 279 independent Bernoulli trials with probability 0.1,
for three reasons. The genetic values are the same in every replication, so
each individual has its own case probability rather than a common 0.1. The
default fixed residual mode holds the residual sample variance constant. And
the threshold is recomputed from each replication's sample mean and standard
deviation. Each removes part of the sampling variation. The
replication count is a simulation design choice; it does not change the
genetic model.

### Cases by population

The four HapMap populations differ in allele frequencies at the causal markers,
so they differ in mean genetic value and therefore in realized prevalence.

[]{#binary-populations}

```{r}
#| label: binary-populations
values |>
  left_join(samples, by = "id") |>
  group_by(population) |>
  summarise(
    n = n(),
    cases = sum(case),
    prevalence = mean(case),
    mean_genetic_value = mean(genetic_value)
  )
```

This difference is an outcome of the simulation, not an input. The threshold
is one value for the whole sample, and the genetic effects are the same for
every individual. Groups with a higher mean genetic value at the sampled
causal markers end up with more cases. With a different seed the causal markers
change and so does the ranking of the groups. A simulation of this kind cannot
tell you anything about real disease differences between populations; it tells
you how allele-frequency differences at hypothetical causal loci translate
into prevalence differences under one model.

### Random residual mode

The default fixed residual mode scales the residual to exactly its requested
sample variance. With `residual_mode = "random"` the residual is an ordinary
normal draw, so the liability's sample variance fluctuates. The threshold rule
is unchanged: it standardizes whatever liability it receives.

[]{#binary-random}

```{r}
#| label: binary-random
random_disease <- simulate_phenotype(human, seed = 2024,
                                     residual_mode = "random") |>
  additive(prop = 0.5, n_qtn = 20) |>
  liability_threshold(prop = c(0.9, 0.1))

tibble(
  mode = c("fixed", "random"),
  liability_variance = c(var(disease$liability$value),
                         var(random_disease$liability$value)),
  cases = c(sum(disease$pheno$value == 2),
            sum(random_disease$pheno$value == 2))
)
```

The genetic values are identical in both modes because the seed and the
additive layer are the same; only the residual generation differs. For
simulation studies of case/control estimators, random mode corresponds to the
usual model of independent normal residuals on the liability scale.

### A quantitative trait and a disease together

Several traits can share causal loci while only some of them are thresholded.
Here two traits are simulated under the pleiotropy architecture with a genetic
correlation target of 0.5, and only the second trait is made binary with a
prevalence target of 0.2.

[]{#binary-two-traits}

```{r}
#| label: binary-two-traits
pair <- simulate_phenotype(human, seed = 7, n_traits = 2,
                           architecture = "pleiotropy", cor = 0.5) |>
  additive(prop = c(0.4, 0.6), n_qtn = 20) |>
  liability_threshold(prop = c(0.8, 0.2), trait = 2)
print(pair)

phenotypes_wide(pair) |>
  head()

coheritability(pair)
```

`Trait_1` remains continuous and `Trait_2` is a category label. The
co-heritability matrix is computed from the genetic values and the liability,
so it stays interpretable for the binary trait. The
[multiple-trait chapter](multitrait.qmd) discusses the pleiotropy architecture
and its correlation target.

### More than two categories

`prop` may list more than two proportions. The categories are ordered, and the
cut points are the normal quantiles of the cumulative proportions.

[]{#binary-ordinal}

```{r}
#| label: binary-ordinal
severity <- simulate_phenotype(human, seed = 2024) |>
  additive(prop = 0.5, n_qtn = 20) |>
  liability_threshold(prop = c(0.25, 0.5, 0.25))

count(phenotypes_long(severity), value)
```

The liability is the same as in the first simulation, because the seed and the
layer are the same. Only the number and position of the cut points differ.

### Sampling cases and controls

Studies rarely observe the whole population. A common design samples a fixed
number of cases and the same number of controls. We take 25 of each from the
simulated population and compare what the sample shows with what the
population showed.

[]{#binary-ascertainment}

```{r}
#| label: binary-ascertainment
set.seed(1)
sampled <- values |>
  group_by(case) |>
  slice_sample(n = 25) |>
  ungroup()

tibble(
  quantity = c("Prevalence", "Genetic variance", "Liability variance"),
  population = c(mean(values$case), var(values$genetic_value),
                 var(values$liability)),
  case_control_sample = c(mean(sampled$case), var(sampled$genetic_value),
                          var(sampled$liability))
)
```

The sample's prevalence is 0.5 by construction. Its genetic and liability
variances are also different from the population's, because selecting on the
label selects on the liability. Any heritability computed from this sample
refers to the sample's own design, not to the population. The simulation
provides the population truth, so the effect of the design can be measured
rather than assumed.

## Assumptions and theory

This example uses one continuous liability built from an additive layer and a
normal residual, a single threshold rule, and no selection, crossing, or
ascertainment inside the simulation itself. The case/control sampling above is
performed on the output, not by the package.

### The liability-threshold model

The model postulates an unobserved continuous variable, the liability $l_i$,
which in this chapter is the ordinary simulated phenotype

$$
l_i = g_i + e_i ,
$$

with $g_i$ the genetic value scaled to sample variance $\pi$ and $e_i$ the
residual. The observed category is determined by fixed cut points
$t_1 < t_2 < \dots < t_{C-1}$:

$$
y_i = 1 \text{ if } l_i < t_1, \qquad
y_i = k \text{ if } t_{k-1} \le l_i < t_k, \qquad
y_i = C \text{ if } l_i \ge t_{C-1} .
$$ {#eq-threshold-rule}

The package documentation attributes the model to Wright (1934) and Falconer
(1965), as cited in `?liability_threshold`. The construction of $l_i$ is the
same package definition as in @sec-scaling-derivation; the threshold rule is
a package definition implemented by `.apply_threshold()`.

### How the cut points are chosen

Given category proportions $p_1, \dots, p_C$ summing to one, the package sets

$$
t_k = \bar l + s_l \, \Phi^{-1}\!\left(\sum_{j=1}^{k} p_j\right),
\qquad k = 1, \dots, C-1,
$$ {#eq-threshold-cut}

where $\bar l$ and $s_l$ are the sample mean and sample standard deviation of
the liability in the simulated population, and $\Phi^{-1}$ is the standard
normal quantile function. This is a package definition, implemented by
`.threshold_cuts()`. It is applied separately for each thresholded trait and
each replication.

Two consequences follow directly. First, the realized category frequencies
match the requested proportions only in expectation, and only when the
liability is normal. A finite sample is never exactly normal, and 279 times 0.1
is not even a whole number, so the realized prevalence fluctuates, as the
replication table showed. Second, the rule is
not a sample-quantile rule. A sample-quantile rule would fix the number of
cases at the nearest integer to $n p_C$ in every replication; the package does
not do this, so counts vary from replication to replication. When a design
needs an exact case count, sample it from the output as in the ascertainment
example.

A model whose liability is deliberately non-normal, for example one with a
large-effect locus, a vQTL component, or strong epistasis, can produce a
realized prevalence that differs systematically from `prop`, not only by
sampling. The requested proportion is then a normal-approximation target, and
the realized count is the quantity to report.

### Liability scale and observed scale

All variance quantities reported by the package refer to the liability. The
genetic share $\pi$, the realized genetic ratio, the per-marker variance
explained, and the co-heritability are computed from $g_i$ and $l_i$ by
`.realized_h2()` and `coheritability()` through the stored liability table,
not from the labels. This is a package definition.

The labels themselves have a different variance. For a binary trait coded 0/1
with prevalence $K$, the sample variance of the labels is approximately
$K(1-K)$, which for $K = 0.1$ is about 0.09, while the liability has variance
near one. A ratio of genetic variance to label variance therefore has no
interpretation as a heritability on either scale; the cases table above
illustrates that the genetic values overlap between groups while the labels do
not. Formulas that convert a heritability estimated on the observed 0/1 scale
to the liability scale exist in the literature, but they are not implemented
in the package and are not reproduced here; the equation map of this package
records no verified source for them.

### Why sampling cases and controls changes the variances

Selecting individuals by their label is selection on the liability: every case
has $l_i \ge t_1$ and every control has $l_i < t_1$. In a case/control sample
with equal group sizes, the liability distribution is a mixture of the upper
and lower tails in proportions that differ from the population's. Its variance
and the genetic variance within it are therefore not the population values,
and a heritability computed from such a sample does not estimate the
population liability heritability without a correction for the sampling
design. This is an algebraic consequence of selection on the output, not a
package definition.

## Implementation and practical limits

`liability_threshold()` cuts the already realized liability; it does not
redraw the residual. Layers added afterwards re-realize the phenotype and
re-apply the stored thresholds. The relevant internal functions are
`.apply_threshold()`, `.threshold_cuts()`, and `.liability_table()`; all are R
functions, so the threshold step involves no Rust code.

- The threshold is a function of the simulated sample. A population carrying a
  stored trait keeps its base population's absolute cut points, so category
  frequencies can drift across generations; see the
  [trait values across generations chapter](trait.qmd).
- A combined model from `complex_phenotypes()` cannot be thresholded.
  `liability_threshold()` refuses the combined object, and a threshold on an
  input is discarded together with that input's phenotype. Build binary traits
  with layers on a single `simulate_phenotype()` object instead; see the
  [complex architectures chapter](complex.qmd).
- Selection on the phenotype of a thresholded trait ranks category labels, with
  many ties. Selection on genetic values or liability-scale breeding values
  avoids the ties; see the [selection chapter](selection.qmd).
- The HapMap subset is small and pruned for linkage disequilibrium, which makes
  it convenient for examples but unrepresentative of a genome-wide association
  design. Prevalence differences between its populations in a simulation are
  artefacts of which markers were sampled as causal.
