# Complex phenotype architectures {#sec-complex}

Two kinds of complexity are often needed beyond additive, dominance, and
epistatic mean effects. The first is a locus that changes how *variable* a
trait is rather than its mean: a variance QTL. The second is a model that no
single architecture describes, for example two traits that share some causal
loci but not others. This chapter builds both, checks what the simulation
produces against what was requested, and explains why a vQTL is counted as
residual rather than genetic variance.

[]{#complex-setup}

```{r}
#| label: complex-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-complex-example}

### Load the package and the data

[]{#complex-library}

```{r}
#| label: complex-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

A variance QTL is easiest to see when all three genotype classes are present,
so the first part uses the human HapMap genotypes introduced in the
[binary chapter](binary.qmd) and the [input chapter](input.qmd). The second
part combines two architectures on the bundled maize panel.

[]{#complex-genotypes}

```{r}
#| label: complex-genotypes
source("_data.R")
human <- book_human()

data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
maize <- as_numeric(SNP55K_maize282_maf04,
                    to_r = TRUE, to_file = FALSE, verbose = FALSE)
```

### Choose a locus whose variance effect can be seen

A vQTL drawn at random may land on a marker with few individuals in one
genotype class, which makes its effect hard to display. We therefore pick a
marker on which every class has at least 60 individuals and pass it to
`vqtl()` through `qtn`. The additive layer is still drawn at random.

[]{#complex-choose}

```{r}
#| label: complex-choose
class_counts <- human |>
  select(-(1:5)) |>
  apply(1, function(x) table(factor(x, levels = c(-1, 0, 1)))) |>
  t() |>
  as_tibble() |>
  mutate(snp = human$snp, .before = 1)

candidate <- class_counts |>
  filter(`-1` >= 60, `0` >= 60, `1` >= 60) |>
  slice(1)
candidate
```

### A variance QTL on its own locus

We request an additive share of 0.3 on ten markers and a vQTL share of 0.3 on
the chosen marker, leaving 0.4 for the ordinary residual.

[]{#complex-vqtl}

```{r}
#| label: complex-vqtl
sim <- simulate_phenotype(human, seed = 31) |>
  additive(prop = 0.3, n_qtn = 10) |>
  vqtl(prop = 0.3, qtn = candidate$snp)
print(sim)
```

The printout lists the vQTL as a separate row of the variance partition and
states that it is not counted in the broad-sense heritability. The requested
genetic share is 0.3, the additive layer alone. `qtn_table()` lists the vQTL
locus with its layer effect and leaves its `var_explained` empty: a vQTL
explains no variance of the mean.

[]{#complex-vqtl-table}

```{r}
#| label: complex-vqtl-table
qtn_table(sim) |>
  filter(layer == "vqtl") |>
  select(trait, layer, snp, chr, maf, effect, var_explained)
```

### Residual variance by genotype class

Because the phenotype is genetic value plus residual, subtracting the genetic
value recovers the residual. We attach each individual's genotype at the vQTL
and compare the residual variance within the three classes.

[]{#complex-classes}

```{r}
#| label: complex-classes
genotype_at <- function(geno, marker) {
  row <- geno[geno$snp == marker, ]
  tibble(id = names(row)[-(1:5)], genotype = as.numeric(row[-(1:5)]))
}

residuals_of <- function(sim, rep = 1) {
  genetic <- genetic_values(sim, rep = rep) |>
    as_tibble(rownames = "id") |>
    rename(genetic_value = Trait_1)
  phenotypes_long(sim) |>
    filter(rep == !!rep) |>
    left_join(genetic, by = "id") |>
    mutate(residual = value - genetic_value)
}

by_class <- residuals_of(sim) |>
  left_join(genotype_at(human, candidate$snp), by = "id") |>
  group_by(genotype) |>
  summarise(
    n = n(),
    mean_residual = mean(residual),
    residual_variance = var(residual),
    mean_genetic_value = mean(genetic_value)
  )
by_class
```

The residual variance increases from one homozygote through the heterozygote
to the other homozygote, while the mean residual stays near zero in every
class. The mean genetic value also differs somewhat between classes. That
difference does not come from the vQTL, which adds no mean effect; it comes
from the sample association between this marker and the ten additive loci.
In a panel drawn from four populations with different allele frequencies such
associations are common. We have created a locus that affects the spread of
the residual without affecting its average.

### Compare with a model without the vQTL

The same seed and additive layer without the variance component give a
reference whose residual variance does not depend on the genotype at that
marker. Any differences left between classes are sampling variation of a
variance computed from 60 to 143 individuals; the replicated comparison in the
next section separates such noise from the modelled pattern.

[]{#complex-reference}

```{r}
#| label: complex-reference
plain <- simulate_phenotype(human, seed = 31) |>
  additive(prop = 0.3, n_qtn = 10)

residuals_of(plain) |>
  left_join(genotype_at(human, candidate$snp), by = "id") |>
  group_by(genotype) |>
  summarise(n = n(), residual_variance = var(residual))
```

### Replications and the expected pattern

One realization can show a class difference by chance. With `n_reps` the
residual is redrawn many times while the loci and effects stay the same. The
log-linear link described in the theory section predicts that the conditional
variances of the three classes are in geometric progression, with the ratio
between adjacent classes set by the vQTL share and by the marker's
standardized dosage.

[]{#complex-replications}

```{r}
#| label: complex-replications
replicated <- simulate_phenotype(human, seed = 31, n_reps = 200) |>
  additive(prop = 0.3, n_qtn = 10) |>
  vqtl(prop = 0.3, qtn = candidate$snp)

dosage <- genotype_at(human, candidate$snp)
loading <- dosage |>
  mutate(loading = sqrt(0.3) * (genotype - mean(genotype)) / sd(genotype)) |>
  distinct(genotype, loading)

class_variances <- lapply(seq_len(200), function(r) {
  residuals_of(replicated, rep = r) |>
    left_join(dosage, by = "id") |>
    group_by(genotype) |>
    summarise(variance = var(residual), .groups = "drop") |>
    mutate(rep = r)
}) |>
  bind_rows()

class_variances |>
  group_by(genotype) |>
  summarise(mean_variance = mean(variance), sd_variance = sd(variance)) |>
  left_join(loading, by = "genotype") |>
  mutate(relative_factor = exp(loading) / exp(loading[genotype == 0]))
```

The `relative_factor` column is the conditional variance of the heterogeneous
component relative to the heterozygote class, before the component is
rescaled and added to the homogeneous residual. The replicated class
variances follow the same ordering, but they are not proportional to the
factor, because each class's total residual variance is the sum of the
homogeneous part, which is the same in every class, and the heterogeneous
part. The sign of the pattern depends on the sign of the vQTL effect, which
is drawn with the effect series.

### Fixed and random residual mode

In fixed mode the heterogeneous component is centered and scaled so that its
sample variance equals the requested share. In random mode the per-individual
standard deviations are used directly, with their mean squared value equal to
the share, and nothing is rescaled afterwards.

[]{#complex-random}

```{r}
#| label: complex-random
random_sim <- simulate_phenotype(human, seed = 31, residual_mode = "random") |>
  additive(prop = 0.3, n_qtn = 10) |>
  vqtl(prop = 0.3, qtn = candidate$snp)

fixed_res <- residuals_of(sim)
random_res <- residuals_of(random_sim)

tibble(
  quantity = c("Residual variance", "Phenotypic variance",
               "Genetic / phenotypic variance"),
  parameter = c(0.7, 1, 0.3),
  fixed = c(var(fixed_res$residual), var(fixed_res$value),
            var(fixed_res$genetic_value) / var(fixed_res$value)),
  random = c(var(random_res$residual), var(random_res$value),
             var(random_res$genetic_value) / var(random_res$value))
)
```

In neither mode does the total residual variance equal 0.7 exactly. In fixed
mode the homogeneous and heterogeneous parts each have the requested sample
variance, but their sample covariance is free. In random mode their sample
variances fluctuate as well. The realized genetic ratio is therefore a
diagnostic, as in the [first chapter](first-simulation.qmd).

### A vQTL on the additive loci

By default `vqtl()` reuses the additive loci, so the markers that shift the
mean also change the spread. The number of vQTL loci then follows the additive
layer and `n_qtn` is ignored.

[]{#complex-shared}

```{r}
#| label: complex-shared
shared <- simulate_phenotype(human, seed = 31) |>
  additive(prop = 0.3, n_qtn = 10) |>
  vqtl(prop = 0.3)
print(shared)

qtn_table(shared) |>
  count(layer)
```

Independent variance-only loci are requested with `same_as_add = FALSE` and a
locus count, as in `vqtl(prop = 0.3, same_as_add = FALSE, n_qtn = 2)`.

### Several components in one model

Mean-effect layers and the variance component can be stacked. Here an
additive layer, an epistatic layer, and a vQTL on the additive loci are
combined on the maize panel. The maize lines are inbred, so a dominance layer
is not used here; the [nonadditive chapter](nonadditive.qmd) shows how to
prepare a panel for dominance.

[]{#complex-stack}

```{r}
#| label: complex-stack
stacked <- simulate_phenotype(maize, seed = 5) |>
  additive(prop = 0.3, n_qtn = 10) |>
  epistasis(prop = 0.2, n_pairs = 3) |>
  vqtl(prop = 0.2)
print(stacked)
stacked$var_budget
```

The variance budget lists every component with its share. The genetic share
is the sum of the mean-effect layers, 0.5; the vQTL share belongs to the
residual side of the budget, and the homogeneous residual takes what remains.
The one-call form `simulate_phenotype(maize, h2 = 0.5, model = "AE",
n_qtn = 10)` builds the two mean-effect layers with equal shares; a vQTL has
no one-call shortcut and is always added with `vqtl()`.

### Partial pleiotropy by combining architectures

Suppose two maize traits share some causal loci and also have loci of their
own. No single architecture expresses this, but two models can be combined.
The first uses the pleiotropy architecture with a genetic correlation target
of 0.8; the second uses independent loci for each trait.

[]{#complex-inputs}

```{r}
#| label: complex-inputs
shared_loci <- simulate_phenotype(maize, architecture = "pleiotropy",
                                  n_traits = 2, seed = 10, cor = 0.8) |>
  additive(prop = 0.4, n_qtn = 10)

own_loci <- simulate_phenotype(maize, n_traits = 2, seed = 11) |>
  additive(prop = 0.4, n_qtn = 10)

tibble(
  model = c("shared loci", "own loci"),
  genetic_correlation = c(cor(genetic_values(shared_loci))[1, 2],
                          cor(genetic_values(own_loci))[1, 2])
)
```

The two inputs were given different seeds so that they draw different loci.
With one seed, both models would draw the same markers for the first trait;
the pleiotropy model reuses them for the second trait while the independent
model draws a separate set, so the inputs would no longer be partially
pleiotropic. `complex_phenotypes()` warns about the differing seeds and records
the first.

[]{#complex-combine}

```{r}
#| label: complex-combine
combined <- complex_phenotypes(shared_loci, own_loci, h2 = 0.5)
print(combined)
```

The combined model has architecture `complex`. Its genetic value for each
trait is the sum of the two inputs' genetic values, rescaled to the requested
share, and a new common residual is added; the inputs' own residuals are
discarded. The genetic correlation of the combined traits lies between the two
inputs' correlations.

[]{#complex-correlation}

```{r}
#| label: complex-correlation
combined_genetic <- genetic_values(combined)
summed <- genetic_values(shared_loci) + genetic_values(own_loci)

tibble(
  quantity = c("Genetic correlation", "Genetic variance, Trait_1",
               "Correlation with the summed inputs, Trait_1"),
  parameter = c(NA, 0.5, 1),
  simulated = c(cor(combined_genetic)[1, 2], var(combined_genetic[, 1]),
                cor(combined_genetic[, 1], summed[, 1]))
)

coheritability(combined)
```

The combined genetic value is an exact linear rescaling of the summed inputs,
so the correlation with the sum is one. Each input contributes in proportion
to its genetic variance; with equal shares of 0.4, the two contribute equally
and the correlation is close to the average of 0.8 and the near-zero value of
the independent model. Unequal shares weight the inputs differently.

### Replicated records and correlated residuals in the combined model

The combined model accepts the same residual options as a single model: entry
means over several records, a residual correlation between traits, and the
random residual mode.

[]{#complex-combine-options}

```{r}
#| label: complex-combine-options
combined_means <- complex_phenotypes(shared_loci, own_loci, h2 = 0.5,
                                     reps = 4, resid_cor = 0.3,
                                     residual_mode = "random")
print(combined_means)
```

The printout separates the single-record target share from the entry-mean
realized ratio, which is larger because the residual of a mean of four records
is divided by two. The [trait values chapter](trait.qmd) discusses entry means.

### What a combined model cannot do

The combined object is terminal. Layers cannot be added to it, its
`qtn_table()` is empty, and it cannot be thresholded: `liability_threshold()`
refuses the combined object, and a threshold applied to an input is discarded
with that input's phenotype, because the combined phenotype is rebuilt from the
inputs' genetic values plus a new residual. The inputs must also be built on the same genotypes
with the same individuals and the same number of traits.

[]{#complex-terminal}

```{r}
#| label: complex-terminal
#| error: true
additive(combined, prop = 0.1)
nrow(qtn_table(combined))
```

## Assumptions and theory

The vQTL examples use one additive layer, one vQTL layer, and a normal
residual. The combined-architecture examples use additive layers only. No
selection or crossing is involved.

### The variance-QTL model

Let $s_{iv}$ be the score of individual $i$ at vQTL layer $v$, the sum of its
coded genotypes at the layer's loci multiplied by the layer's effect series,
exactly as for an additive layer. The package standardizes each layer's score
and weights it by the square root of the layer's share $\pi_v$ to form a
loading

$$
L_i = \sum_v \sqrt{\pi_v}\; \frac{s_{iv} - \bar s_v}{s_{s_v}} ,
$$ {#eq-vqtl-loading}

where the bar and $s_{s_v}$ denote the sample mean and sample standard
deviation of the scores. The heterogeneous residual component is then

$$
\log \operatorname{Var}(E_{v,i} \mid \text{genotype}) = \text{constant} + L_i,
\qquad E_{v,i} = z_i \, e^{L_i / 2},
$$ {#eq-vqtl-link}

with $z_i$ a standard normal draw. In fixed mode the vector $E_v$ is centered
and scaled to sample variance $\pi_v$; in random mode the factors
$e^{L_i/2}$ are divided by the root mean square of the factors and multiplied
by $\sqrt{\pi_v}$, so that the mean of the per-individual variance parameters
equals $\pi_v$, and nothing is rescaled afterwards. The component is added to
the homogeneous residual. These are package definitions, implemented by
`.apply_vqtl()`; the equation map records them as entries V2-grammar-19 and
V2-grammar-20.

The exponential link guarantees a positive conditional variance for any
loading. For a single vQTL on one marker, the loading takes three values, one
per genotype class, so the conditional variances form a geometric progression
across classes. This is why the replication table showed a monotone pattern.
The link is a package choice; the documentation of `vqtl()` cites Rönnegård
and Valdar (2011) and Murphy et al. (2022) for the vQTL concept and its
detection and states explicitly that neither is the source of this formula.

### Why a vQTL is residual variance

The heterogeneous component has mean zero within every genotype class in
expectation, so it does not shift the trait mean and contributes nothing to
the covariance between relatives through the mean. It changes the residual
distribution, not the genetic value. The package therefore keeps it in the
residual share of the budget: the residual share is
$1 - \sum_\ell \pi_\ell - \sum_v \pi_v$, with the sum over mean-effect layers
$\ell$ and vQTL layers $v$ (equation map entry V2-grammar-9, implemented by
`.realize_phenotype()`), and the realized genetic ratio uses the vQTL share
only in the denominator (entry V2-grammar-14, `.realized_h2()`). A detection
method that models the residual variance, such as a double generalized linear
model, is the appropriate way to recover such a locus; a model for the mean
will not find it unless the locus also carries a mean effect.

### How the combined model is built

For trait $t$, let $G^{(m)}_{it}$ be the genetic value of individual $i$ in
input model $m$. `complex_phenotypes()` forms the sum $\sum_m G^{(m)}_{it}$,
centers it, divides by its sample standard deviation, and multiplies by
$\sqrt{h^2_t}$, so the combined genetic value has sample variance $h^2_t$. A
fresh residual with share $1 - h^2_t$ is drawn under the chosen residual mode
and divided by $\sqrt{\text{reps}}$ when entry means are requested. This is a
package definition, implemented in `complex_phenotypes()` itself.

Two algebraic consequences follow. First, the combined genetic value is a
linear function of the summed inputs, which the correlation of one above
confirmed. Second, if the inputs have genetic variances $v_1$ and $v_2$ and
genetic correlations $r_1$ and $r_2$ between the two traits, each input gives
both traits the same genetic variance, and the inputs are uncorrelated with
each other, the genetic correlation of the sum is the variance-weighted average

$$
r = \frac{v_1 r_1 + v_2 r_2}{v_1 + v_2} .
$$ {#eq-complex-correlation}

The inputs are not exactly uncorrelated in a finite sample, and the
pleiotropy input's realized correlation is itself a sample quantity, so the
combined correlation is near this value rather than equal to it. With
unequal per-trait shares the general form is
$r = \sum_m r_m \sqrt{v_{m1} v_{m2}} \big/ \sqrt{\sum_m v_{m1} \sum_m v_{m2}}$,
where $v_{mt}$ is input $m$'s genetic variance for trait $t$. Both forms
assume the inputs' genetic values are uncorrelated, which linkage
disequilibrium between their causal loci would violate.

## Implementation and practical limits

All of the computations in this chapter are R functions: the vQTL loading and
heterogeneous residual are built in `.apply_vqtl()`, the residual budget in
`.realize_phenotype()`, and the combination in `complex_phenotypes()`.

- A vQTL on a marker without variation, or with a constant combined loading,
  is an error; on a nearly fixed marker it is legal but uninformative, which is
  why this chapter chose a balanced locus.
- Under the pleiotropy and linkage architectures, `vqtl(qtn = ...)` must
  respect the architecture's rules for shared or paired loci.
- The vQTL share is a marginal phenotypic-variance share. It is not a
  heritability and does not add to one; requesting `prop` for it is mandatory
  even when the simulation has an `h2` budget.
- A combined model keeps no per-locus state: breeding-value selection and
  marker-based methods that need effects refuse it, and `mediation_split()`
  returns nothing for it. Apply such methods to the inputs, or build a single
  model instead.
- Inputs built with different seeds are combined under the first seed with a
  warning; the warning is a reminder that the combined object reproduces only
  if all inputs are rebuilt identically.
