# Meiosis and crossing {#sec-cross}

Suppose you have two inbred maize lines and want to build a mapping or breeding
population from them: an F1, its F2 or doubled-haploid progeny, several
generations of selfing, and a record of who descends from whom. This chapter
simulates those steps on the bundled maize genotypes and checks what the
simulation produces against three expectations that any meiosis simulator must
meet: Mendelian segregation ratios, the halving of heterozygosity under selfing,
and recombination fractions that follow the genetic map.

[]{#cross-setup}

```{r}
#| label: cross-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-cross-example}

### Load the package and the founders

[]{#cross-library}

```{r}
#| label: cross-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

Meiosis needs two things a dosage table does not carry: the phase of each
individual's two homologous chromosomes, and the genetic distances between
markers. `as_population()` supplies the first by building a phased
`Population`; the `cm` column of the genotype table supplies the second.

[]{#cross-founders}

```{r}
#| label: cross-founders
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)

founders <- as_population(geno, individuals = c("33-16", "38-11"))
founders
n_individuals(founders)
```

The bundled map is synthetic, built from physical positions with
`synthetic_map()` ([maps and founders](founders.qmd)). The genetic lengths it
reports are therefore a model of maize recombination, not measured values.
Both founders are inbred lines, so phasing their rare heterozygotes
arbitrarily loses almost nothing; for outbred material that phasing is a real
assumption, discussed below.

### An F1 is heterozygous where the parents differ

`cross()` takes one mother and one father, each a single-individual
`Population`, and returns `n` progeny. Every progeny receives one recombinant
gamete from each parent.

[]{#cross-f1}

```{r}
#| label: cross-f1
f1 <- cross(founders[1], founders[2], n = 1, seed = 11)
f1
```

The markers that can tell the two founders apart are those where both are
homozygous, for different alleles. At those **informative markers** the F1
must be heterozygous, because it received one homologue from each parent.
`dosages()` returns the familiar `-1`, `0`, `1` coding with markers in rows.

[]{#cross-f1-het}

```{r}
#| label: cross-f1-het
d_founders <- dosages(founders)
informative <- which(
  d_founders[, 1] != d_founders[, 2] &
    d_founders[, 1] != 0 & d_founders[, 2] != 0
)

tibble(
  markers = nrow(d_founders),
  informative = length(informative),
  f1_heterozygous_at_informative = mean(dosages(f1)[informative, 1] == 0)
)
```

Requiring both founders to be homozygous matters. A locus where one founder is
heterozygous also differs between them, but the F1 is heterozygous there only
half the time, and including such loci would blur every ratio that follows.

### An F2 segregates 1:2:1

`selfcross()` pollinates an individual with itself. The two gametes come from
independent meioses of the same plant, so at an informative locus the F2
carries each parental homozygote with probability one quarter and the
heterozygote with probability one half.

[]{#cross-f2}

```{r}
#| label: cross-f2
f2 <- selfcross(f1, n = 300, seed = 13)
f2

f2_counts <- table(dosages(f2)[informative, ])
tibble(
  genotype = c("homozygote 33-16 or 38-11", "heterozygote", "other homozygote"),
  parameter = c(0.25, 0.5, 0.25),
  simulated = as.numeric(prop.table(f2_counts))
)
```

The `parameter` column gives the Mendelian expectation; `simulated` is the
proportion over all informative markers and all F2 plants. Which homozygote
appears first in the table depends on which allele is coded `1` at each marker,
so the two homozygote rows are interchangeable.

### Doubled haploids are homozygous by construction

`double_haploid()` draws a single gamete and doubles it, so every doubled
haploid is homozygous at every marker and, at an informative locus, carries
either parent's allele with probability one half.

[]{#cross-dh}

```{r}
#| label: cross-dh
dh <- double_haploid(f1, n = 1000, seed = 14)

tibble(
  any_heterozygote = any(dosages(dh) == 0),
  share_from_first_parent =
    mean(dosages(dh)[informative, ] == d_founders[informative, 1])
)
```

### Selfing halves heterozygosity each generation

Repeated selfing is the classical route from an F2 to inbred lines. Here we
advance 40 lines by single-seed descent for four generations and record the
mean heterozygosity at the informative markers. `mate()` executes one selfing
per line in a single call; its plan names the same individual as mother and
father, which the package records as a self.

[]{#cross-selfing}

```{r}
#| label: cross-selfing
heterozygosity <- function(pop) {
  mean(dosages(pop)[informative, ] == 0)
}

lines <- selfcross(f1, n = 40, seed = 20)
het <- heterozygosity(lines)
for (g in 2:4) {
  ids <- parentage(lines)$id
  plan <- data.frame(mother = ids, father = ids, n = 1)
  lines <- mate(plan, lines, seed = 20 + g)
  het <- c(het, heterozygosity(lines))
}

tibble(
  generation = paste0("S", 1:4),
  parameter = 0.5^(1:4),
  simulated = het
)
```

The `parameter` column is $1/2^t$ for generation $S_t$, the expectation
derived below. Averaging over 40 lines is what makes the agreement visible.
Any single line is a poor estimate, because heterozygosity survives in long
chromosomal blocks rather than locus by locus, so a genome carries far fewer
independent pieces of information than it has markers.

### Pedigree and families

Every `Population` records its pedigree. `parentage()` returns one row per
individual with its parents, generation, founder pool, and the mating design
that produced it. With `ancestors = TRUE` it returns every recorded ancestor as
well.

[]{#cross-parentage}

```{r}
#| label: cross-parentage
parentage(f2) |>
  select(id, mother, father, generation, design) |>
  head(3)

parentage(f2, ancestors = TRUE) |>
  select(id, mother, father, generation, design) |>
  head(4)
```

The hidden `key` columns link individuals by a stable identifier rather than
by display name, so pooled populations with clashing names keep their pedigree
links. `families()` turns the pedigree into a grouping factor, which the
[selection chapter](selection.qmd) uses for within- and among-family selection.

[]{#cross-families}

```{r}
#| label: cross-families
third <- as_population(geno, individuals = "B73")
mixed <- c(
  cross(founders[1], founders[2], n = 3, seed = 31),
  cross(founders[1], third, n = 2, seed = 32),
  selfcross(f1, n = 2, seed = 33)
)

tibble(
  id = parentage(mixed)$id,
  full_sib = as.character(families(mixed, "full_sib")),
  maternal_half_sib = as.character(families(mixed, "maternal_half_sib")),
  selfed = as.character(families(mixed, "selfed"))
)
```

The three crosses sharing mother `33-16` form one maternal half-sib group, the
selfed progeny of `prog_1` form a `selfed` family, and the crosses are `NA`
under `"selfed"` because they are not products of selfing.

### Recombination follows the genetic map

Segregation ratios confirm that alleles are transmitted, but say nothing about
linkage. To check recombination we ask, for pairs of markers on one
chromosome, how often a doubled haploid carries the first parent's allele at
one marker and the second parent's at the other. Doubled haploids make this
easy because their single allele names the founder it came from.

[]{#cross-linkage}

```{r}
#| label: cross-linkage
on_chr1 <- geno$chr == "1"
p1 <- d_founders[on_chr1, 1]
p2 <- d_founders[on_chr1, 2]
inf1 <- which(p1 != p2 & p1 != 0 & p2 != 0)
cm1 <- geno$cm[on_chr1][inf1]

from_p1 <- dosages(dh)[on_chr1, ][inf1, ] == p1[inf1]
pick <- c(5, 20, 60, 150, 400)
distance <- (cm1[pick] - cm1[1]) / 100

recombination <- tibble(
  distance_cM = distance * 100,
  simulated = rowMeans(from_p1[pick, ] != from_p1[rep(1, length(pick)), ]),
  haldane = (1 - exp(-2 * distance)) / 2
)
recombination
```

The simulated recombination fractions rise with distance and never exceed one
half, but at intermediate distances they sit **above** Haldane's map function.
That is expected: the default crossover model includes interference, which
suppresses nearby double crossovers and therefore raises the recombination
fraction between loci tens of centiMorgans apart. The theory section gives the
package's documented formula for the default model. Rerunning the doubled
haploids with `interference = "poisson"` removes interference and recovers
Haldane's function.

[]{#cross-interference}

```{r}
#| label: cross-interference
gamma_r <- function(d, nu = 2.6, p = 0) {
  y <- 2 * nu * (1 - p) * d
  p0 <- exp(-2 * p * d) *
    (1 - pgamma(y, nu + 1) - (y / nu) * (1 - pgamma(y, nu)))
  (1 - p0) / 2
}

dh_poisson <- double_haploid(f1, n = 1000, seed = 14, interference = "poisson")
from_p1_poisson <- dosages(dh_poisson)[on_chr1, ][inf1, ] == p1[inf1]

recombination |>
  mutate(
    gamma_model = gamma_r(distance),
    simulated_poisson = rowMeans(
      from_p1_poisson[pick, ] != from_p1_poisson[rep(1, length(pick)), ]
    )
  ) |>
  select(distance_cM, simulated, gamma_model, simulated_poisson, haldane)
```

Each simulated fraction is an average over 1000 lines, so its standard error is
roughly $\sqrt{r(1-r)/1000}$, below 0.02. The default draws track the gamma
model column and the Poisson draws track Haldane's column within that
precision.

### Reproducibility

All random quantities are drawn by R, so a pedigree is reproducible from its
seed. A seeded call also restores the caller's random-number state on exit, so
it does not disturb other draws in a script.

[]{#cross-reproduce}

```{r}
#| label: cross-reproduce
again <- selfcross(f1, n = 300, seed = 13)
identical(dosages(again), dosages(f2))
```

## Assumptions and theory

This chapter simulates diploid meiosis without sex-specific recombination,
without mutation, and without selection. Parents are phased founders or their
simulated descendants. Chromosomes assort independently, and the genetic map
is taken as given.

### How a gamete is made

For each chromosome the package draws the number of crossovers and their
positions, then draws which homologue the gamete starts on. The deterministic
part, turning those positions into a recombinant chromosome, is the ancestry
mask: the gamete copies the starting homologue up to the first crossover,
switches at every crossover, and the result is the gamete's strand (package
definition; `.draw_meiosis()` draws in R, `chromosome_mask()` and
`recombine()` in the Rust module `meiosis.rs` apply the draws).

Two crossover models are available.

- **Poisson, `interference = "poisson"`.** The number of crossovers on a
  chromosome is Poisson with mean equal to the chromosome's length in Morgans,
  and their positions are uniform along it. This is the count-location model
  of the isqg package, and the package's draws reproduce isqg's random stream
  exactly (package definition, `.draw_meiosis()`; reference Toledo et al. 2019
  as cited in `?cross`). The chromosome length is its last map position, not
  its span, so crossovers upstream of the first marker only swap the whole
  chromosome and do not change the recombination fraction between markers.
- **Gamma interference, the default.** Chiasmata on the four-strand bivalent
  follow a stationary renewal process whose gaps are gamma distributed with
  shape $\nu = 2.6$ and, optionally, a share $p$ of non-interfering chiasmata
  (here $p = 0$). A gamete takes part in each chiasma independently with
  probability one half. These defaults are AlphaSimR's and approximate
  Kosambi's map function; the package documents them as following the gamma
  model of McPeek and Speed (1995) and the two-pathway extension of Housworth
  and Stahl (2003), while stating that its formulas were derived and checked
  numerically in its own tests rather than copied from those papers. Both
  models give one crossover per Morgan on average, so the genetic map keeps
  its meaning under either.

Under the Poisson model the recombination fraction between two loci $d$
Morgans apart is Haldane's

$$
r(d) = \tfrac{1}{2}\left(1 - e^{-2d}\right),
$$ {#eq-haldane}

because the loci recombine when an odd number of crossovers falls between
them, and for a Poisson count the probability of an odd number is
$(1 - e^{-2d})/2$. The equation map records this as Haldane (1919), page
unverified, and as a property of the Poisson process rather than a formula the
code evaluates. Under the gamma model the package documents

$$
r(d) = \frac{1 - P_0(d)}{2},\qquad
P_0(d) = e^{-2pd}\left\{1 - F_{\nu+1}(y) - \frac{y}{\nu}\left[1 - F_\nu(y)\right]\right\},
\quad y = 2\nu(1-p)\,d,
$$ {#eq-gamma-r}

where $P_0(d)$ is the probability of no chiasma in the interval and $F_a$ is
the distribution function of a Gamma($a$, 1) variable. The `gamma_r()`
function above evaluates this expression as written in `?cross`; it is the
package's documented formula, not an independent derivation in this book.

### Why selfing halves heterozygosity

Consider one locus at which a plant is heterozygous. Each of its gametes
carries either allele with probability one half, independently, so the selfed
offspring is heterozygous with probability one half and homozygous otherwise.
A homozygous locus stays homozygous. Averaging over loci and lines, the
expected heterozygosity after $t$ generations of selfing is

$$
H_t = H_0 / 2^t,
$$ {#eq-selfing}

with $H_0$ the heterozygosity of the starting plant, here the F1, which is one
at every informative marker. This is an algebraic consequence of Mendelian
transmission at a single locus; linkage changes the variance among lines but
not this expectation. The equation map attributes the result to Falconer and
Mackay (1996), chapter 5, page unverified.

### Doubled haploids and the F2

A doubled haploid copies one gamete onto both homologues, so it is homozygous
everywhere and shows the gamete's allele directly. At an informative locus the
gamete carries each parent's allele with probability one half. An F2 combines
two independent gametes of the F1, giving the 1:2:1 ratio. Both facts are
package definitions of `double_haploid()` and `selfcross()` that follow from
the gamete construction above.

### Phasing of founders

A dosage matrix says that an individual is heterozygous but not which
homologue carries which allele. `as_population()` puts allele 1 on the first
strand and allele 2 on the second at every heterozygote (package convention;
equation map entry V2-crossing-1). For an inbred panel this is nearly
lossless. For an outbred sample it imposes a coupling phase on every
heterozygote, which distorts the haplotypes that recombination then shuffles.
Use `population_from_haplotypes()` when phased haplotypes are available
([maps and founders](founders.qmd)).

## Implementation and practical limits

R draws every random quantity, in isqg's order, and passes the results to
Rust; the Rust core never calls a random-number generator. This is the
boundary described in the [implementation overview](architecture.qmd) and is
what makes the Poisson draws bit-identical to isqg and every pedigree
reproducible from its seed. `mate()` batches many matings into one pass
through the Rust core while keeping the same draws as the equivalent
sequence of `cross()`, `selfcross()`, and `double_haploid()` calls.

- Both parents must share one marker map and the same allele coded `1`. Convert
  panels together, or with a fixed `ref_allele`, before crossing populations
  built from separately converted files.
- The map must have no missing `cm` values and must be non-decreasing within
  each chromosome. A physical-position-only table needs `synthetic_map()`.
- A chromosome's crossover count depends on its last map position. A map
  whose first marker sits far from zero adds crossovers that swap whole
  chromosomes and cost computing time without changing recombination between
  markers.
- Seeded results of every meiosis function changed when the default moved
  from Poisson to gamma interference in version 2.0.0.9003; pass
  `interference = "poisson"` to reproduce older results, or set the option
  `simplePHENOTYPES.interference` once for a session.
- The pedigree records parents, not per-locus founder origin, so realized
  ancestry fractions along the genome are not tracked. The
  [mating designs chapter](mating.qmd) returns to this point.

::: {.callout-tip title="A check worth keeping"}
Segregation ratios cannot see phasing or linkage bugs. The recombination table
can: an earlier development version produced haplotypes the founders never
had, and its F2 and doubled-haploid ratios still looked perfect while the
recombination fractions came out near one. Keep the linkage check in any
pipeline that modifies meiosis.
:::
