# Transcriptome simulation and counts {#sec-expression}

Suppose you want gene expression for your genotyped individuals: a matrix of
genes by individuals in which some of the variation traces to the genome through
cis and trans regulatory effects, some is shared co-expression, and the rest is
gene-specific noise. You also want to know the truth behind it, so that an eQTL
or co-expression analysis can be scored, and you may need the same expression
as integer sequencing counts. This chapter builds such a transcriptome on the
maize genotypes, checks what the requested parameters mean once realized, applies
the same architecture to new individuals, and converts the expression to counts.

[]{#expression-setup}

```{r}
#| label: expression-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-expression-example}

### Load the package and the genotypes

[]{#expression-library}

```{r}
#| label: expression-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)

data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
```

The maize panel and its coding are described in
[A first simulation](first-simulation.qmd). Expression simulation needs marker
chromosomes and physical positions, because cis regulation is defined by
distance from a gene's transcription start site. The panel provides both.

### Simulate a transcriptome

`simulate_transcriptome()` draws a gene set with synthetic coordinates, assigns
each gene a target expression heritability and a cis fraction, samples cis
markers near each gene and trans hub markers for latent regulatory factors,
and returns the expression matrix together with the truth tables.

[]{#expression-simulate}

```{r}
#| label: expression-simulate
tx <- simulate_transcriptome(geno, n_genes = 400, seed = 11)
tx
dim(tx$expression)
```

The default `h2 = "beta"` draws each gene's target heritability from a
Beta(1.5, 6) distribution with mean 0.20, and the default `cis_fraction = 0.25`
asks that one quarter of each gene's genetic variance come from cis markers.
The print method summarizes the realized heritabilities, the number of
cis-eQTL, and the number of trans hub markers. The object holds the expression
matrix, the genetic part of expression, a per-gene table, the eQTL truth
tables, and a per-gene variance budget.

[]{#expression-genes}

```{r}
#| label: expression-genes
tx$genes |>
  select(gene_id, chr, tss, module, h2_target, h2_realized, h2_allocated,
         cis_fraction_target, cis_fraction_realized, n_cis) |>
  head()
```

Each gene belongs to a co-expression module, which is also the latent factor
carrying its trans regulation. The table reports three heritability columns,
explained below, together with the realized cis fraction and the number of
cis-eQTL drawn for the gene.

### Compare requested and realized heritability

A per-gene target is a parameter; the realized value is a sample quantity. The
package reports both and does not force them to agree. The table compares the
distribution of targets with the distribution of realized values across genes.

[]{#expression-h2}

```{r}
#| label: expression-h2
h2_summary <- function(x) c(mean = mean(x), median = median(x),
                            min = min(x), max = max(x))

tibble(
  quantity = names(h2_summary(tx$genes$h2_target)),
  target = h2_summary(tx$genes$h2_target),
  allocated = h2_summary(tx$genes$h2_allocated),
  realized = h2_summary(tx$genes$h2_realized)
)
```

`h2_allocated` equals the target on the reference panel by construction: the
genetic and residual components of each gene are scaled separately to sample
variances `h2` and `1 - h2`. `h2_realized` divides the genetic variance by the
variance of the expression actually produced, which includes the chance sample
covariance between the two components. It therefore scatters around the target,
exactly as the realized variance ratio did in the first chapter. The following
chunk shows the scatter for a constant target, where it is easiest to see.

[]{#expression-fixed-h2}

```{r}
#| label: expression-fixed-h2
tx_fixed <- simulate_transcriptome(
  geno, n_genes = 400, h2 = 0.3, cis_fraction = 0.5, seed = 12
)

tibble(
  quantity = c("Expression heritability", "Cis fraction"),
  parameter = c(0.3, 0.5),
  realized_mean = c(mean(tx_fixed$genes$h2_realized),
                    mean(tx_fixed$genes$cis_fraction_realized)),
  realized_min = c(min(tx_fixed$genes$h2_realized),
                   min(tx_fixed$genes$cis_fraction_realized)),
  realized_max = c(max(tx_fixed$genes$h2_realized),
                   max(tx_fixed$genes$cis_fraction_realized))
)
```

The realized cis fraction varies for the same reason. The cis score and the
trans score of a gene are standardized separately and blended with weights
$\sqrt{\omega}$ and $\sqrt{1-\omega}$, so their marginal shares equal the
requested fraction. But the two scores covary through linkage disequilibrium
and population structure, and the realized fraction is defined relative to the
variance of the combined genetic component, which contains that covariance.

### Inspect the eQTL truth tables

The cis table lists, for each gene, the markers inside its cis window and their
effects on centered dosage. The factor table lists the hub markers driving each
latent factor. Joining the cis table with the gene table confirms the window
rule: every cis marker is on the gene's chromosome and within `cis_window`
(one million positions by default) of its transcription start site.

[]{#expression-cis}

```{r}
#| label: expression-cis
head(tx$cis_eqtl)
head(tx$factor_eqtl)

cis_check <- tx$cis_eqtl |>
  left_join(
    tx$genes |> select(gene_id, gene_chr = chr, tss),
    by = "gene_id"
  ) |>
  mutate(same_chromosome = chr == gene_chr,
         distance = abs(pos - tss))

tibble(
  cis_markers = nrow(cis_check),
  genes_with_cis = n_distinct(cis_check$gene_id),
  all_on_gene_chromosome = all(cis_check$same_chromosome),
  max_distance_from_tss = max(cis_check$distance),
  cis_window = tx$reference$cis_window
)
```

Only markers with minor allele frequency of at least 0.05 and a known
chromosome and position are eligible as eQTL. This threshold is fixed in the
generator. The bundled maize panel was already filtered to a high minor allele
frequency, so every marker is eligible here. A sparser panel can leave a gene
without any cis marker; that gene then realizes its heritability through its
trans factor alone, with a realized cis fraction of zero. Only a gene with
neither a cis marker nor an active module factor realizes zero heritability.

### Reconstruct the genetic expression from the truth

The truth tables are complete. For a gene without epistasis, the genetic part of
its expression is the cis effects applied to dosages centered on the reference
marker means, plus the gene's `trans_scale` times the hub effects of its module.
The chunk reconstructs it for every gene and compares with the stored matrix.

[]{#expression-reconstruct}

```{r}
#| label: expression-reconstruct
dosage <- t(as.matrix(geno[, -(1:5)]))
colnames(dosage) <- geno$snp
centered <- sweep(dosage, 2, tx$reference$marker_mean[colnames(dosage)])

reconstruct <- function(g) {
  gene <- tx$genes[g, ]
  cis <- tx$cis_eqtl |> filter(gene_id == gene$gene_id)
  hubs <- tx$factor_eqtl |> filter(factor == gene$module)
  as.numeric(centered[, cis$snp, drop = FALSE] %*% cis$effect) +
    gene$trans_scale * as.numeric(centered[, hubs$snp, drop = FALSE] %*% hubs$hub_effect)
}
genetic_hat <- sapply(seq_len(tx$n_genes), reconstruct)

tibble(
  quantity = "Largest absolute difference, reconstructed vs stored genetic expression",
  value = max(abs(genetic_hat - t(tx$genetic_expression)))
)
```

The agreement is at floating-point precision. The same reconstruction is what
`predict()` applies to new individuals below. The variance budget uses these
parts: for each gene, the cis variance, the trans variance, and twice their
covariance add up to the variance of the genetic expression.

[]{#expression-budget}

```{r}
#| label: expression-budget
budget <- tx$var_budget |>
  mutate(genetic_variance = apply(tx$genetic_expression, 1, var),
         closure = v_cis + v_trans + cis_trans_cov - genetic_variance)

budget |>
  select(gene_id, v_cis, v_trans, cis_trans_cov, genetic_variance, closure) |>
  head()
```

The `gr_cov` column of the budget, not shown, records twice the sample
covariance between the genetic and residual parts; it is the term that moves
`h2_realized` away from `h2_allocated`.

### Co-expression without genetics

Genes co-express through their module's non-genetic factor, controlled by
`residual_module_fraction`, and through shared trans regulation. To see the
non-genetic part alone, simulate a transcriptome with no genotypes. In this
mode every gene has zero heritability and `n_ind` sets the sample size. The
table compares mean absolute correlations within and between modules for two
settings of the module fraction.

[]{#expression-coexpression}

```{r}
#| label: expression-coexpression
module_correlation <- function(kappa) {
  tx0 <- simulate_transcriptome(
    n_genes = 200, n_ind = 200, n_factors = 4,
    residual_module_fraction = kappa, seed = 7
  )
  r <- cor(t(tx0$expression))
  same <- outer(tx0$genes$module, tx0$genes$module, "==")
  diag(same) <- NA
  c(within = mean(abs(r[which(same)])), between = mean(abs(r[which(!same)])))
}

tibble(
  residual_module_fraction = c(0, 0.5),
  within_module = c(module_correlation(0)["within"],
                    module_correlation(0.5)["within"]),
  between_module = c(module_correlation(0)["between"],
                     module_correlation(0.5)["between"])
)
```

With a zero module fraction, the within-module and between-module correlations
are both at the level expected from noise alone. With half of the residual
variance carried by the module factor, genes in the same module are clearly
correlated while genes in different modules are not. A naive co-expression
network analysis of this genotype-free transcriptome would recover the modules
even though no gene is heritable: co-expression is not evidence of shared
genetic control.

### Apply the architecture to new individuals

A simulated transcriptome is tied to its reference individuals through the
marker means used for centering and the scaling of each gene. `predict()`
applies the stored architecture to different genotypes without recomputing any
constant, so the genetic expression of a given genotype is the same whichever
population it sits in. The example fits the architecture on 200 maize lines and
predicts the remaining 80.

[]{#expression-predict}

```{r}
#| label: expression-predict
reference <- geno[, c(1:5, 5 + 1:200)]
new_lines <- geno[, c(1:5, 5 + 201:280)]

tx_ref <- simulate_transcriptome(reference, n_genes = 400, seed = 11)
tx_new <- predict(tx_ref, new_lines, seed = 5)
tx_new

tibble(
  panel = c("Reference (200 lines)", "New (80 lines)"),
  mean_target = c(mean(tx_ref$genes$h2_target), mean(tx_new$genes$h2_target)),
  mean_realized = c(mean(tx_ref$genes$h2_realized), mean(tx_new$genes$h2_realized)),
  sd_realized_minus_target = c(sd(tx_ref$genes$h2_realized - tx_ref$genes$h2_target),
                               sd(tx_new$genes$h2_realized - tx_new$genes$h2_target))
)
```

On the new lines the residual is drawn afresh, and nothing is rescaled. The
realized heritability is therefore emergent: it depends on the allele
frequencies and linkage of the new individuals, and it scatters more widely in
a smaller panel. `predict(..., residual = FALSE)` returns the noiseless genetic
expression, which on the reference panel itself reproduces `genetic_expression`.

[]{#expression-predict-reference}

```{r}
#| label: expression-predict-reference
noiseless <- predict(tx_ref, reference, residual = FALSE)
tibble(
  quantity = "Largest absolute difference, noiseless prediction vs stored genetic expression",
  value = max(abs(noiseless$expression - tx_ref$genetic_expression))
)
```

The new genotypes must carry every marker named in the eQTL tables with the
same allele orientation; only names and the dosage coding are checked, as the
[genotype-input chapter](input.qmd) discusses for merged panels.

### Add epistatic expression

`epistasis` assigns each gene one or two marker pairs whose centered dosage
product contributes to its expression, with a requested share of the genetic
variance. The pairs are reported in an additional truth table.

[]{#expression-epistasis}

```{r}
#| label: expression-epistasis
tx_epi <- simulate_transcriptome(geno, n_genes = 100, epistasis = 0.2, seed = 13)
head(tx_epi$epi_eqtl)

tibble(
  quantity = "Epistatic share of realized genetic variance",
  parameter = 0.2,
  realized_mean = mean(tx_epi$genes$epistasis_realized),
  realized_min = min(tx_epi$genes$epistasis_realized),
  realized_max = max(tx_epi$genes$epistasis_realized)
)
```

As with the cis fraction, `epistasis` is a marginal blend weight, and the
realized share differs from it through the covariance between the epistatic
score and the additive scores. The function documentation gives the exact
relation.

### Calibrate to an existing expression matrix

`mimic` takes a genes-by-individuals expression matrix for the same genotyped
individuals and sets the generator's targets from it: the per-gene mean and
variance, a per-gene heritability estimated by a single-component REML fit on
the genomic relationship matrix, the number of co-expression factors, and the
residual module fraction. Here the matrix simulated above stands in for a
user's data, which lets us compare the estimated targets with the known truth.

[]{#expression-mimic}

```{r}
#| label: expression-mimic
tx_mimic <- simulate_transcriptome(geno, mimic = tx$expression, seed = 4)
tx_mimic

calibration <- tx_mimic$calibration$h2 |>
  left_join(tx$genes |> select(gene_id, h2_target, h2_realized), by = "gene_id")

tibble(
  quantity = c("Largest absolute difference in gene means",
               "Largest absolute difference in gene variances"),
  value = c(
    max(abs(rowMeans(tx_mimic$expression) - rowMeans(tx$expression))),
    max(abs(apply(tx_mimic$expression, 1, var) - apply(tx$expression, 1, var)))
  )
)

tibble(
  quantity = c(
    "Correlation of REML heritability with the realized truth",
    "Mean REML heritability minus the realized truth",
    "Estimated factors (5 were simulated)",
    "Estimated residual module fraction (0.15 was simulated)"
  ),
  value = c(
    cor(calibration$h2_greml, calibration$h2_realized),
    mean(calibration$h2_greml - calibration$h2_realized),
    tx_mimic$calibration$n_factors,
    tx_mimic$calibration$kappa
  )
)
```

The means and variances are met to rounding. The heritability calibration is a
distribution-level match: the per-gene estimates correlate with the truth but
are noisy at 280 individuals, and the documentation warns that they are less
reliable below 100. The eQTL effects, loadings, and module memberships are
drawn afresh, so a mimicked transcriptome reproduces the moments and the
heritability distribution of the input, not its individual genes.

### Observe sequencing counts

`observe_counts()` maps the latent expression to integer counts through a
negative-binomial model with a log link. Per gene, the latent expression is
standardized, multiplied by a coupling slope, and added to a log baseline; per
individual, a library-size factor scales the expected count. With a zero
coupling the expected count is constant across individuals, which isolates the
negative-binomial mean-variance relation, $\operatorname{Var} = \mu + \phi\mu^2$.

[]{#expression-counts}

```{r}
#| label: expression-counts
count_moments <- function(phi) {
  counts <- observe_counts(
    tx, coupling = 0, baseline = log(50), dispersion = phi, seed = 3
  )$counts
  c(mean = mean(rowMeans(counts)), variance = mean(apply(counts, 1, var)))
}
moments <- sapply(c(0, 0.1, 0.5), count_moments)

tibble(
  dispersion = c(0, 0.1, 0.5),
  parameter_mean = 50,
  simulated_mean = moments["mean", ],
  parameter_variance = 50 + c(0, 0.1, 0.5) * 50^2,
  simulated_variance = moments["variance", ]
)
```

A dispersion of zero is the Poisson limit, where the variance equals the mean.
Larger dispersions add the quadratic term. The default coupling of 0.5 means a
one-standard-deviation change in latent expression multiplies the expected
count by $e^{0.5}$. The library-size factor enters multiplicatively, which the
column totals show.

[]{#expression-library-size}

```{r}
#| label: expression-library-size
depth <- c(rep(0.5, 140), rep(2, 140))
counts <- observe_counts(tx, library_size = depth, seed = 4)
totals <- colSums(counts$counts)

tibble(
  quantity = "Ratio of mean column totals, deep to shallow libraries",
  parameter = 2 / 0.5,
  simulated = mean(totals[depth == 2]) / mean(totals[depth == 0.5])
)
counts$counts[1:4, 1:4]
```

The count matrix and the realized expected-count matrix are stored in the
returned object. Counts are an observation layer for count-based methods. The
share of count variance explained by the genome is not the latent heritability,
because the link is nonlinear and mean-dependent.

## Assumptions and theory

The expression model, the eQTL architecture, and the count model are package
definitions. The literature references below are those recorded in the
package's equation map and documentation; where the map marks a page as
unverified, that status is kept here.

### The expression model

For gene $g$ and individual $i$, the normalized expression is

$$
E_{gi} = \mu_g + G_{gi} + R_{gi},
$$ {#eq-expression-model}

a gene location $\mu_g$ (zero unless `mimic` is used), a genetic component
$G_{gi}$, and a residual component $R_{gi}$. The additive dosage model and the
definition of heritability follow Falconer and Mackay (1996; page unverified).
The genetic and residual components are drawn independently, so their
covariance is zero in expectation but not in a finite sample.

Dosages are centered on the reference marker means, $Z_{ij} = x_{ij} - \bar x_j$,
a package convention that fixes the genetic scale on the reference panel. The
cis score of a gene sums normal effects over markers in its window,
$c_g = \sum_{j \in W(g)} \beta_{gj} Z_{j}$, with one to three cis markers per
gene (GTEx Consortium 2020, recorded in the map as verified). The trans score is
the gene's latent factor, $f_q = \sum_{k \in H(q)} \gamma_{qk} Z_k$, driven by
one or two hub markers outside the cis windows of the module's genes (Albert et
al. 2018, recorded as verified). Each component is standardized with the
sample standard deviation, denominator $n-1$.

The additive genetic score blends the standardized cis and trans scores and is
then scaled to the target heritability:

$$
G^0_g = \sqrt{\omega_g}\,\tilde c_g + \sqrt{1-\omega_g}\,\tilde t_g,
\qquad
G_g = \sqrt{h^2_g}\;\frac{G^0_g}{s_{G^0_g}}.
$$ {#eq-expression-genetic}

Here $\omega_g$ is the cis fraction and $h^2_g$ the target heritability. The
residual is built the same way from the module factor $u_{q(g)}$ and
gene-specific noise $\varepsilon_g$ with the residual module fraction $\kappa$:

$$
R^0_g = \sqrt{\kappa}\,\tilde u_{q(g)} + \sqrt{1-\kappa}\,\tilde\varepsilon_g,
\qquad
R_g = \sqrt{1-h^2_g}\;\frac{R^0_g}{s_{R^0_g}}.
$$ {#eq-expression-residual}

These are package definitions implemented in `simulate_transcriptome()`; the
map records them as the package's own composite design. With epistasis, the
standardized sum of centered dosage products for the gene's marker pairs is
blended with the additive score with weights $\sqrt{1-\epsilon}$ and
$\sqrt{\epsilon}$ before the final scaling.

### Three heritability columns

Because $G_g$ and $R_g$ are each scaled on the reference panel, the allocation
$\operatorname{Var}(G_g) / (\operatorname{Var}(G_g) + \operatorname{Var}(R_g))$
equals $h^2_g$ there. The realized heritability is

$$
h^2_{\text{realized},g} = \frac{\operatorname{Var}(G_g)}{\operatorname{Var}(E_g)}
= \frac{h^2_g}{1 + 2\operatorname{Cov}(G_g, R_g)},
$$ {#eq-expression-realized}

where the second form holds only when the two variances sum to one, that is,
without `mimic` rescaling. This is the same sample-variance identity derived in
@sec-variance-derivation; the covariance is reported in the budget as `gr_cov`
rather than projected away. The quantity is not bounded by one, and the
documentation warns that it is unreliable below 30 individuals.

### The count model

With $z_{gi}$ the per-gene standardized latent expression,

$$
\log \mu_{gi} = \log L_i + \alpha_g + \sigma_g z_{gi},
\qquad
Y_{gi} \sim \mathrm{NB}(\mu_{gi}, \phi_g),
\qquad
\operatorname{Var}(Y_{gi}) = \mu_{gi} + \phi_g \mu_{gi}^2.
$$ {#eq-expression-counts}

$L_i$ is the library-size factor, $\alpha_g$ the log baseline, $\sigma_g$ the
coupling, and $\phi_g$ the dispersion, with $\phi_g = 0$ giving the Poisson
limit. The variance parameterization follows the edgeR convention (McCarthy,
Chen and Smyth 2012; Robinson and Smyth 2008; page-level equation numbers not
verified). The log link to the latent expression and the library-size factor
are the package's own design, implemented in `observe_counts()`.

### What mimic estimates

The per-gene heritability under `mimic` is a single-variance-component REML
estimate on a genomic relationship matrix built from reference-centered
dosages (`.tx_grm()`, after VanRaden 2008 method 1 up to a constant; the map
notes the page was not re-verified). The profile likelihood is the EMMA form of
Kang et al. (2008), recorded as verified, implemented in `.greml_h2()`. The
factor count uses the Marchenko-Pastur edge of the sample covariance spectrum
(`.tx_estimate_factors()`; page unverified), and the residual module fraction
comes from a spiked-covariance estimator (`.tx_estimate_kappa()`). That
estimator subtracts the genetic trans share implied by the GREML heritability
and by the `cis_fraction` passed to the mimic call, so it over-estimates the
fraction when the assumed cis fraction is higher than the input's.

## Implementation and practical limits

The generator is R code in `simulate_transcriptome()`, with `predict.transcriptome_sim()`
for re-scoring, `observe_counts()` for counts, and the `mimic` estimators in
the `.tx_*` and `.greml_h2()` helpers. No Rust kernel is involved.

- The default profile is a documented benchmarking compromise, not a set of
  biological constants. Report the profile and parameters with any result.
- Eligibility requires minor allele frequency of at least 0.05 and known
  coordinates. On a small or sparse panel, a gene can lack both a cis marker
  and an active hub, and then realizes zero heritability; the gene table shows it.
- Random draws flow gene by gene from one stream, so changing an option for one
  gene changes the draws of every later gene. Per-gene reproducibility holds only
  for the same full call and seed.
- `mimic` reproduces moments, the heritability distribution, and the factor
  count; it does not retain the input's loadings or their signs, so
  anti-correlated blocks of the input become positively co-expressed.
- Count-scale heritability differs from latent heritability. Use counts as an
  observation layer for methods that model counts.
- Directed regulatory networks and tissue specificity are outside the scope of
  this generator.

The next chapter, [Expression-mediated phenotypes](mediation.qmd), uses a
transcriptome from this generator as a source of phenotypic variation and
separates the part of that variation that traces to the genome.
