# A first simulation {#sec-first-simulation}

Suppose you have genotyped individuals and want to simulate a continuous trait
controlled by 20 additive causal markers on their observed genotypes.
You want to retain the simulated genetic values, inspect the phenotype, and save
the result for a downstream analysis. This chapter carries out that workflow
and checks what a requested genetic variance allocation of 0.5 actually means.

[]{#first-setup}

```{r}
#| label: first-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-first-example}

### Load the package

Use the development version corresponding to this book. The commands below assume
the package is installed; see the [build instructions](development.qmd#sec-build)
if you are working from a source checkout.

[]{#first-library}

```{r}
#| label: first-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
library(readr)
```

### Bring your own genotype data

The starting point is your genotype dataset. `as_numeric()` reads supported
formats, including VCF and HapMap, and prepares the numeric input used by the
simulation. Replace the path below with your own file. This recipe is not executed
when building the book because it refers to a reader-supplied file.

```{r}
#| eval: false
geno <- as_numeric(
  "path/to/your_genotypes.vcf",
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
```

The same workflow accepts a HapMap file by replacing the path with, for example,
`"path/to/your_genotypes.hmp.txt"`. Format-specific dependencies and input checks
are covered in the [genotype input chapter](input.qmd). Before simulating, check
individual identifiers, allele coding, missingness, and marker quality; importing
a file does not establish that it is suitable for your study.

### Follow along with observed maize genotypes

To make this chapter runnable without an external download, we use the bundled
`SNP55K_maize282_maf04` dataset: 10,650 markers and 280 maize individuals from
an inbred association panel. The dataset name refers to the original 282-line
panel; the bundled table contains 280 individual columns. Its documentation
attributes the marker data to Cook et al. (2012); see
`?SNP55K_maize282_maf04` for provenance.

These genotypes are already numeric. Here `as_numeric()` normalizes the input
schema; it is not converting raw nucleotide calls. If you loaded your own file
above, keep that `geno` object and skip this replacement block. All subsequent
steps use the same object name.

[]{#first-genotypes}

```{r}
#| label: first-genotypes
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)

tibble(markers = nrow(geno), individuals = ncol(geno) - 5)

geno |>
  select(1:9) |>
  head()
```

Markers occupy rows; the first five columns contain marker metadata and the
remaining columns identify individuals. Genotypes use `-1`, `0`, `1` coding;
see [coding conventions](conventions.qmd#sec-coding). The bundled data were
filtered to minor allele frequency greater than 0.4 to reduce their size, so
they do not represent the full allele-frequency spectrum of the original panel.

::: {.callout-note title="Observed genotypes, simulated phenotype"}
The simulation uses the observed marker patterns, including their associations
across loci, to generate a new trait. It does not use measured maize phenotypes.
The bundled `cm` column is a **synthetic genetic map**, not measured recombination
distances. This additive phenotype example does not simulate meiosis.
:::

### Specify and inspect the trait

[]{#first-model}

```{r}
#| label: first-model
sim <- simulate_phenotype(geno, seed = 901) |>
  additive(prop = 0.5, n_qtn = 20)
print(sim)
```

`simulate_phenotype()` establishes the simulation and its seed. `additive()`
samples the causal markers and constructs the genetic component. The default
effect series is geometric, so the 20 markers do not contribute equally.
The object already contains realized phenotypes after the layer is added.

[]{#first-tables}

```{r}
#| label: first-tables
long <- phenotypes_long(sim)
wide <- phenotypes_wide(sim)
truth <- genetic_values(sim)
qt <- qtn_table(sim)

head(long)
qt |>
  select(snp, effect, var_explained) |>
  head()
```

The long table has one row per individual, trait, and simulation replication.
The wide table has one trait column per trait. `genetic_values()` returns the
simulated genetic values, and `qtn_table()` identifies the causal markers and
their model effects. These are known simulation quantities, not estimates from
phenotypes. The `effect` column is the layer's effect series; component scaling
also enters the realized genetic values, as explained below.

### Check the realized variance

We compare what we requested with what the simulation produced. First, join
phenotypes and genetic values by individual ID. `as_tibble(..., rownames = "id")`
turns the genetic-value matrix into a table and keeps its individual names as an
`id` column. `left_join()` matches individuals by that column, so row order does
not matter.

In this example the phenotype is a genetic value plus a residual. Subtracting
the genetic value therefore recovers the residual.

[]{#first-variance}

```{r}
#| label: first-variance
genetic <- truth |>
  as_tibble(rownames = "id") |>
  rename(genetic_value = Trait_1)

values <- long |>
  left_join(genetic, by = "id") |>
  mutate(residual = value - genetic_value)

comparison <- tibble(
  quantity = c(
    "Genetic variance", "Residual variance", "Genetic-residual covariance",
    "Phenotypic variance", "Genetic / phenotypic variance"
  ),
  parameter = c(0.5, 0.5, 0, 1, 0.5),
  simulated = c(
    var(values$genetic_value), var(values$residual),
    cov(values$genetic_value, values$residual), var(values$value),
    var(values$genetic_value) / var(values$value)
  )
)

comparison
```

The `parameter` column shows the design values. The last two entries are the
benchmarks obtained when genetic and residual components are uncorrelated;
0.5 is not a claim about the average of the sample variance ratio across runs.
The genetic and residual sample variances equal 0.5 under the default settings.
Their sample covariance can be positive or negative, so the last two rows can
differ from their benchmarks. We explain why below.

[]{#first-plot}

```{r}
#| label: first-plot
#| fig-height: 6
#| fig-cap: "The default diagnostic plots for the maize simulation."
plot(sim)
```

The default plot shows the variance budget, phenotype distribution, marker
effects, and requested versus realized genetic variance ratio. The budget panel
shows the model's allocations; it is not a measurement of every sample variance.
The histogram shows the resulting distribution; additive effects alone do not
guarantee a bell-shaped phenotype distribution.

### Export and reproduce the result

Create a results folder and save the phenotype and causal-marker tables.
`write_phenotypes()` and `write_qtn_table()` use the simulation object directly.

[]{#first-export}

```{r}
#| label: first-export
dir.create("results", showWarnings = FALSE)
write_phenotypes(sim, file = "results/phenotypes.tsv", sep = "\t")
write_qtn_table(sim, file = "results/causal-markers.tsv", sep = "\t")

saved <- read_tsv("results/phenotypes.tsv", show_col_types = FALSE)
saved |> head()
```

Use the same input and seed to repeat the simulation. The table below places the
two sets of phenotypes next to each other.

[]{#first-reproduce}

```{r}
#| label: first-reproduce
again <- simulate_phenotype(geno, seed = 901) |>
  additive(prop = 0.5, n_qtn = 20)

first_run <- phenotypes_long(sim) |>
  rename(first_value = value)

second_run <- phenotypes_long(again) |>
  rename(second_value = value)

first_run |>
  left_join(second_run, by = c("id", "trait", "rep")) |>
  select(id, first_value, second_value) |>
  head()
```

The first rows illustrate reproducibility in this software environment. For a
reproducible study, keep the input data, seed, and package version with your script.

## Assumptions and theory

This example uses one continuous trait, one additive layer, one simulation
replication, one record per individual, default mean zero, and a nondegenerate
genetic component. There is no selection, crossing, dominance, epistasis,
expression-mediated component, or threshold transformation.

### How the trait is built

Each selected marker contributes its coded genotype multiplied by its effect.
The package adds those contributions to obtain a genetic score for each
individual. It then shifts the scores to mean zero and rescales them so their
sample variance equals the requested allocation, here 0.5.

The residual represents variation outside this genetic component. By default,
the package draws normal random values, shifts them to mean zero, and rescales
them to the remaining variance, also 0.5. The phenotype is simply

$$
\text{phenotype} = \text{genetic value} + \text{residual}.
$$

Rescaling controls the spread of each component. It does not control whether an
individual with a high genetic value happens to receive a high residual too.
The detailed normalization formulas are explained in @sec-scaling-derivation.

### Why the realized ratio can differ from 0.5

Imagine two runs with exactly the same genetic values. In one run, individuals
with high genetic values happen to receive mostly positive residuals. This
reinforces their differences and increases the phenotype's spread. In another
run, those individuals happen to receive mostly negative residuals, reducing
the spread. This chance agreement or disagreement is measured by **sample
covariance**.

The exact accounting rule is

$$
\operatorname{Var}(y) = \operatorname{Var}(g) + \operatorname{Var}(e)
                         + 2\operatorname{Cov}(g,e).
$$ {#eq-sample-variance}

Here $y$ means phenotype, $g$ genetic value, and $e$ residual. `Var` measures
spread; `Cov` measures whether two sets of values tend to rise together. These
are sample quantities, calculated from the simulated individuals.

Our two variances are 0.5 each. Their sum is one, but the covariance term can
move phenotypic variance above or below one. Dividing genetic variance by this
changing total therefore need not give exactly 0.5.

The package labels this sample ratio `realized H²`. It is a diagnostic of one
simulation, not an estimated population heritability. In unusual samples it can
even exceed one; @sec-variance-derivation explains the algebra and gives an
example. This does not imply that population heritability exceeds one.

### Allow residual sample variance to fluctuate

Keep the default `residual_mode = "fixed"` when you want the residual component
to have the same sample variance in each run. Choose `"random"` when you want
ordinary normal sampling variation, as in AlphaSimR's standard residual draw.
In random mode, 0.5 describes the variance of the distribution supplying the
residuals. A finite sample from that distribution need not have variance exactly
0.5, and its mean need not be exactly zero.

[]{#first-random-residuals}

```{r}
#| label: first-random-residuals
random_sim <- simulate_phenotype(geno, seed = 901, residual_mode = "random") |>
  additive(prop = 0.5, n_qtn = 20)

random_values <- phenotypes_long(random_sim) |>
  left_join(genetic, by = "id") |>
  mutate(residual = value - genetic_value)

tibble(
  quantity = c("Residual mean", "Residual variance", "Genetic-residual covariance"),
  parameter = c(0, 0.5, 0),
  fixed = c(
    mean(values$residual), var(values$residual),
    cov(values$genetic_value, values$residual)
  ),
  random = c(
    mean(random_values$residual), var(random_values$residual),
    cov(random_values$genetic_value, random_values$residual)
  )
)
```

The same seed and genetic specification retain the same genetic values, which
is why we can reuse the `genetic` table. Random mode changes only how residuals
are generated. Both modes leave genetic-residual sample covariance free to vary.
For repeated sampling studies of estimators or confidence intervals, random mode
matches a model with independent normal errors. Fixed mode offers controlled
component variances for comparisons between simulation settings.

For multiple traits, `resid_cor` specifies their residual correlations. Random
mode draws correlated normal residuals without fixing their sample variances.
With `reps`, both modes divide residuals by the square root of the number of
records to represent an entry mean. A carried population trait remembers its
residual mode unless you explicitly choose another. For vQTLs, random mode also
lets the heterogeneous residual component fluctuate; the average of its
individual variance parameters equals the requested allocation.

### Should the residual covariance be forced to zero?

Usually there is no need to force it to zero. Independent residual generation
allows chance covariance in a finite sample; that is part of the sampling model.
One could instead adjust the residuals using the genetic values until their
sample covariance is zero, then rescale them. For this simple model that would
make the variance ratio exactly 0.5, but it would impose another constraint on
the sample. Such adjustment is a different simulation design and is not provided
by either residual mode here.

### Comparison with AlphaSimR

AlphaSimR 2.1.0 provides a useful comparison. Its ordinary `setPheno()` path
draws normal residuals and adds them to genetic values. It neither projects those
residuals off the genetic-value vector nor standardizes their realized sample
variance. Its `h2` and `H2` arguments calibrate the error-variance parameter using
founder-population variances. They therefore do not promise the requested sample
ratio in every simulated population. See the
[official setPheno documentation](https://gaynorr.github.io/AlphaSimR/reference/setPheno.html)
and the `addError()` implementation in
[phenotypes.R](https://github.com/gaynorr/AlphaSimR/blob/master/R/phenotypes.R).

| Ordinary single-trait path, one record | Residual sample variance fixed? | Genetic–residual sample covariance forced to zero? |
|---|---|---|
| simplePHENOTYPES `residual_mode = "fixed"` (default) | Yes | No |
| simplePHENOTYPES `residual_mode = "random"` | No | No |
| AlphaSimR 2.1.0 `setPheno()` | No | No |
| Hypothetical exact sample calibration | Yes | Yes |

The optional repository script `dev/book/compare_alphasimr.R` reproduces the
comparison without making AlphaSimR a dependency of this book. The matching residual-draw principle does not make the packages identical:
AlphaSimR and simplePHENOTYPES use different trait-calibration workflows. GxE and
other specialized model paths require separate interpretation.

## Implementation and practical limits

The scientific calculations above are R functions. QTN sampling and effect and
residual draws stay in R. The [implementation diagram](architecture.qmd) shows
where Rust participates in other workflows.

- Preserve genotype orientation and individual identifiers when importing your own data.
- A requested positive allocation needs a component with nonzero variation.
- Interpret per-marker `var_explained` values as marginal contributions; covariance
  among marker contributions can prevent their sum from equaling the layer share.
- This chapter's residual calculation does not apply unchanged to threshold traits,
  expression-mediated phenotypes, or other added components.

::: {.callout-tip title="Extending the maize example"}
This is an inbred panel. Before adding dominance, inspect heterozygosity: loci
without heterozygotes cannot express variation in a heterozygote-indicator column.
The high-MAF subset is a teaching dataset, not a universal marker-filtering rule.
:::

### Plant, animal, and human applications

The book's worked examples use observed genotype datasets from all three
contexts. Plant examples cover quantitative traits and breeding workflows on
this maize panel; the [prediction chapter](prediction.qmd) uses heterogeneous-stock
mice for relationships, prediction, and accuracy; and the
[binary phenotype chapter](binary.qmd) uses HapMap individuals to introduce
`liability_threshold()`, distinguishing latent liability from observed
case/control status. The [genotype input chapter](input.qmd) documents where
each dataset comes from and how it is prepared.

The common starting point is a user's genotype data. Dataset-specific features,
such as inbreeding, relatedness, and population structure, belong in the scientific
interpretation of each example.

## Explore the simulation

Change the seed and compare the tables again. The component allocations should
remain stable, while the sample covariance and realized ratio change. Then reduce
the number of causal markers to one and inspect the phenotype histogram.
Also repeat with `residual_mode = "random"` and watch the residual variance
change. These comparisons distinguish model specification from the empirical distribution of
one simulated population.
