# Maps, haplotypes, and founders {#sec-founders}

Suppose you want to simulate crosses, selfing, or doubled haploids from your
genotyped individuals. A dosage table is not enough: meiosis needs to know
which allele sits on which of the two homologous chromosomes and how far apart
markers are in recombination units. This chapter builds a `Population` from
real genotypes, inspects its haplotypes, supplies a genetic map when the data
have none, and generates founders from a coalescent model when you have no
genotypes at all.

[]{#founders-setup}

```{r}
#| label: founders-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-founders-example}

### Load the packages

[]{#founders-library}

```{r}
#| label: founders-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

### Founders from the maize panel

The bundled maize data carry a `cm` column, so `as_population()` can use them
directly. The print method summarizes the individuals, markers, genetic map,
and pedigree status.

[]{#founders-maize}

```{r}
#| label: founders-maize
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
maize <- as_numeric(SNP55K_maize282_maf04, to_r = TRUE, to_file = FALSE,
                    verbose = FALSE)

founders <- as_population(maize)
founders
n_individuals(founders)
```

Every individual is recorded as a pedigree founder in generation 0. The
population keeps the marker map and a pedigree table; both are plain data
frames you can inspect.

[]{#founders-components}

```{r}
#| label: founders-components
founders$map |> head(3)
founders$pedigree |> select(id, mother, father, generation, design) |> head(3)
```

### Dosages and haplotypes

`dosages()` returns the markers-by-individuals `-1/0/1` matrix.
`haplotypes()` returns the two strands, called `cis` and `trans`, with `1`
marking the counted allele. The dosage is the sum of the two strands minus
one, and we check that identity on the whole panel.

[]{#founders-haplotypes}

```{r}
#| label: founders-haplotypes
d <- dosages(founders)
h <- haplotypes(founders)

d[1:3, 1:4]
h$cis[1:3, 1:4]

tibble(
  quantity = c("markers", "individuals", "dosage equals cis + trans - 1"),
  value = c(nrow(d), ncol(d), identical(h$cis + h$trans - 1L, d))
)
```

The strands have no parental meaning. Meiosis treats them symmetrically, so
either assignment is valid as long as both strands of an individual stay
together.

### Selecting individuals

Subsetting with `[` keeps the selected individuals in the order given, by
name, position, or logical mask. A repeated subscript duplicates an
individual: the copy receives a unique id but shares the original's pedigree
key, so the pedigree still records one individual.

[]{#founders-subset}

```{r}
#| label: founders-subset
parents <- founders[c("33-16", "38-11")]
parents

twice <- founders[c(1, 1, 2)]
twice$ids
nrow(twice$pedigree)
```

`as_population()` accepts the same selection through its `individuals`
argument, which avoids phasing the full panel when only a few lines are
needed.

### Outbred founders and the phasing guess

The maize lines are almost completely homozygous, so their haplotypes are
nearly determined by the dosages. The human panel is different. We load it
with the helper described in [Genotype input and quality control](input.qmd),
give it a synthetic map, and phase the first 50 individuals.

[]{#founders-human}

```{r}
#| label: founders-human
source("_data.R")
human <- book_human()
human$cm <- synthetic_map(human$chr, human$pos, cm_per_mb = 1)

people <- as_population(human, individuals = 1:50)
people

strands <- haplotypes(people)
het <- dosages(people) == 0

tibble(
  heterozygous_calls = sum(het),
  cis_carries_counted_allele = sum(strands$cis[het] == 1),
  trans_carries_counted_allele = sum(strands$trans[het] == 1)
)
```

Every heterozygote has been placed with the counted allele on the first
strand. That is a rule, not an inference: the dosage table does not say which
allele sits on which chromosome, and the package does not import external
phase. For these founders, all heterozygous sites on a chromosome are in
perfect coupling, so their first-generation gametes carry more linkage
disequilibrium than real gametes would. The founders' own dosages are exact,
but every gamete they produce, from the first cross on, carries this
coupling-phase linkage disequilibrium. Use `population_from_haplotypes()` with
phased data whenever meiosis of outbred founders matters.

### Founders from known haplotypes

When the phase is known, from statistically phased data or another
simulator, `population_from_haplotypes()` builds the population from the two
strand matrices directly. Here we rebuild the human founders from their own
strands and confirm that the dosages agree, then construct a two-individual
example by hand where one individual is heterozygous at two linked markers
with a specified phase.

[]{#founders-from-haplotypes}

```{r}
#| label: founders-from-haplotypes
rebuilt <- population_from_haplotypes(strands$cis, strands$trans, people$map,
                                      pool = "hapmap")
identical(dosages(rebuilt), dosages(people))

map <- data.frame(snp = paste0("m", 1:4), chr = 1, pos = c(100, 200, 300, 400),
                  cm = c(0, 10, 20, 30))
cis <- cbind(ind1 = c(1, 1, 0, 0), ind2 = c(0, 0, 0, 0))
trans <- cbind(ind1 = c(1, 0, 1, 0), ind2 = c(0, 0, 0, 0))
known <- population_from_haplotypes(cis, trans, map)
dosages(known)
haplotypes(known)$cis
```

Individual `ind1` is heterozygous at `m2` and `m3`, with the counted allele
at `m2` on the first strand and at `m3` on the second: repulsion phase.
`as_population()` would have put both on the first strand. The
[meiosis chapter](cross.qmd) shows how that difference changes the gametes.

### A map for data without one

The mouse genotypes carry physical positions but no genetic map, so
`as_population()` refuses them until `cm` is filled. `synthetic_map()`
converts base pairs into a plausible centiMorgan map. It starts each
chromosome at 0 and is non-decreasing within a chromosome.

[]{#founders-mice-map}

```{r}
#| label: founders-mice-map
mice <- book_mice()
mice$cm <- synthetic_map(mice$chr, mice$pos)

tibble(chr = mice$chr, pos = mice$pos, cm = mice$cm) |>
  group_by(chr) |>
  summarise(markers = n(), span_mb = max(pos) / 1e6,
            length_cm = max(cm), .groups = "drop") |>
  head(5)

mouse_founders <- as_population(mice, individuals = 1:100)
mouse_founders
```

::: {.callout-note title="This map is a model"}
The `cm` values are not estimates of mouse recombination distances. They give
the simulation a recombination landscape with the right order of magnitude
and centromeric suppression. If you have a measured map for your markers,
put it in `cm` instead.
:::

The default rate of 0.73 cM per Mb is a maize-scale value. The shape of the
map comes from a local rate that dips around the centromere, which defaults
to the midpoint of each chromosome's span. Setting `suppression = 0` gives a
uniform map, and `total_cm` fixes chromosome lengths directly.

[]{#founders-map-shape}

```{r}
#| label: founders-map-shape
#| fig-cap: "Synthetic genetic map for mouse chromosome 1 under the default centromeric suppression and under a uniform rate. The physical positions are real; the cM values are a model."
on1 <- mice$chr == "1"
uniform <- synthetic_map(mice$chr, mice$pos, suppression = 0)
plot(mice$pos[on1] / 1e6, mice$cm[on1], type = "l", lwd = 2,
     xlab = "Physical position on chromosome 1 (Mb)",
     ylab = "Genetic position (cM)")
lines(mice$pos[on1] / 1e6, uniform[on1], lwd = 2, lty = 2, col = "grey50")
legend("topleft", c("centromere-suppressed", "uniform"),
       lwd = 2, lty = c(1, 2), col = c("black", "grey50"), bty = "n")
```

[]{#founders-map-options}

```{r}
#| label: founders-map-options
fixed <- synthetic_map(mice$chr, mice$pos, total_cm = 80)
tibble(chr = mice$chr, default = mice$cm, uniform = uniform, fixed = fixed) |>
  group_by(chr) |>
  summarise(across(c(default, uniform, fixed), max), .groups = "drop") |>
  head(4)
```

### Founders without genotypes

`founders_coalescent()` simulates founder haplotypes under a sequential
Markov coalescent with recombination and mutation, using the same demographic
presets that AlphaSimR's `runMacs()` builds. The result is a `Population`
whose counted allele is the derived allele. The seed makes it reproducible.

[]{#founders-coalescent}

```{r}
#| label: founders-coalescent
generic <- founders_coalescent(50, n_chr = 2, seg_sites = 500, seed = 1)
generic

again <- founders_coalescent(50, n_chr = 2, seg_sites = 500, seed = 1)
identical(dosages(again), dosages(generic))

attr(generic, "coalescent")$n_total
```

`seg_sites` keeps a uniform random subset of the segregating sites the
coalescent produced; the attribute records how many existed before
subsetting. A constant-size neutral coalescent sample has a spectrum skewed
toward rare variants. The `"GENERIC"` preset instead follows AlphaSimR's
demography, a recent strong bottleneck, which depletes rare variants, so the
median minor allele frequency in the summary is well above what a
constant-size sample would give.

[]{#founders-coalescent-spectrum}

```{r}
#| label: founders-coalescent-spectrum
x <- dosages(generic)
p <- rowMeans(x + 1) / 2
tibble(
  quantity = c("minimum MAF", "median MAF", "fraction heterozygous"),
  value = c(min(pmin(p, 1 - p)), median(pmin(p, 1 - p)), mean(x == 0))
)
```

Species presets change the chromosome length, mutation and recombination
parameters, and the population-size history. `inbred = TRUE` duplicates each
simulated haplotype so every founder is homozygous, the usual starting point
for a line-breeding program. `split` separates the founders into two
subpopulations with no migration for a number of generations before the
present.

[]{#founders-coalescent-presets}

```{r}
#| label: founders-coalescent-presets
maize_like <- founders_coalescent(50, n_chr = 2, seg_sites = 500,
                                  species = "MAIZE", seed = 1)
maize_like

lines <- founders_coalescent(20, seg_sites = 300, inbred = TRUE, seed = 2)
all(dosages(lines) != 0)

split_pop <- founders_coalescent(40, seg_sites = 300, split = 50, seed = 3)
xs <- dosages(split_pop)
p1 <- rowMeans(xs[, 1:20] + 1) / 2
p2 <- rowMeans(xs[, 21:40] + 1) / 2
tibble(mean_absolute_frequency_difference = mean(abs(p1 - p2)))
```

The last table is a descriptive statistic of the two halves' allele
frequencies, not an estimator of a population parameter.

## Assumptions and theory

### Phasing and dosage rules

For an individual's dosage $x_j \in \{-1, 0, 1\}$ at marker $j$,
`as_population()` sets

$$
\text{cis}_j = \mathbf{1}[x_j \ge 0], \qquad
\text{trans}_j = \mathbf{1}[x_j > 0],
$$ {#eq-phasing-rule}

so homozygotes are determined and every heterozygote has its counted allele
on the first strand. This is a package convention; the equation map records
it for `as_population()`. The inverse,

$$
x_j = \text{cis}_j + \text{trans}_j - 1,
$$ {#eq-dosage-rule}

is the isqg `genotype_num` convention that `dosages()` and the Rust genome
module implement. The identity check above executes it.

### The synthetic map model

`synthetic_map()` assigns each physical position $p$ on a chromosome of span
$L$ a local rate

$$
r(p) = 1 - s \exp\!\left(-\tfrac{1}{2}\left(\frac{p - c}{wL}\right)^2\right),
$$ {#eq-synthetic-map}

where $c$ is the centromere position, $s$ the suppression strength and $w$ the
relative width of the suppressed region. The rate is integrated across marker
intervals and rescaled so that the chromosome reaches its target length,
either `total_cm` or `cm_per_mb` times the span in Mb. Because $r(p)$ is
positive for $s < 1$, cM is a monotone function of bp.

The Gaussian form is a phenomenological package choice. The package
documentation cites Bauer et al. (2013) for the pericentromeric suppression
pattern in maize and the genome-wide scale the defaults approximate, and
states that the functional form and parameters are not taken from that
source. The map's length matters for what follows: the expected number of
crossovers on a chromosome equals its last map position in Morgans, as the
[meiosis chapter](cross.qmd) explains.

### The coalescent founders

`founders_coalescent()` follows the model documented for the function: time
in units of $4N_0$ generations, per-chromosome $\theta = 4N_0\mu$ and
$\rho = 4N_0 r$, a piecewise-constant population-size history relative to
$N_0$, infinite-sites mutation, and a sequential Markov coalescent (SMC')
walk along the chromosome. The genetic map is linear in physical position and
starts at 0 cM at the first segregating site. The species presets reproduce
the demographic commands that AlphaSimR 2.1.0's `runMacs()` builds; their
parameter values are AlphaSimR's choices, not estimates made in this package.
The package references are Marjoram and Wall (2006), Chen, Marjoram and Wall
(2009), Hudson (2002), and Gaynor, Gorjanc and Hickey (2021), as given in the
function documentation. This book does not verify the coalescent algorithm
against those sources; the allele-frequency summary above is a sanity check
of the output, not a validation of the model.

### Randomness and reproducibility

This function is the one place where random numbers are drawn in Rust. Its
pseudorandom generator is seeded from R, so a `seed` reproduces the founders
and the caller's R random state is restored afterwards. Meiosis, QTN sampling,
and effect and residual draws remain in R, as described in the
[implementation overview](architecture.qmd).

## Implementation and practical limits

`as_population()` and `population_from_haplotypes()` are R functions in the
crossing module; `dosages()` combines the strands through the Rust genome
code; `founders_coalescent()` calls `coalescent_chromosome_core()` per
chromosome and assembles the result with `population_from_haplotypes()`. The
[developer index](developer-index.qmd) gives current locations.

- `cm` must be in centiMorgans, non-decreasing within a chromosome, with no
  missing values; a map whose largest value is at most 5 across 20 or more
  markers draws a warning because it looks like Morgans.
- Chromosomes are processed in a canonical order (numeric labels first), so
  `chr` stored as text or as numbers gives the same seeded progeny.
- Populations built from separately converted panels can count different
  alleles as `1`. Convert jointly or carry the `counted` column, as the
  [input chapter](input.qmd) describes; `cross()` and `c.Population()` stop
  when the records disagree.
- Phasing a large outbred panel is fast, but the phase is a guess. Use
  `population_from_haplotypes()` when real phase information exists.
- `founders_coalescent()` can return fewer segregating sites than requested
  only by stopping with an error; ask for fewer sites or more individuals.
- A filtered population (see `filter_geno()`) has a different marker set from
  the unfiltered one and cannot be crossed or pooled with it.
