# Genotype input and quality control {#sec-input}

Suppose you have a genotype file in one of the usual formats and want to use it
as the basis for a simulation. You need to know which allele the numeric
coding counts, what happened to missing calls, and which markers are suitable
for the genetic model you plan to simulate. This chapter converts files and
in-memory tables with `as_numeric()`, checks orientation and missingness, and
applies marker filters with `filter_geno()`, using the bundled maize panel and
a real human dataset distributed with the SNPRelate package.

[]{#input-setup}

```{r}
#| label: input-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-input-example}

### Load the packages

[]{#input-library}

```{r}
#| label: input-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
library(readr)
```

### A small HapMap file, written and read back

To see exactly what `as_numeric()` does, start with a table small enough to
check by hand. The HapMap text format has eleven metadata columns followed by
one column per individual, with diploid calls such as `AG` and `NN` for
missing. We build such a table, write it to the `results` folder, and read the
file back as a reader would read their own file.

[]{#input-toy-hapmap}

```{r}
#| label: input-toy-hapmap
hmp <- tibble(
  `rs#` = c("m1", "m2", "m3", "m4"),
  alleles = c("A/G", "C/T", "G/T", "A/C"),
  chrom = c(1, 1, 2, 2),
  pos = c(100, 500, 200, 900),
  strand = "+", `assembly#` = NA, center = NA, protLSID = NA,
  assayLSID = NA, panelLSID = NA, QCcode = NA,
  L1 = c("AA", "CC", "GT", "AA"),
  L2 = c("AG", "TT", "TT", "CC"),
  L3 = c("GG", "CT", "NN", "AC"),
  L4 = c("AA", "CC", "GG", "CC"),
  L5 = c("GG", "CC", "TT", "AA"),
  L6 = c("AA", "TT", "GT", "CC")
)

dir.create("results", showWarnings = FALSE)
write_tsv(hmp, "results/toy.hmp.txt", na = "NA")

toy <- as_numeric("results/toy.hmp.txt", to_r = TRUE, to_file = FALSE,
                  verbose = FALSE)
toy
```

The format is detected from the file contents. The result has the five
metadata columns `snp`, `allele`, `chr`, `pos`, `cm`, then one integer column
per individual. `cm` is `NA` because HapMap files carry no genetic map; see
[Maps, haplotypes, and founders](founders.qmd) for how to supply one.

### Which allele is coded 1

Compare marker `m3` with its calls. Allele `T` occurs more often than `G`
across the six individuals, so `T` is the reference: `TT` is `1`, `GT` is
`0`, `GG` is `-1`, and `NN` is `NA`. The counted allele at each marker is
recorded as an attribute of the returned table.

[]{#input-counted}

```{r}
#| label: input-counted
attr(toy, "counted_allele")
```

By default (`method = "frequency"`) the most frequent allele of **this**
dataset becomes `1`. Two panels converted separately can therefore count
different alleles at the same marker. To fix the orientation yourself, pass a
reference allele per marker; here we force the first-listed allele to be the
reference, which flips `m3` and `m4` relative to the frequency-based result.

[]{#input-reference}

```{r}
#| label: input-reference
by_reference <- as_numeric("results/toy.hmp.txt", to_r = TRUE, to_file = FALSE,
                           verbose = FALSE, method = "reference",
                           ref_allele = c("A", "C", "G", "A"))
by_reference
```

The `counted` column makes the record durable. It is written into text files
and moves with its row when the table is subset. The attribute is kept
unchanged by base row subsetting, so after subsetting or reordering it no
longer lines up with the rows.

[]{#input-counted-column}

```{r}
#| label: input-counted-column
as_numeric("results/toy.hmp.txt", to_r = TRUE, to_file = FALSE,
           verbose = FALSE, counted_column = TRUE)
```

### Alternative codings and missing calls

`code_as = "012"` counts copies of the reference allele instead of centering
on the heterozygote. `impute` replaces missing calls; `"Middle"` sets them to
the heterozygote, `"Major"` to the homozygote coded 1 (the counted allele) and
`"Minor"` to the homozygote coded -1. Imputation follows orientation, so with
`method = "reference"` the counted allele, and therefore the `"Major"` fill,
need not be the more frequent allele.

[]{#input-coding}

```{r}
#| label: input-coding
as_numeric("results/toy.hmp.txt", to_r = TRUE, to_file = FALSE,
           verbose = FALSE, code_as = "012")

as_numeric("results/toy.hmp.txt", to_r = TRUE, to_file = FALSE,
           verbose = FALSE, impute = "Major")
```

Imputed values are not observations. For a simulation they only decide which
genotype class the individual is assigned at that marker; nothing in the
output distinguishes an imputed call from a real one.

### Writing the numeric file

With `to_file = TRUE` the same conversion is written to disk. Reading that file
again returns the same table, which is how a converted panel can be reused
across sessions without repeating the conversion.

[]{#input-to-file}

```{r}
#| label: input-to-file
as_numeric("results/toy.hmp.txt", to_r = FALSE, to_file = TRUE,
           file_name = "results/toy_numeric.txt", verbose = FALSE)

from_file <- as_numeric("results/toy_numeric.txt", to_r = TRUE,
                        to_file = FALSE, verbose = FALSE)

tibble(
  same_values = identical(from_file[, -(1:5)], toy[, -(1:5)]),
  counted_allele_kept = !is.null(attr(from_file, "counted_allele"))
)
```

The genotype values are the same, but the counted-allele attribute is gone:
a plain text file has nowhere to store it, which is what `counted_column`
is for. An already numeric input cannot be re-oriented or imputed;
`as_numeric()` stops with an error if you ask for it. Keep the original
allele calls when you may need a different orientation later.

### Other formats

VCF, GDS and PLINK bed/ped files follow the same call. For VCF, only complete
biallelic diploid calls are used; haploid and partially missing calls are set
to missing with a warning. Among the external formats, only PLINK `.bim`/`.map`
files supply a genetic map; a table already in the package's numeric format
keeps whatever `cm` it has. These recipes are not executed here
because they refer to reader-supplied files.

```{r}
#| eval: false
geno <- as_numeric("path/to/study.vcf", to_r = TRUE, to_file = FALSE)
geno <- as_numeric("path/to/study.bed", to_r = TRUE, to_file = FALSE)
geno <- as_numeric("path/to/study.gds", to_r = TRUE, to_file = FALSE)
```

### A real human dataset

The SNPRelate package ships a HapMap Phase II subset: 279 individuals from the
CEU, YRI, HCB and JPT panels, with 9,088 SNPs after LD pruning. It is a GDS
file, which `as_numeric()` reads directly. The first attempt fails.

[]{#input-gds-error}

```{r}
#| label: input-gds-error
#| error: true
library(SNPRelate)
src <- snpgdsExampleFileName()
as_numeric(src, to_r = TRUE, to_file = FALSE, verbose = FALSE)
```

Downstream functions index markers by name, so duplicated marker IDs are an
error rather than a warning. The file uses `---` for SNPs without an rs
identifier and lists one rs number twice. We inspect the problem with the GDS
tools and write a cleaned copy that keeps only uniquely named SNPs.

[]{#input-gds-clean}

```{r}
#| label: input-gds-clean
gds <- snpgdsOpen(src)
ids <- read.gdsn(index.gdsn(gds, "snp.id"))
rs <- read.gdsn(index.gdsn(gds, "snp.rs.id"))
snpgdsClose(gds)

tibble(id = ids, rs = rs) |>
  filter(rs == "---" | rs %in% rs[duplicated(rs)]) |>
  count(rs)

keep <- ids[rs != "---" & !(rs %in% rs[duplicated(rs)])]
clean <- "results/hapmap_clean.gds"
if (file.exists(clean)) file.remove(clean)
snpgdsCreateGenoSet(src, clean, snp.id = keep, verbose = FALSE)
```

Now the conversion succeeds. We first convert without imputation to measure
the missing calls, then convert again with `impute = "Middle"`, which is the
version later chapters use.

[]{#input-human}

```{r}
#| label: input-human
human_raw <- as_numeric(clean, to_r = TRUE, to_file = FALSE, verbose = FALSE)
calls <- human_raw[, -(1:5)]

tibble(
  markers = nrow(human_raw),
  individuals = ncol(calls),
  missing_calls = sum(is.na(calls)),
  missing_fraction = mean(is.na(calls)),
  markers_with_missing = sum(rowSums(is.na(calls)) > 0)
)

human <- as_numeric(clean, to_r = TRUE, to_file = FALSE, verbose = FALSE,
                    impute = "Middle")
sum(is.na(human[, -(1:5)]))
```

The file `_data.R` in the book source wraps exactly these steps in
`book_human()`, so later chapters can load the same object with one call.
Setting a missing call to the heterozygote is a neutral choice for an additive
model, but it creates heterozygotes that were never observed. There is no
call-rate filter in the package; if missingness matters for your study,
remove individuals or markers before conversion.

### Genotype classes in an inbred and an outbred panel

The maize panel is an inbred association panel; the human panel is outbred.
The mean frequency of each genotype class shows the difference directly.

[]{#input-classes}

```{r}
#| label: input-classes
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
maize <- as_numeric(SNP55K_maize282_maf04, to_r = TRUE, to_file = FALSE,
                    verbose = FALSE)

class_frequencies <- function(geno) {
  x <- as.matrix(geno[, -(1:5)])
  tibble(
    major_homozygote = mean(x == 1),
    heterozygote = mean(x == 0),
    minor_homozygote = mean(x == -1)
  )
}

bind_rows(maize = class_frequencies(maize),
          human = class_frequencies(human), .id = "data")
```

Heterozygotes are nearly absent in maize. A dominance layer, which assigns an
effect to the heterozygote class, would have almost nothing to act on there.
The maize panel was also filtered to minor allele frequency above 0.4 to keep
the bundled data small, so its allele-frequency spectrum is far from a typical
genotyping panel.

### Minor allele frequency and heterozygosity filters

`filter_geno()` applies monomorphic removal, minor allele frequency bounds, a
heterozygosity rule, and LD pruning, in that order. With `verbose = TRUE` it
reports how many markers each step removed; the tables below count them.

[]{#input-filter-maf}

```{r}
#| label: input-filter-maf
human_common <- filter_geno(human, maf_above = 0.05)

maize_hets <- filter_geno(maize, hets = "include")

tibble(
  panel = c("human", "maize"),
  before = c(nrow(human), nrow(maize)),
  after = c(nrow(human_common), nrow(maize_hets))
)
```

`hets = "include"` keeps only markers with at least one heterozygote. On the
maize panel it removes about half the markers, which is the first thing to
check before simulating dominance on inbred lines. `hets = "remove"` does the
opposite and is useful for strictly homozygous material.

### LD pruning

Three pruners follow PLINK's definitions. `indep_pairwise` drops one marker of
every pair whose composite genotype $r^2$ exceeds the threshold inside a
sliding window; `indep_pairphase` uses the haplotypic $r^2$ instead;
`indep` removes markers by variance inflation factor. The window is counted
in markers unless `window_unit = "kb"`.

[]{#input-filter-ld}

```{r}
#| label: input-filter-ld
pruned <- filter_geno(human, maf_above = 0.05,
                      indep_pairwise = c(50, 5, 0.2))

pruned_kb <- filter_geno(human, maf_above = 0.05,
                         indep_pairwise = c(500, 50, 0.2),
                         window_unit = "kb")

vif_pruned <- filter_geno(human, maf_above = 0.05, indep = c(50, 5, 2))

tibble(
  pruner = c("MAF filter only", "pairwise, 50 markers", "pairwise, 500 kb", "VIF, 50 markers"),
  markers = c(nrow(human_common), nrow(pruned), nrow(pruned_kb), nrow(vif_pruned))
)
```

The human example data were already LD pruned by their distributors. The
marker-count windows still remove roughly an eighth of the common markers,
while the 500 kb window removes very few, because the markers are sparse
enough that few pairs fall within one window at this threshold. Window choice
matters as much as the $r^2$ threshold. The
phased pruner is slower because it solves a two-locus maximum-likelihood
problem for each pair.

[]{#input-filter-pairphase}

```{r}
#| label: input-filter-pairphase
phased <- filter_geno(human, maf_above = 0.05,
                      indep_pairphase = c(50, 5, 0.2))
```

### Haplotype blocks

`blocks = TRUE` defines haplotype blocks by the method of Gabriel et al.
(2002), as PLINK implements it, and keeps one tag marker per block. It needs
physical positions. We apply it to one maize chromosome, where linkage
disequilibrium is extensive.

[]{#input-filter-blocks}

```{r}
#| label: input-filter-blocks
maize_chr10 <- maize |> filter(chr == "10")
tagged <- filter_geno(maize_chr10, blocks = TRUE, block_max_kb = 500)

tibble(markers = nrow(maize_chr10), tag_markers = nrow(tagged))
```

The block definitions reproduce PLINK's; collapsing each block to its
highest-MAF marker is a package convenience, not part of the block method.

### Filtering other genotype objects

The same function accepts an individuals-by-markers matrix coded `-1/0/1` and
a `Population` from `as_population()`. A filtered population keeps its map,
individuals and pedigree, but it has a different marker set from the original,
so filter founders once before crossing them.

[]{#input-filter-population}

```{r}
#| label: input-filter-population
pop <- as_population(maize, individuals = 1:20)
filter_geno(pop, maf_above = 0.45, verbose = FALSE)
```

A matrix input uses `code_as` to declare whether it is `-1/0/1` or `0/1/2`,
because a marker with only `0` and `1` present is ambiguous between the two.

## Assumptions and theory

### Allele frequency and minor allele frequency

For a marker with $n$ called individuals and dosage $d_i \in \{0, 1, 2\}$
counting copies of one allele,

$$
p = \frac{\sum_i d_i}{2n}, \qquad \text{MAF} = \min(p, 1 - p).
$$ {#eq-maf}

Missing calls are excluded from $n$. These are package definitions following
PLINK's `--maf` semantics; the equation map records them for `filter_geno()`.
The `maf_above` and `maf_below` bounds are inclusive. Heterozygosity filters
use the count of heterozygotes, not a rate; there is no heterozygosity-rate
threshold and no Hardy-Weinberg test in the package.

### The orientation rule

With `method = "frequency"` the package flips a marker so that the more
frequent homozygote class is coded `1`. Heterozygotes cancel in that
comparison. When the two homozygote classes are exactly equally frequent, no
flip occurs and the first-listed allele of the input stays the reference. This
tie rule is a package convention, and it is applied identically by every
reader, so a file path and the same table in memory give the same dosages.

### Composite $r^2$

For two markers with dosage vectors $x$ and $y$ over $n$ individuals, the
composite genotype $r^2$ used by `indep_pairwise` and `indep` is the squared
Pearson correlation of the dosages,

$$
r^2 = \frac{\left(n \sum xy - \sum x \sum y\right)^2}
           {\left(n \sum x^2 - (\sum x)^2\right)\left(n \sum y^2 - (\sum y)^2\right)}.
$$ {#eq-composite-r2}

This is PLINK's `--indep-pairwise` quantity, computed from integer sums. When
a pair has missing calls, the package falls back to a pairwise-complete
correlation, which is not claimed to match PLINK's missing-data path. A pair is
pruned when $r^2$ exceeds the threshold, and the marker with the lower MAF is
dropped; ties keep the earlier marker.

### Haplotypic $r^2$ and the variance inflation factor

`indep_pairphase` estimates haplotype frequencies for each pair by resolving
the double-heterozygote phase with PLINK's exact two-locus maximum-likelihood
solver, then computes

$$
r^2 = \frac{D^2}{p_1 q_1 p_2 q_2}, \qquad D = f_{11} - f_{1\cdot} f_{\cdot 1},
$$ {#eq-hap-r2}

where $f_{11}$ is the estimated frequency of the haplotype carrying the first
allele at both loci and $p_1, q_1, p_2, q_2$ are the allele frequencies. The
equation map cites Hill and Robertson (1968) for this quantity, with the page
of the equation not verified, and the cubic solution of Gaunt, Rodríguez and
Day (2007) for the phase step, also page unverified.

The VIF pruner inverts the correlation matrix $R$ of the markers in a window,
takes $\text{VIF}_k = [R^{-1}]_{kk}$, and removes the largest-VIF marker while
any VIF exceeds the threshold. This follows PLINK's `--indep`, attributed to
Purcell et al. (2007). The treatment of singular windows is a package
implementation detail that the equation map records as unverifiable against
PLINK's source.

### What the filters do not do

- No call-rate filter and no Hardy-Weinberg test exist; if your study needs
  them, apply them before conversion.
- LD is computed from every individual in the object, including progeny in a
  crossing population. PLINK uses founders only. Pass the founders if you want
  to reproduce a PLINK pruning.
- Pruning changes which markers can become causal in a later simulation; it
  does not change the genotypes of the markers that remain.

## Implementation and practical limits

`as_numeric()` dispatches on the detected format and calls a Rust kernel,
`numericalize_core()`, for the numeric coding and imputation; orientation is
decided in R by `compute_flip()`. `filter_geno()` is R code; the pruners
`.ld_sweep()`, `.plink_hap_rsq()` and `.gabriel_blocks()` port PLINK's
algorithms. The [developer index](developer-index.qmd) lists their locations.

- Marker and individual identifiers must be unique. Fix them in the source
  file or, as above, subset the file before conversion.
- The `counted_allele` attribute is not subset with the table rows, so it
  becomes stale after subsetting or reordering. Use `counted_column = TRUE` when panels converted separately will be
  crossed or pooled.
- Convert panels that will be crossed **jointly**, or give each the same
  `ref_allele`, so that `1` means the same allele in both.
- A numeric table cannot be re-oriented or imputed. Keep the allele-coded
  source.
- `window_unit = "kb"` and `blocks = TRUE` need `chr` and `pos` for every
  marker.

::: {.callout-tip title="Which panel for which model"}
Inbred lines suit additive models and crossing designs; dominance needs
heterozygotes, which arise naturally in the progeny of crosses or in an outbred
sample such as the human panel. Check the genotype class table before
choosing a model rather than after inspecting a disappointing variance budget.
:::
