# Legacy interface and migration {#sec-legacy}

Scripts written for the original package call `create_phenotypes()`, the
function described in Fernandes and Lipka (2020). That function still ships,
unchanged in signature and in its random-number arithmetic, so old scripts keep
producing their old numbers. It is not a wrapper around the composable grammar
of the other chapters: the two implementations coexist and share no code on the
simulation path. This chapter runs the legacy function on the maize data, shows
the files it writes, places the equivalent grammar call next to each legacy
call, and explains what "equivalent" does and does not mean.

[]{#legacy-setup}

```{r}
#| label: legacy-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-legacy-example}

[]{#legacy-library}

```{r}
#| label: legacy-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
library(readr)
```

### A single additive trait, the legacy way

The legacy function takes the genotype object directly, writes its results to
disk, and optionally returns them to R with `to_r = TRUE`. It needs a home
directory and creates a run folder inside it; use an absolute path for
`home_dir`, because the function changes the working directory while it runs
and resolves a relative `home_dir` against itself. The call below reproduces
the first example of the original vignette: three additive markers, a large
effect on the first one, a geometric series on the rest, heritability 0.7, and
two replications.

[]{#legacy-single}

```{r}
#| label: legacy-single
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
dir.create("results/legacy", showWarnings = FALSE, recursive = TRUE)
home <- normalizePath("results/legacy")

v1_single <- create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  add_QTN_num = 3,
  add_effect = 0.2,
  big_add_QTN_effect = 0.9,
  rep = 2,
  h2 = 0.7,
  model = "A",
  seed = 901,
  home_dir = home,
  output_dir = "single",
  to_r = TRUE,
  quiet = TRUE,
  verbose = FALSE
)

head(v1_single)
list.files(file.path(home, "single"))
```

For a single trait the returned data frame is wide: one row per individual and
one column per replication. The run folder holds the phenotypes, the selected
markers, the genetic values, and a log. The log records the master seed and
the sample heritability averaged over replications:

[]{#legacy-files}

```{r}
#| label: legacy-files
read_tsv(file.path(home, "single", "Additive_QTNs.txt"),
         show_col_types = FALSE)

read_lines(file.path(home, "single", "Log_Sim.txt")) |>
  tail(8) |>
  cat(sep = "\n")
```

The `add_eff_t1` column shows the effect series: the large effect on the first
marker, then the geometric series `0.2`, `0.2^2`. This is where the two
interfaces first differ. The legacy effect series is the final per-marker
effect; the grammar scales a whole layer to a requested share, as the
[first chapter](first-simulation.qmd) explains.

### The same design in the grammar

In the grammar, heritability is a sum of layer proportions rather than an
argument, and there is no `big_add_QTN_effect`; a large-effect marker is
expressed through an explicit effect series. The phenotypes are returned as an
object and written only on request.

[]{#legacy-single-v2}

```{r}
#| label: legacy-single-v2
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)

v2_single <- simulate_phenotype(geno, n_reps = 2, seed = 901) |>
  additive(prop = 0.7, n_qtn = 3, effect = c(0.9, 0.2, 0.04))
v2_single
```

### No bit-parity between the interfaces

The two calls use the same seed. Their numbers are nevertheless different,
because the legacy function keeps its own seed arithmetic under
`RNGversion("3.5.1")` while the grammar derives sub-seeds for each layer and
replication. The decision that the grammar owes the legacy function no
bit-parity is recorded in the repository's decision log (DECISION-009).

[]{#legacy-parity}

```{r}
#| label: legacy-parity
v2_wide <- phenotypes_wide(v2_single) |>
  filter(rep == 1) |>
  select(id, grammar = Trait_1)

v1_single |>
  select(id = taxa, legacy = rep1) |>
  left_join(v2_wide, by = "id") |>
  head()

v1_markers <- read_tsv(file.path(home, "single", "Additive_QTNs.txt"),
                       show_col_types = FALSE)$snp
tibble(legacy = v1_markers, grammar = qtn_table(v2_single)$snp)
```

Different markers are drawn and different values result. What the two
interfaces share is the design: the realized genetic-to-phenotypic variance
ratio of each run is close to the requested 0.7. The legacy log above and the
grammar's print method both report the mean of the per-replication ratios,
which is why the printed grammar value differs from the single replications in
the table below. Both are sample diagnostics in the sense of the
[first chapter](first-simulation.qmd#sec-first-example).

[]{#legacy-ratio}

```{r}
#| label: legacy-ratio
v1_genetic <- read_tsv(file.path(home, "single", "Genetic_values.txt"),
                       show_col_types = FALSE)$Trait_1
v2_genetic <- genetic_values(v2_single)[, 1]

tibble(
  interface = c("legacy, rep 1", "legacy, rep 2", "grammar, rep 1", "grammar, rep 2"),
  parameter = 0.7,
  simulated = c(
    var(v1_genetic) / var(v1_single$rep1),
    var(v1_genetic) / var(v1_single$rep2),
    var(v2_genetic) / var(filter(phenotypes_wide(v2_single), rep == 1)$Trait_1),
    var(v2_genetic) / var(filter(phenotypes_wide(v2_single), rep == 2)$Trait_1)
  )
)
```

### Pleiotropic traits with dominance

The legacy multi-trait call takes one effect size per trait and one
heritability per trait. `QTN_variance = TRUE` adds per-marker variance files.
The returned data frame is long by replication, with one column per trait.

[]{#legacy-pleiotropic}

```{r}
#| label: legacy-pleiotropic
v1_pleio <- create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  ntraits = 3,
  architecture = "pleiotropic",
  model = "AD",
  add_QTN_num = 3,
  dom_QTN_num = 4,
  add_effect = c(0.04, 0.2, 0.1),
  dom_effect = c(0.04, 0.2, 0.1),
  big_add_QTN_effect = c(0.3, 0.3, 0.3),
  h2 = c(0.2, 0.4, 0.4),
  rep = 2,
  seed = 10,
  QTN_variance = TRUE,
  home_dir = home,
  output_dir = "pleiotropic",
  to_r = TRUE,
  quiet = TRUE,
  verbose = FALSE
)

head(v1_pleio)
list.files(file.path(home, "pleiotropic"))
```

The grammar version is close but not identical in structure. The legacy call
draws its four dominance loci separately from the three additive loci, because
`same_add_dom_QTN` defaults to `FALSE`; the grammar version places dominance on
the additive loci with `same_as_add = TRUE`. Use `dominance(n_qtn = 4,
same_as_add = FALSE)` to mirror the legacy design exactly. The per-trait heritabilities become per-trait
proportions, and the per-marker variances are always available from
`qtn_table()` rather than through an export switch. In this legacy call,
which gives no `cor`, the genetic correlation among traits is a by-product of
the effect sizes. The legacy `cor` argument imposes it by Cholesky colouring,
which the grammar replaces with PleioArch through its own `cor` argument, as the
[multiple-trait chapter](multitrait.qmd) shows.

[]{#legacy-pleiotropic-v2}

```{r}
#| label: legacy-pleiotropic-v2
geno_hets <- filter_geno(geno, hets = "include", verbose = FALSE)

v2_pleio <- simulate_phenotype(geno_hets, architecture = "pleiotropy",
                               n_traits = 3, n_reps = 2, seed = 10) |>
  additive(prop = c(0.15, 0.3, 0.3), n_qtn = 3) |>
  dominance(prop = c(0.05, 0.1, 0.1), same_as_add = TRUE)

qtn_table(v2_pleio) |>
  select(trait, layer, snp, effect, var_explained) |>
  head()
```

The inbred maize panel has almost no heterozygotes. The grammar refuses a
dominance layer on loci without heterozygotes, so the markers are first
restricted with `filter_geno(hets = "include")` to those that have at least
one heterozygous individual; with `same_as_add = TRUE` the dominance layer
then uses the additive loci. The legacy function instead resamples its marker
set until the whole set contains a heterozygote, a weaker check described in
the theory section below.

### Partial pleiotropy and linkage

The legacy `architecture = "partially"` combines shared and trait-specific
markers in one call, with a user-supplied correlation matrix applied to the
genetic values. The grammar builds the same thing from two models combined by
`complex_phenotypes()`, which the [complex architectures chapter](complex.qmd)
demonstrates. The legacy `architecture = "LD"` corresponds to the grammar's
`"ld"` architecture, with renamed arguments.

[]{#legacy-ld}

```{r}
#| label: legacy-ld
v1_ld <- create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  ntraits = 2,
  architecture = "LD",
  type_of_ld = "indirect",
  ld_min = 0.2,
  ld_max = 0.8,
  model = "A",
  add_QTN_num = 3,
  add_effect = c(0.02, 0.05),
  h2 = c(0.2, 0.4),
  rep = 2,
  seed = 200,
  home_dir = home,
  output_dir = "ld",
  to_r = TRUE,
  quiet = TRUE,
  verbose = FALSE
)

read_tsv(file.path(home, "ld", "LD_Summary_Additive.txt"),
         show_col_types = FALSE) |>
  select(rep, SNP_causing_LD, QTN_for_trait_1, QTN_for_trait_2,
         LD_between_QTNs)
```

[]{#legacy-ld-v2}

```{r}
#| label: legacy-ld-v2
v2_ld <- simulate_phenotype(geno, architecture = "ld", n_traits = 2,
                            ld_type = "indirect", r2_min = 0.2, r2_max = 0.8,
                            n_reps = 2, seed = 200) |>
  additive(prop = c(0.2, 0.4), n_qtn = 3)

qtn_table(v2_ld) |>
  select(trait, snp, QTN_t1, QTN_t2, ld_r2)
```

The legacy summary reports a signed composite linkage-disequilibrium
coefficient between the two traits' markers; the grammar reports a squared
correlation. The scales differ, so the `ld_min` and `ld_max` bounds of one
interface are not the `r2_min` and `r2_max` bounds of the other.

### Argument mapping

The table lists each legacy argument and its grammar counterpart. It is copied
from the package specification, which is the maintained reference for this
mapping.

| Legacy `create_phenotypes()` | Grammar | Notes |
|---|---|---|
| `geno_obj` / `geno_file` / `geno_path` | `simulate_phenotype(geno=)` | Convert files once with `as_numeric()` |
| `ntraits` | `n_traits` | |
| `architecture = "pleiotropic"` | `architecture = "pleiotropy"` | |
| `architecture = "partially"` | `complex_phenotypes()` | Combine a pleiotropic and an independent model |
| `architecture = "LD"` | `architecture = "ld"` | |
| `add_QTN_num`, `dom_QTN_num`, `epi_QTN_num` | `additive(n_qtn=)`, `dominance(n_qtn=)`, `epistasis(n_pairs=)` | |
| `add_effect`, `dom_effect`, `epi_effect` | `effect=` on each layer | Geometric series by default |
| `big_add_QTN_effect` | none | Give an explicit `effect` series |
| `h2` | Sum of layer `prop`, or `complex_phenotypes(h2=)` | Emerges from the proportions |
| `same_add_dom_QTN` | `dominance(same_as_add=)` | |
| `degree_of_dom` | none | Per-component scaling absorbs it; a per-locus degree is available through `additive(orthogonal = TRUE, a=, d=)` |
| `epi_interaction` | `epistasis(interaction=)` | |
| `sim_method = "geometric"` / `"custom"` | `dist = "geometric"` / `effect=` | |
| `type_of_ld`, `ld_min`, `ld_max` | `ld_type`, `r2_min`, `r2_max` | Different linkage-disequilibrium scales |
| `cor` | `cor` | Legacy Cholesky colouring replaced by the PleioArch engine |
| `cor_res` | `resid_cor` | Grammar re-standardizes each trait after mixing |
| `rep`, `vary_QTN` | `n_reps`, `vary_qtn` | |
| `constraints = list(maf_above, maf_below, hets)` | `filter_geno()` upstream | Legacy filtered only the sampled markers |
| `seed` | `seed` | Same name, different arithmetic |
| `output_format`, `out_geno`, `home_dir`, `output_dir`, `to_r` | `write_phenotypes()`, `write_qtn_table()` | Object always returned |

## Assumptions and theory

### Heritability is an argument, not a sum

The legacy function builds a genetic value, measures its sample variance, and
derives the residual variance from the requested heritability. The implemented
expression, as recorded in the repository's equation map (entry
V1-core-linkage-7, "page unverified" against Falconer and Mackay), is

$$
\sigma_e^2 = \frac{V_G\,(1-h^2)}{h^2}, \qquad
V_G = \frac{1}{n-1}\sum_i (g_i-\bar g)^2 .
$$ {#eq-legacy-residual}

Here $h^2$ is the requested heritability, $V_G$ is the sample variance of the
realized genetic values $g_i$, and $n$ is the number of individuals. The
residuals are then drawn as normal values with that variance and added to the
centered genetic values, so that $y_{ij} = g_i + e_{ij} + \mu$
for replication $j$ (entry V1-core-linkage-8). Rearranging @eq-legacy-residual
shows the intent: $V_G/(V_G+\sigma_e^2) = h^2$ exactly for the variance
parameter. Two exceptions apply. With $h^2 = 0$ the function bypasses the
formula and returns unit-variance noise with no genetic value, and in a
multiple-trait run a zero for any trait does this for every trait. For several
traits, the residuals are drawn jointly and correlated through `cor_res`. The realized sample ratio is not fixed, because the residual sample
variance and the genetic-residual covariance fluctuate. This is the same
random-residual behaviour that the grammar offers through
`residual_mode = "random"`, but the two reach it by different routes: the
legacy function scales the residual to the genetic sample variance, while the
grammar scales the genetic component to the phenotypic budget.

Both consequences follow from the definitions. First, the legacy effect sizes
matter only through $V_G$; doubling every effect doubles the residual standard
deviation and leaves the variance ratio unchanged. Second, the per-trait
heritabilities of a multi-trait legacy run are applied after the optional
colouring transformation that imposes `cor`, so the genetic correlation and
the heritabilities are set in sequence rather than jointly.

### The legacy genetic value

The additive value sums effect times coded genotype over the selected markers;
the dominance value sums effects over heterozygous markers; the epistatic
value multiplies the coded genotypes of the interacting markers (entries
V1-core-linkage-1 to 3, "page unverified"). The total is centered so that the
phenotype mean equals the `mean` argument. The geometric series assigns
$a_k = a^k$ to the $k$-th marker, with `big_add_QTN_effect` replacing the first
term (entry V1-core-linkage-6). With `same_add_dom_QTN = TRUE`, the dominance
effect of each shared marker is `degree_of_dom` times its additive effect
(entry V1-core-linkage-5). The legacy samplers check heterozygosity for the
whole marker set, not per marker: they redraw, up to ten times, while no
selected marker has a heterozygote (entry V1-pleio-effects-10), so a set can
pass with a single heterozygous marker and the others contribute no dominance
variance.

The grammar uses the same per-marker scores as raw components, but then
centers and scales each layer to its proportion. The legacy
`var_explained` quantities and the grammar's are therefore not the same
numbers even for identical markers and effects.

### Partial pleiotropy and vary_QTN

Under `vary_QTN = TRUE`, each replication samples its own markers and the
residual variance is recomputed per replication from that replication's
genetic variance. The grammar's `vary_qtn` has the same meaning at the level of
the design; its seeds and draws differ.

### What is verified about the legacy engine

The legacy engine's behaviour is pinned by reference outputs captured from
version 1.3.0 and stored with the package. The test file `test-v130-parity.R`
compares the current legacy results against those stored objects, so a change
in legacy numbers is detected as a test failure. This is a regression contract,
not a statement that the legacy theory is the reference for the grammar.
The equation map lists a number of legacy findings, such as the uncentered
epistatic product and the per-marker variances that ignore linkage
covariance, which are retained because the function is frozen.

## Implementation and practical limits

The legacy engine lives in the `legacy_*.R` files and shares no simulation
code with the grammar. Its random draws run under `RNGversion("3.5.1")`, and
the function sets and restores the working directory around a run. The source
locations in the [developer index](developer-index.qmd) point to the current
definitions.

- Pass an absolute `home_dir`. The function resolves a relative path after
  changing into it, which nests the run folder one level deeper than intended.
- Each run creates a new folder; an existing `output_dir` name gets a numbered
  suffix rather than being overwritten.
- `to_r = TRUE` returns a wide table for one trait, with columns `taxa`,
  `rep1`, `rep2`, and so on, and a long-by-replication table for several
  traits, whose columns encode the heritability (`Trait_k_H2_x`).
- A single-trait run rejects `QTN_list` and the `"ADE"` and `"DE"` models with
  an explicit error.
- The legacy function is bugfix-only. New features, including
  `residual_mode`, controlled genetic correlation for any number of traits,
  orthogonal dominance coding, and the transcriptome layers, exist only in the
  grammar.
- To migrate a script, translate the design with the mapping table, then
  compare variance partitions and marker counts rather than individual values.
