# Marker-assisted breeding {#sec-markers}

A donor line carries an allele you want, at a known marker, inside a genome
you do not want. Marker-assisted backcrossing moves that allele into an elite
recurrent parent while recovering the recurrent genome faster than repeated
backcrossing alone would. A related task stacks several target alleles into
one line. This chapter runs both on two maize inbreds, B73 and Mo17, and
checks the recovery of the recurrent genome against its Mendelian expectation.

[]{#markers-setup}

```{r}
#| label: markers-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-markers-example}

### Founders and informative markers

[]{#markers-library}

```{r}
#| label: markers-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

The two parents come from the bundled maize panel. B73 is the recurrent
parent and Mo17 the donor. A marker is **informative** for backcrossing only
when the two parents are homozygous for different alleles, because only then
does a progeny genotype reveal which parent each allele came from.

[]{#markers-founders}

```{r}
#| label: markers-founders
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
founders <- as_population(geno, individuals = c("B73", "MO17"))
recurrent <- founders["B73"]
donor <- founders["MO17"]

d <- dosages(founders)
informative <- d[, "B73"] * d[, "MO17"] == -1
tibble(markers = nrow(d), informative = sum(informative))
```

`dosages()` returns the markers-by-individuals matrix of `-1`, `0`, `1`
codes. The product of the two parental dosages is `-1` exactly when they are
opposite homozygotes.

### Choose a target and its flanking markers

The target is an informative marker on chromosome 1 near 60 cM of the
bundled synthetic map. Two informative markers about five centimorgans to
either side serve as flanking markers for recombinant selection.

[]{#markers-target}

```{r}
#| label: markers-target
map <- founders$map
chr1 <- which(informative & map$chr == "1")
target_row <- chr1[which.min(abs(map$cm[chr1] - 60))]
target <- map$snp[target_row]

left <- chr1[chr1 < target_row & map$cm[chr1] <= map$cm[target_row] - 5]
right <- chr1[chr1 > target_row & map$cm[chr1] >= map$cm[target_row] + 5]
flanking <- map$snp[c(max(left), min(right))]

map |>
  filter(snp %in% c(target, flanking)) |>
  select(snp, chr, pos, cm)
```

The donor allele at the target is the one coded by Mo17's dosage there.

### Backcross with and without background selection

We make the F1, then three backcross generations of 100 progeny each. In the
first scheme, `mabc_select()` keeps the single best candidate each
generation: it must carry the donor allele at the target (foreground
selection), is preferred when both flanking markers are already homozygous
for B73 (recombinant selection), and among those is ranked by its recovery of
the B73 genome at the remaining informative markers (background selection).
The target region of 10 cM on either side is excluded from the background
score, because that is the donor segment we intend to keep.

[]{#markers-mabc}

```{r}
#| label: markers-mabc
f1 <- cross(recurrent, donor, n = 1, seed = 1)
region <- list(chr = "1", from = map$cm[target_row] - 10, to = map$cm[target_row] + 10)

parent <- f1
mabc <- list()
for (t in 1:3) {
  parent_recovery <- recurrent_parent_recovery(parent, recurrent, donor)
  cohort <- cross(parent, recurrent, n = 100, seed = 10 + t)
  cohort_recovery <- recurrent_parent_recovery(cohort, recurrent, donor)
  parent <- mabc_select(
    cohort, recurrent, donor,
    target_markers = target, flanking_markers = flanking,
    exclude_interval = region, n = 1, seed = 20 + t
  )
  mabc[[t]] <- tibble(
    generation = paste0("BC", t),
    unselected = 1 - 0.5^(t + 1),
    expected_cohort = 0.5 + 0.5 * parent_recovery,
    cohort_mean = mean(cohort_recovery),
    cohort_best = max(cohort_recovery),
    selected = recurrent_parent_recovery(parent, recurrent, donor)
  )
}
bc3 <- parent
bind_rows(mabc)
```

`recurrent_parent_recovery()` scores each individual by its fraction of
B73 alleles over the informative markers. A backcross progeny receives one
complete recurrent gamete and half of its parent's genome, so a cohort's
expected recovery is one half plus half of its parent's recovery
(`expected_cohort`). The `unselected` column is the Mendelian value
$1 - 2^{-(t+1)}$ for a pedigree with no selection; it applies to BC1 only here,
because from BC2 on each cohort descends from a selected plant. The simulated
cohort means track `expected_cohort`. The selected individual exceeds its cohort because it was chosen for
exactly that score. The last selected candidate's record shows how the three
stages were applied:

[]{#markers-mabc-record}

```{r}
#| label: markers-mabc-record
record <- attr(bc3, "mabc")
record |>
  filter(selected) |>
  select(id, target, feasible, flank_recurrent, background_recovery, rank)

record |>
  count(target, feasible)
```

The `target` column codes each candidate's genotype at the target as `R`,
`H` or `D` for recurrent homozygote, heterozygote or donor homozygote. In a
backcross about half the progeny are heterozygous at the target; only those
are feasible.

In the second scheme we keep a carrier of the donor allele but choose it at
random, which is foreground selection alone. `marker_select()` does this with
the favourable allele given explicitly; with no `rank_on` score, feasible
candidates are ranked by the number of targets met, which is one for all of
them, and the tie is broken at random.

[]{#markers-foreground}

```{r}
#| label: markers-foreground
donor_allele <- dosages(donor)[target, 1]

carrier <- f1
foreground <- numeric(3)
for (t in 1:3) {
  cohort <- cross(carrier, recurrent, n = 100, seed = 10 + t)
  carrier <- marker_select(
    cohort, markers = target, favorable = donor_allele,
    requirement = "carrier", n = 1, seed = 30 + t
  )
  foreground[t] <- recurrent_parent_recovery(carrier, recurrent, donor)
}

bind_rows(mabc) |>
  select(generation, unselected, with_background = selected) |>
  mutate(foreground_only = foreground)
```

Only the BC1 cohort is shared. From BC2 on the cohorts differ because they
descend from different selected plants, so the table compares the two rules
compounded over three generations.

### Linkage drag around the target

Recovery over the whole genome hides what happens next to the target. The
donor segment carried with the target shrinks only through recombination
nearby, so recovery on chromosome 1 lags behind the other chromosomes. We
score the selected BC3 plant from the background-selected scheme separately
on chromosome 1 and on the rest of the genome.

[]{#markers-drag}

```{r}
#| label: markers-drag
best_bc3 <- bind_rows(mabc)$selected[3]
chr1_markers <- map$snp[informative & map$chr == "1"]
other_markers <- map$snp[informative & map$chr != "1"]
tibble(
  region = c("Genome", "Chromosome 1", "Other chromosomes"),
  recovery = c(
    best_bc3,
    recurrent_parent_recovery(bc3, recurrent, donor, markers = chr1_markers),
    recurrent_parent_recovery(bc3, recurrent, donor, markers = other_markers)
  )
)
```

The `weights = "interval"` option scores each marker by the length of genome
it represents instead of equally, which matters when markers cluster.

[]{#markers-weights}

```{r}
#| label: markers-weights
tibble(
  weights = c("equal", "interval"),
  recovery = c(
    recurrent_parent_recovery(bc3, recurrent, donor),
    recurrent_parent_recovery(bc3, recurrent, donor, weights = "interval")
  )
)
```

### Gene pyramiding in an F2

Pyramiding stacks several target alleles in one individual. We select three
informative markers on different chromosomes and ask which F2 plants of the
B73 x Mo17 cross are homozygous for the Mo17 allele at all three. Under
independent assortment, one plant in 64 should qualify.

[]{#markers-pyramid}

```{r}
#| label: markers-pyramid
targets <- sapply(c("2", "5", "8"), function(chr) {
  k <- which(informative & map$chr == chr)
  map$snp[k[which.min(abs(map$cm[k] - 50))]]
})
favourable <- dosages(donor)[targets, 1]

f2 <- selfcross(f1, n = 400, seed = 3)
pyramid <- marker_select(
  f2, markers = targets, favorable = favourable, requirement = "homozygote"
)
status <- attr(pyramid, "marker_select")
head(status)

tibble(
  requirement = c("Homozygous at all three", "Carrier at all three"),
  expected = c(400 / 64, 400 * (3 / 4)^3),
  simulated = c(
    sum(status$feasible),
    sum(attr(marker_select(
      f2, markers = targets, favorable = favourable, requirement = "carrier"
    ), "marker_select")$feasible)
  )
)
```

The `state` column lists the genotype at each target as `F`, `H` or `U`
(favourable homozygote, heterozygote, unfavourable homozygote). A staged
pyramid relaxes the requirement to some of the targets and ranks candidates
by the number met, which lets a program select now and finish the stack
later. Candidates that already meet all three targets rank first, so with
enough of them the staged selection still returns complete stacks.

[]{#markers-staged}

```{r}
#| label: markers-staged
staged <- marker_select(
  f2, markers = targets, favorable = favourable,
  requirement = "homozygote", min_markers = 2, n = 5, seed = 4
)
attr(staged, "marker_select_info")$staged
attr(staged, "marker_select") |>
  filter(selected) |>
  select(id, state, n_met, rank)
```

### Ranking feasible candidates on a marker index

`rank_on` orders the feasible candidates by any score. With the simulation's
own causal loci and their additive effects, `additive_value()` is a marker
index of the kind used in marker-assisted recurrent selection, here with the
true effects rather than estimated ones. We define a trait on the F2 and keep
the five carriers with the highest index.

[]{#markers-index}

```{r}
#| label: markers-index
sim <- simulate_phenotype(f2, h2 = 0.5, seed = 5) |>
  additive(n_qtn = 30)
template <- template_effects(sim)
index <- additive_value(f2, template$qtn, template$a)

mars <- marker_select(
  f2, markers = targets, favorable = favourable, requirement = "carrier",
  rank_on = index, n = 5, seed = 6
)
attr(mars, "marker_select") |>
  filter(selected) |>
  select(id, state, score, rank)
```

## Assumptions and theory

### Informative markers and the recovery score

For the recurrent parent homozygous with dosage $r_m \in \{-1, +1\}$ at an
informative marker $m$ and a candidate with dosage $x_{mi}$, the package
defines the recurrent-allele score

$$
s_{mi} = \frac{x_{mi}\, r_m + 1}{2},
$$ {#eq-recurrent-score}

which is 1, 1/2 or 0 for a recurrent homozygote, a heterozygote or a donor
homozygote (`.mabc_founders()`, a package definition). Recovery is the
weighted mean of these scores over the background set $B$,

$$
R_i = \frac{\sum_{m \in B} w_m s_{mi}}{\sum_{m \in B} w_m},
$$ {#eq-recovery}

with equal weights, interval weights or user weights (`.mabc_recovery()`,
package definition). The interval weight of a marker is the centimorgan
length of genome nearer to it than to any other scored marker on its
chromosome, minus any part inside an excluded interval (`.mabc_weights()`).
This is recovery of recurrent-parent **marker alleles**, not identity by
descent: a `Population` carries no per-locus ancestry, so a marker allele
shared by both parents is simply uninformative and is dropped.

### Expected recovery without selection

Each backcross joins one gamete from the recurrent parent, which is entirely
recurrent, to one gamete from the current backcross plant, which carries on
average that plant's recurrent fraction. For an unselected cohort after $t$
backcrosses following the F1,

$$
R_0 = \tfrac12, \qquad R_t = \tfrac12 + \tfrac12 R_{t-1}, \qquad
R_t = 1 - 2^{-(t+1)}.
$$ {#eq-expected-recovery}

The equation map records this as Mendelian, attributed to Frisch & Melchinger
2005 *Genetics* 170:909–917 (paper verified; location unverified). The
expectation holds for the cohort mean; the selected individual is drawn from
the upper tail of that cohort.

### Staged selection in `mabc_select()`

Feasibility requires the donor allele at every target: at least one donor
allele for `"donor_carrier"`, both for `"donor_homozygote"`. Among feasible
candidates the ranking is lexicographic: the number of flanking markers in
the recurrent-homozygous state first, then the background recovery, then a
seeded random tie-break. The equation map attributes the three-stage
procedure to Frisch & Melchinger 2001 *Crop Sci* 41:1485–1494 (verified;
page unverified); the feasibility rule and the recovery estimator are package
definitions.

### Feasibility and ranking in `marker_select()`

With $f_m \in \{-1, +1\}$ the favourable homozygote at target $m$, the
package computes $z_{mi} = x_{mi} f_m$ and declares the requirement met when
$z_{mi} \ge 0$ (carrier) or $z_{mi} \ge 1$ (homozygote). A candidate is
feasible when it meets the requirement at `min_markers` or more targets.
The equation map records this as DECISION-029, a package definition. A
`rank_on` index of the form $I_i = \sum_j w_j x_{ij}$ is the marker index of
Lande & Thompson 1990 *Genetics* 124:743–756 (verified; page unverified);
with the simulation's own effects as weights it is an oracle index, an upper
bound on what estimated effects could achieve.

## Implementation and practical limits

Marker selection is deterministic R code apart from the seeded tie-break;
the meioses that produce each cohort are simulated by the crossing engine,
with the random draws made in R.

- Every target and flanking marker must be informative; the function errors
  rather than guess which allele is the donor's.
- The founders, the candidates and the donor must share one allele coding.
  That holds when the candidates are bred from the founders, as here. For
  separately converted panels, convert them in one `as_numeric()` call or
  use `method = "reference"` with the same `ref_allele`, because the coding
  cannot be checked from dosages.
- Both weightings drop an excluded interval from the denominator. With equal
  weights the denominator is a marker count, so a densely genotyped region
  counts more; `marker_weights = "interval"` weights each marker by the map
  length it represents.
- A one-sided set of flanking markers is accepted with a warning; without any
  flanking markers no recombinant selection takes place.
- `marker_select()` is an error when fewer candidates are feasible than `n`
  requests. Check the `feasible` column of its attribute first when the stack
  is large or the population small.
- Phenotypic truncation with [`select_ind()`](selection.qmd) is not
  marker-assisted selection; the two can be combined through `rank_on`.
