# Mating designs and crossbreeding {#sec-mating}

Suppose you have two groups of inbred lines, say two heterotic pools, and want
to plan the crosses among them, run them all in one call, trace each progeny's
expected ancestry, and ask how much of the hybrid advantage survives a
backcross or a rotation. This chapter writes mating plans with
`mating_design()`, executes them with `mate()`, follows ancestry with
`breed_composition()` and `crossbreed()`, and checks the heterosis produced by
a dominance architecture against its expected value.

[]{#mating-setup}

```{r}
#| label: mating-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-mating-example}

### Two founder pools from the maize panel

[]{#mating-library}

```{r}
#| label: mating-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

The maize panel has no breed labels, so we form two pools of eight inbred
lines each by taking the first sixteen lines of the table. The `pool`
argument of `as_population()` records the label in the pedigree; it is the
only thing that distinguishes the pools, and it is what `breed_composition()`
and `heterosis()` trace. The pools are therefore an arbitrary split of one
panel, not two diverged populations, and the heterosis they show is whatever
their allele-frequency differences happen to produce.

[]{#mating-pools}

```{r}
#| label: mating-pools
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
ids <- colnames(geno)[-(1:5)]

pool_a <- as_population(geno, individuals = ids[1:8], pool = "A")
pool_b <- as_population(geno, individuals = ids[9:16], pool = "B")
pool_a
```

### Write a mating plan

`mating_design()` returns a data frame with `mother`, `father`, and `n`
columns. The standard designs differ only in which pairs they list.

[]{#mating-design}

```{r}
#| label: mating-design
mating_design(pool_a[1:3], pool_b[1:2], design = "factorial",
              progeny_per_cross = 4)

designs <- tibble(
  design = c("factorial 3 x 2", "half_diallel 4", "diallel 4",
             "diallel 4 with selfs", "nested 2 fathers x 2 mothers", "random 5"),
  rows = c(
    nrow(mating_design(pool_a[1:3], pool_b[1:2], design = "factorial")),
    nrow(mating_design(pool_a[1:4], design = "half_diallel")),
    nrow(mating_design(pool_a[1:4], design = "diallel")),
    nrow(mating_design(pool_a[1:4], design = "diallel", allow_self = TRUE)),
    nrow(mating_design(pool_a[1:4], pool_b[1:2], design = "nested",
                       mothers_per_father = 2)),
    nrow(mating_design(pool_a, pool_b, design = "random", n_crosses = 5,
                       seed = 1))
  ),
  parameter = c(3 * 2, choose(4, 2), 4 * 3, 4 * 3 + 4, 2 * 2, 5)
)
designs
```

The `parameter` column is the number of pairs each design defines: every
mother with every father for the factorial (North Carolina design II), every
unordered pair once for the half diallel, every ordered pair for the full
diallel, plus one self per parent when selfs are allowed, each father with its
own set of mothers for the nested design (North Carolina design I), and the
requested number of random pairs. `?mating_design` names the Griffing (1956)
methods that the diallel variants correspond to.

### Execute the plan across pools

When parents come from several populations, the plan needs `mother_pool` and
`father_pool` columns naming the arguments of `mate()`. All rows run in one
pass, and the returned `Population` carries the plan with each row's progeny
identifiers.

[]{#mating-mate}

```{r}
#| label: mating-mate
plan <- mating_design(pool_a[1:3], pool_b[1:2], design = "factorial",
                      progeny_per_cross = 4) |>
  mutate(mother_pool = "A", father_pool = "B")

hybrids <- mate(plan, A = pool_a, B = pool_b, seed = 1, prefix = "H")
hybrids

attr(hybrids, "plan") |>
  select(mother, father, n, progeny) |>
  head(3)

table(families(hybrids, "full_sib"))
```

Each row produced a full-sib family of four, and the family labels name the
two parents. Progeny identifiers follow the `prefix`.

### Expected ancestry

`breed_composition()` gives each individual's expected fraction of genes from
each founder pool, traced through the pedigree: a founder is one for its own
pool, and every progeny is the mean of its parents. We build a backcross
series by hand with `cross()` to see the familiar halving of the donor
fraction.

[]{#mating-composition}

```{r}
#| label: mating-composition
f1 <- hybrids[1]
bc1 <- cross(f1, pool_a[1], n = 1, seed = 2)
bc2 <- cross(bc1, pool_a[1], n = 1, seed = 3)
bc3 <- cross(bc2, pool_a[1], n = 1, seed = 4)

series <- c(f1, bc1, bc2, bc3)
breed_composition(series) |>
  as_tibble(rownames = "id") |>
  mutate(generation = c("F1", "BC1", "BC2", "BC3"),
         parameter_A = c(1 / 2, 3 / 4, 7 / 8, 15 / 16)) |>
  select(generation, id, A, B, parameter_A)
```

These are **expected** fractions. The realized share of an individual's
genome that came from pool A varies around them, and the `Population` does not
record per-locus founder origin, so the realized share is not available from
the pedigree. The [marker-assisted breeding chapter](markers.qmd) estimates
it from markers instead.

### Crossbreeding systems

`crossbreed()` runs the classical systems on named pure-breed populations:
each generation mates random pairs between the pools the system prescribes,
and the attribute `history` records the mean expected composition per
generation. Sex is not modelled; "dam" and "sire" are the roles of the two
pools in the plan.

[]{#mating-crossbreed}

```{r}
#| label: mating-crossbreed
breeds <- list(A = pool_a, B = pool_b)

two_way <- crossbreed(breeds, "two_way", n_progeny = 60, seed = 5)
backcross <- crossbreed(breeds, "backcross", n_progeny = 60, seed = 6)
rotation <- crossbreed(breeds, "rotational", n_progeny = 60,
                       generations = 5, seed = 7)

attr(backcross, "history")
attr(rotation, "history")
```

A two-breed rotation alternates the sire breed, and its composition converges
toward two thirds of the current sire breed and one third of the other. The
three-way and terminal systems add a third pool.

[]{#mating-three-way}

```{r}
#| label: mating-three-way
pool_c <- as_population(geno, individuals = ids[17:24], pool = "C")
three <- list(A = pool_a, B = pool_b, C = pool_c)

attr(crossbreed(three, "three_way", n_progeny = 60, seed = 8), "history")
attr(crossbreed(three, "terminal", n_progeny = 60, sire_breed = "C",
                seed = 9), "history")
```

Both systems reach the same composition, one quarter from each of the first
two pools and one half from the third. The difference is in use: a terminal
cross is not bred further.

### Heterosis on a dominance architecture

`heterosis()` scores a crossbred population on a frozen additive-plus-dominance
architecture ([trait values](trait.qmd)) and reports its mean minus the
composition-weighted mean of the pure pools. It also returns the **expected**
F1 heterosis of each pair of pools, computed from the pools' own gamete
frequencies. We take 30 marker loci at random as causal, give each the same
additive effect and the same dominance deviation, and compare.

[]{#mating-heterosis}

```{r}
#| label: mating-heterosis
set.seed(1)
causal <- sample(nrow(geno), 30)
a_eff <- rep(0.5, 30)
d_dev <- rep(0.3, 30)

h_f1 <- heterosis(two_way, breeds, qtn = causal, a = a_eff, d = d_dev)
h_bc <- heterosis(backcross, breeds, qtn = causal, a = a_eff, d = d_dev)
h_additive <- heterosis(two_way, breeds, qtn = causal, a = a_eff, d = 0)

p_a <- rowMeans(dosages(pool_a)[causal, ] == 1)
p_b <- rowMeans(dosages(pool_b)[causal, ] == 1)
expected_inbred <- sum(d_dev * (p_a + p_b - 2 * p_a * p_b))

tibble(
  quantity = c("Expected F1 heterosis (package)",
               "Expected F1 heterosis (inbred-line formula)",
               "Realized F1 heterosis", "Realized backcross heterosis",
               "Realized F1 heterosis, d = 0"),
  value = c(h_f1$expected_f1["A", "B"], expected_inbred,
            h_f1$realized, h_bc$realized, h_additive$realized)
)
h_f1$breed_means
```

The package's expected value and the inbred-line formula agree, because every
line in both pools is homozygous at the causal loci. The realized F1 value
scatters around the expectation, since 60 progeny of random pairs sample the
pool allele frequencies. With `d = 0` the realized heterosis is close to
zero: the additive part cancels against the mid-parent except for that
sampling of parents.

The backcross retains more than the classical one half of the F1 heterosis.
That is not an error. The one-half rule assumes the recurrent breed is in
Hardy-Weinberg proportions at the locus, and a pool of inbred lines that
differ at a locus has no heterozygotes at all. The theory section gives the
exact condition.

## Assumptions and theory

### Expected composition

Let $F_{i,b}$ be the expected fraction of individual $i$'s genes from pool
$b$. A founder has $F_{i,b} = 1$ for its own pool and $0$ otherwise, and a
progeny of mother $m$ and father $f$ has

$$
F_{i,b} = \tfrac{1}{2}\left(F_{m,b} + F_{f,b}\right),
$$ {#eq-composition}

with a self or doubled haploid equal to its parent (package definition,
`breed_composition()`). This is the expectation of the share of alleles
inherited from each pool and says nothing about which loci came from where.

### Realized and expected heterosis

The realized heterosis of a crossbred population is its mean total genotypic
value minus the pool means weighted by its mean expected composition,

$$
H = \bar G_{\text{cross}} - \sum_b \bar F_b\,\bar G_b,
$$ {#eq-realized-heterosis}

the mid-parent value for an F1 (package definition, `heterosis()`; the equation
map cites Falconer and Mackay 1996, chapter 14, page unverified, for mid-parent
heterosis). For the expected F1 heterosis between pools A and B, write
$p_A$ and $p_B$ for the gamete frequencies of the allele coded $+a$ at a locus
and $h_A$, $h_B$ for the pools' own heterozygote frequencies. The F1's allele
frequency is the average of the two, so the additive part cancels against the
mid-parent and what remains is

$$
H_{F1} = \sum_j d_j\left[h_{AB,j} - \tfrac{1}{2}\left(h_{A,j} + h_{B,j}\right)\right],
\qquad h_{AB} = p_A(1 - p_B) + (1 - p_A)\,p_B .
$$ {#eq-expected-heterosis}

The package documents this as its own derivation from gamete frequencies,
which is why linkage does not enter. Two special cases follow algebraically.
When each pool is in Hardy-Weinberg proportions, $h = 2p(1-p)$ and the
expression reduces to $\sum_j d_j (p_{Aj} - p_{Bj})^2$, the textbook
$H_{F1} = \sum d\,y^2$ with $y$ the allele-frequency difference (Falconer and
Mackay 1996, page unverified, as cited in `?heterosis`). When each pool
consists of fully inbred lines, $h = 0$ and it reduces to
$\sum_j d_j\,(p_{Aj} + p_{Bj} - 2p_{Aj}p_{Bj})$, the formula evaluated in the
worked example. A purely additive architecture gives zero in every case.

### How much heterosis later generations retain

The classical fractions, one half in an F2 or backcross and two thirds in a
two-breed rotation at equilibrium, come from the probability that an
individual's two alleles at a locus originate from different pools under a
dominance model. The package's documentation states the exact conditions:
at a locus with non-zero F1 heterosis, a backcross to pool A retains one half
if and only if $h_A = 2p_A(1 - p_A)$, that is, only the recurrent pool needs to
be in Hardy-Weinberg proportions; an F2 retains one half if and only if the
two pools' deviations $h - 2p(1-p)$ sum to zero. The condition for the
two-thirds rotation figure is not derived in the package and is given as the
classical value for pools in Hardy-Weinberg proportions. Our inbred-line
pools violate the backcross condition at every segregating locus, with $h_A =
0 < 2p_A(1-p_A)$, which is why the backcross retained more than one half.
`heterosis()` computes the realized value from the actual genotypes and does
not assume any retention fraction.

### Mating designs

The designs are bookkeeping, not genetics: they enumerate pairs. Individuals
are compared by pedigree identity rather than display name, so two pools that
reuse names stay distinct provided they were imported under their own pool
labels, and a parent that appears twice under different names enters a design
once. The random design draws a mother uniformly and then one of her
admissible fathers uniformly, which is uniform over pairs only when every
mother has the same number of admissible fathers.

## Implementation and practical limits

- Pool labels are the only record of origin. Import each breed or heterotic
  group with its own `pool`, or two founders with the same name and genotype
  are the same individual, and mating them is a self.
- All populations in one plan must share one marker map and the same allele
  coded `1`; see the [crossing chapter](cross.qmd) on allele orientation.
- `crossbreed()` mates random dam-sire pairs with one progeny each; the
  number of progeny per generation is `n_progeny`, and no sexes are modelled.
- `breed_composition()` and the `history` attribute are expectations from the
  pedigree. Realized genome shares require marker-based estimation, as in
  `recurrent_parent_recovery()`.
- `heterosis()` requires every founder that the crossbred pedigree traces to
  be present in the supplied pools, and scores the frozen architecture per se;
  epistasis is outside the function.
- The pools here are an arbitrary split of one panel. For a study of
  heterotic groups, import populations with documented provenance and let the
  observed allele-frequency differences, not a label, carry the biology.
