# Expression-mediated phenotypes {#sec-mediation}

Suppose part of a trait's variation acts through gene expression: a few genes
whose expression levels shift the phenotype, in addition to markers that act
directly. If the expression itself is partly heritable, some of that
expression-mediated variation traces back to the genome and some does not. This
chapter builds such a phenotype on the maize genotypes, reads the causal genes
from the usual tables, and checks how the package splits the mediated component
into its genetic and environmental parts.

[]{#mediation-setup}

```{r}
#| label: mediation-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-mediation-example}

### Load the package, the genotypes, and a transcriptome

[]{#mediation-library}

```{r}
#| label: mediation-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)

data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
tx <- simulate_transcriptome(geno, n_genes = 400, seed = 11)
tx
```

The transcriptome is the one built in the
[previous chapter](expression.qmd): 400 genes on 280 maize lines, each gene
with its own expression heritability. The genetic part of every gene's
expression is stored in the object, which is what makes the split below possible.

### Build a phenotype with an expression layer

Pass the transcriptome to `simulate_phenotype()`, then add a `transcriptome()`
layer that selects causal genes and a direct `additive()` layer on markers.
The expression layer's `prop` is a separate variance category: it is not drawn
from the marker heritability budget, and all layers together may not exceed one.

[]{#mediation-model}

```{r}
#| label: mediation-model
sim <- simulate_phenotype(geno, seed = 2, transcriptome = tx) |>
  transcriptome(prop = 0.3, n_genes = 20) |>
  additive(prop = 0.2, n_qtn = 10)
sim
```

The print output shows three budget rows: the expression component, the
additive marker component, and the residual. The requested genetic share names
the marker layer only. The realized broad-sense ratio is larger than that share,
because the genetic part of the expression component also counts as genetic.
The last lines preview the split explained next.

### Read the causal genes

`qtn_table()` lists the causal genes as rows of the `transcriptome` layer. The
`snp` column holds gene identifiers, the `effect` column holds the slopes, and
`var_explained` is each gene's marginal share of phenotypic variance.

[]{#mediation-qtn}

```{r}
#| label: mediation-qtn
qt <- qtn_table(sim)
table(qt$layer)

qt |>
  filter(layer == "transcriptome") |>
  select(trait, layer, snp, effect, var_explained) |>
  head()
```

The slopes are drawn from a standard normal distribution when `slopes` is not
given. Their absolute scale is irrelevant, because the component is scaled to
`prop`; only their relative sizes matter. The marginal shares need not add to
0.3, because causal genes co-express and their contributions covary.

### Split the mediated component

`mediation_split()` returns the realized shares of phenotypic variance carried
by the genetically mediated part, the environmentally mediated part, and twice
their covariance.

[]{#mediation-split}

```{r}
#| label: mediation-split
mediation_split(sim)
```

The three columns sum to the realized share of the expression component. We
now reproduce them from public outputs. The component scores each causal gene's
expression standardized by its own standard deviation, weights it by the slope
divided by the largest absolute slope, sums, centers, and scales the sum to
sample variance 0.3. The genetic part applies the same weights and the same
denominators to the genetic expression of the same genes.

[]{#mediation-reconstruct}

```{r}
#| label: mediation-reconstruct
causal <- qt |> filter(layer == "transcriptome")
expression <- tx$expression[causal$snp, ]
genetic_expression <- tx$genetic_expression[causal$snp, ]

gene_sd <- apply(expression, 1, sd)
weights <- causal$effect / max(abs(causal$effect))

z_total <- (expression - rowMeans(expression)) / gene_sd
z_genetic <- (genetic_expression - rowMeans(genetic_expression)) / gene_sd

raw_total <- as.numeric(weights %*% z_total)
raw_genetic <- as.numeric(weights %*% z_genetic)
scale <- sqrt(0.3) / sd(raw_total)

mediated <- scale * (raw_total - mean(raw_total))
mediated_genetic <- scale * (raw_genetic - mean(raw_genetic))
mediated_env <- mediated - mediated_genetic

wide <- phenotypes_wide(sim)
individuals <- colnames(geno)[-(1:5)]
y <- wide$Trait_1[match(individuals, wide$id)]

tibble(
  quantity = c("Expression component", "Genetic-mediated", "Environment-mediated",
               "Covariance term"),
  package = c(sum(mediation_split(sim)[, 2:4]), mediation_split(sim)$genetic_mediated,
              mediation_split(sim)$env_mediated, mediation_split(sim)$covariance),
  reconstructed = c(var(mediated), var(mediated_genetic), var(mediated_env),
                    2 * cov(mediated_genetic, mediated_env)) / var(y)
)
```

The reconstruction matches the package. The expression component's sample
variance is exactly 0.3 by construction, but its share of phenotypic variance
differs slightly from 0.3 because the phenotypic variance includes the chance
covariances among the components, as the [first chapter](first-simulation.qmd)
explained for a single additive layer.

The genetic values returned by `genetic_values()` are the marker component plus
the genetic-mediated part. Subtracting the reconstructed genetic-mediated part
leaves the additive marker component, whose sample variance is the requested 0.2.

[]{#mediation-genetic-values}

```{r}
#| label: mediation-genetic-values
g <- genetic_values(sim)[, "Trait_1"]
g <- g[match(individuals, names(g))]

tibble(
  quantity = c("Additive marker component", "Genetic value", "Genetic / phenotypic variance"),
  parameter = c(0.2, NA, NA),
  simulated = c(var(g - mediated_genetic), var(g), var(g) / var(y))
)
```

### Choose genes by their heritability

How much of the expression component is genetic depends on which genes are
causal. The chunk builds two phenotypes with the same slope on ten genes each:
the ten genes with the lowest realized expression heritability, and the ten
with the highest.

[]{#mediation-genes}

```{r}
#| label: mediation-genes
ranked <- tx$genes |> arrange(h2_realized)
low_genes <- head(ranked$gene_id, 10)
high_genes <- tail(ranked$gene_id, 10)

split_for <- function(genes) {
  simulate_phenotype(geno, seed = 5, transcriptome = tx) |>
    transcriptome(prop = 0.4, genes = genes, slopes = rep(1, 10)) |>
    mediation_split()
}

bind_rows(
  split_for(low_genes) |> mutate(genes = "Lowest expression h2"),
  split_for(high_genes) |> mutate(genes = "Highest expression h2")
) |>
  mutate(mean_gene_h2 = c(mean(head(ranked$h2_realized, 10)),
                          mean(tail(ranked$h2_realized, 10)))) |>
  select(genes, mean_gene_h2, genetic_mediated, env_mediated, covariance)
```

With weakly heritable genes almost all of the expression component is
environmental; with strongly heritable genes most of it is genetic. The
component's sample variance is 0.4 in both cases; its share of phenotypic
variance differs slightly only through its chance covariance with the residual.
The two phenotypes therefore have very different broad-sense ratios despite
identical `prop`.

### Use a real expression matrix instead

When the expression source is an observed matrix passed through
`expression =`, the package has no basis for asserting which part of it is
genetic. The whole component is then treated as an environmental predictor:
it is excluded from the genetic values and from the realized ratio, and
`mediation_split()` returns `NULL`. The chunk uses the simulated matrix as if it
were observed data.

[]{#mediation-real}

```{r}
#| label: mediation-real
sim_real <- simulate_phenotype(geno, seed = 2, expression = tx$expression) |>
  transcriptome(prop = 0.3, n_genes = 20) |>
  additive(prop = 0.2, n_qtn = 10)
sim_real
is.null(mediation_split(sim_real))

tibble(
  quantity = "Genetic / phenotypic variance",
  derived_source = var(genetic_values(sim)[, "Trait_1"]) / var(phenotypes_wide(sim)$Trait_1),
  real_source = var(genetic_values(sim_real)[, "Trait_1"]) / var(phenotypes_wide(sim_real)$Trait_1)
)
```

The same 20-gene component appears in both phenotypes, but only the derived
source contributes to the genetic value. A phenotype can also be built from
expression alone, with no genotypes. Then individuals come from the expression
columns, there are no markers, and only `transcriptome()` layers are valid.

[]{#mediation-genotype-free}

```{r}
#| label: mediation-genotype-free
simulate_phenotype(expression = tx$expression, seed = 3) |>
  transcriptome(prop = 0.5, n_genes = 15)
```

### Several traits

`prop` can be given per trait. The genes are drawn independently for each trait
under the default `independent` architecture, and the split is reported per trait.

[]{#mediation-multitrait}

```{r}
#| label: mediation-multitrait
sim_two <- simulate_phenotype(geno, n_traits = 2, seed = 6, transcriptome = tx) |>
  transcriptome(prop = c(0.3, 0.1), n_genes = 10) |>
  additive(prop = 0.3, n_qtn = 20)
sim_two
mediation_split(sim_two)
```

Under the `pleiotropy` and `ld` architectures the correlation design covers
the marker layers only. The genetic-mediated part then adds a cross-trait
genetic correlation set by the genes and slopes, so the total genetic
correlation is no longer the targeted one; the package warns in that case. The
[multiple-trait chapter](multitrait.qmd) discusses the targeted correlations.

## Assumptions and theory

The expression layer and its split are package definitions. The equation map
records them as the package's own design, mirroring the scaling of
`additive()`, with the mediation split described as an analogue of the
mediated-expression estimand used in the literature on expression-mediated
heritability (recorded without a verified page, and not relied on here).

### The expression component

For trait $t$ with causal gene set $S$, slopes $b_g$, and expression $E_g$,

$$
T_x = c \sum_{g \in S} w_g z_g, \qquad
w_g = \frac{b_g}{\max_{g' \in S} \lvert b_{g'} \rvert}, \qquad
z_g = \frac{E_g - \bar E_g}{s_{E_g}}, \qquad
c = \frac{\sqrt{\pi_x}}{s_{\sum_g w_g z_g}}.
$$ {#eq-mediation-layer}

The bar is the sample mean over the scored individuals and $s$ the sample
standard deviation; $\pi_x$ is the requested share `prop`. The component has
sample variance $\pi_x$ exactly. This is implemented in `.tx_raw()` and
`.transcriptome_matrix()`; the chunk above reproduced it.

### The split

Because each gene's expression is its genetic part plus its residual,
$E_g = \mu_g + G_g + R_g$, standardizing both parts by the same denominator
$s_{E_g}$ decomposes the standardized expression additively,
$z_g = z_g^{\text{genetic}} + z_g^{\text{env}}$. Applying the same weights and
the same scale $c$ gives

$$
T_{x,g} = c \sum_{g \in S} w_g \frac{G_g - \bar G_g}{s_{E_g}}, \qquad
T_{x,e} = T_x - T_{x,g}.
$$ {#eq-mediation-split}

The reported shares are $\operatorname{Var}(T_{x,g})/V_P$,
$\operatorname{Var}(T_{x,e})/V_P$, and $2\operatorname{Cov}(T_{x,g}, T_{x,e})/V_P$,
with $V_P$ the sample variance of the phenotype. They sum to
$\operatorname{Var}(T_x)/V_P$ by the sample variance identity of
@sec-variance-derivation. The covariance term is close to zero because the
generator draws the genetic and residual parts of expression independently, but
it is reported rather than assumed. The split is computed in
`.tx_raw(which = "genetic")` and `.mediation_budget()`, averaged over
simulation replications.

### The realized broad-sense ratio

With $\text{Gen}$ the sum of the marker layers, the genetic value is
$\text{Gen} + T_{x,g}$ and the realized ratio is

$$
H^2_{\text{realized}} = \frac{\operatorname{Var}(\text{Gen} + T_{x,g})}{V_P}.
$$ {#eq-mediation-h2}

For a real expression source, $T_{x,g}$ is set to zero. The ratio is a sample
diagnostic, not a population parameter, for the reasons given in the
[first chapter](first-simulation.qmd). Each gene's `var_explained` is
$(c\,w_g)^2 / V_P$, a marginal share that ignores covariance among the genes'
standardized expressions, so the shares need not sum to the component's share.

## Implementation and practical limits

The layer lives in `transcriptome()` (R/transcriptome_layer.R) and the scoring
and split in the realization helpers of R/grammar_realize.R. The gene rows of
`qtn_table()` are written by the same export code as marker rows, with
coordinates and allele frequency left missing.

- The layer scores the latent normalized expression, never the counts produced
  by `observe_counts()`. Count data are an observation layer only.
- `prop` for the expression layer is outside the marker `h2` budget; the sum of
  all layer shares is checked against one.
- With `reps`, the residual is divided by the square root of the number of
  records, while the genetic value and the expression component are unchanged.
- A real expression matrix contributes no genetic variance by definition, even
  if its genes are in fact heritable. Use a derived transcriptome, possibly
  calibrated with `mimic`, when the genetic content matters.
- Explicit `genes` must name rows of the expression source without duplicates;
  explicit `slopes` must have one finite value per causal gene.
- Cross-population reuse follows `predict()` from the previous chapter: a
  transcriptome predicted for new individuals can be passed to
  `simulate_phenotype()` for those individuals, but the phenotype's residual and
  layer draws are new.
