# Pleiotropy, linkage, and multiple traits {#sec-multitrait}

Two traits can be correlated because the same loci affect both (pleiotropy),
because their distinct causal loci are inherited together (linkage), or
because their non-genetic influences are shared (residual correlation). The
package separates these mechanisms into three architectures and a residual
correlation argument. This chapter simulates each one on the maize panel,
compares requested with realized correlations, and explains which requests are
attainable and why a realized correlation is a random quantity.

[]{#multitrait-setup}

```{r}
#| label: multitrait-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-multitrait-example}

[]{#multitrait-library}

```{r}
#| label: multitrait-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)

data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)

genetic_cor <- function(sim) cor(genetic_values(sim))
```

The helper `genetic_cor()` is used throughout: it returns the sample
correlation matrix of the simulated genetic values, the realized genetic
correlation.

### Independent traits

`architecture = "independent"` draws separate causal markers for each trait.
`prop` may be a vector with one share per trait.

[]{#multitrait-independent}

```{r}
#| label: multitrait-independent
indep <- simulate_phenotype(geno, n_traits = 3, seed = 31) |>
  additive(prop = c(0.5, 0.3, 0.7), n_qtn = 20)
print(indep)

round(genetic_cor(indep), 3)

qtn_table(indep) |>
  count(trait)
```

The realized genetic correlations are not exactly zero. Different loci on the
same genome still produce correlated values by chance, and in a panel with
extensive linkage disequilibrium the chance correlations are not negligible.
`distinct_chr = TRUE` puts each trait's loci on different chromosomes, which
removes within-chromosome linkage between traits but not chance correlation.

[]{#multitrait-distinct-chr}

```{r}
#| label: multitrait-distinct-chr
indep_chr <- simulate_phenotype(geno, n_traits = 3, seed = 31, distinct_chr = TRUE) |>
  additive(prop = 0.5, n_qtn = 20)

with(qtn_table(indep_chr), table(trait, chr))
```

Each trait's loci occupy a disjoint set of chromosomes.

### Pleiotropy with a controlled genetic correlation

`architecture = "pleiotropy"` draws one set of loci shared by every trait and
assigns each locus correlated effects across traits, so that the genetic
correlation targets `cor`. The algorithm is PleioArch (Prado et al., in
preparation), adopted as the package's pleiotropy engine; the package prints a
citation notice the first time `cor` is used in a session.

[]{#multitrait-pleiotropy}

```{r}
#| label: multitrait-pleiotropy
pleio <- simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                            cor = 0.6, seed = 32) |>
  additive(prop = c(0.5, 0.3), n_qtn = 50)
print(pleio)

qtn_table(pleio) |>
  select(trait, snp, effect, var_explained) |>
  head(4)

tibble(
  quantity = c("Genetic correlation", "Shared causal markers"),
  parameter = c(0.6, 50),
  simulated = c(genetic_cor(pleio)[1, 2], length(unique(qtn_table(pleio)$snp)))
)
```

The effects are no longer a geometric series: they are drawn from a bivariate
normal distribution whose covariance encodes the requested correlation, and
each marker has one effect per trait. The realized correlation differs from
0.6, as the theory section explains.

### Co-heritability

`coheritability()` standardizes the genetic covariance by the phenotypic
standard deviations. Its diagonal is the realized broad-sense heritability of
each trait and its off-diagonal is $r_G h_1 h_2$, the quantity that predicts
the correlated response of one trait to selection on the other in a purely
additive model (Falconer and Mackay 1996, correlated response to selection).

[]{#multitrait-coheritability}

```{r}
#| label: multitrait-coheritability
coheritability(pleio)

h <- sqrt(diag(coheritability(pleio)))
tibble(
  quantity = "Genetic correlation x h1 x h2",
  from_coheritability = coheritability(pleio)[1, 2],
  from_components = genetic_cor(pleio)[1, 2] * h[1] * h[2]
)
```

### Partial pleiotropy

`pi` is the pleiotropic share of each trait's genetic variance. With
`pi = 0.5`, half of each trait's additive variance comes from shared loci and
half from loci specific to that trait. The shared loci appear in both traits'
rows of `qtn_table()`; the specific loci appear in one.

[]{#multitrait-partial}

```{r}
#| label: multitrait-partial
partial <- simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                              cor = 0.4, pi = 0.5, seed = 32) |>
  additive(prop = 0.5, n_qtn = 50)

qtn_table(partial) |>
  count(snp, name = "traits_affected") |>
  count(traits_affected, name = "markers")

tibble(
  quantity = "Genetic correlation",
  parameter = 0.4,
  simulated = genetic_cor(partial)[1, 2]
)
```

Not every combination of `cor` and `pi` is possible. A correlation can only be
as strong as the shared part of the variance allows; the package refuses an
infeasible request instead of approximating it.

[]{#multitrait-infeasible}

```{r}
#| label: multitrait-infeasible
#| error: true
simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                   cor = 0.8, pi = 0.5, seed = 32) |>
  additive(prop = 0.5, n_qtn = 50)
```

### A major pleiotropic locus

`n_pleio_major` and `prop_var_major` reserve a share of the shared variance
for a few major loci. The first row of the table is the major locus.

[]{#multitrait-major}

```{r}
#| label: multitrait-major
major <- simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                            cor = 0.6, n_pleio_major = 1, prop_var_major = 0.5,
                            seed = 32) |>
  additive(prop = 0.5, n_qtn = 50)

qtn_table(major) |>
  filter(trait == "Trait_1") |>
  select(snp, effect, var_explained) |>
  head(3)
```

### Residual correlation

`resid_cor` sets the correlation of the residuals. The phenotypic correlation
then mixes the genetic and residual correlations, weighted by the variance
shares. The chunk recovers the residuals by subtracting genetic values from
phenotypes and reports all three correlations.

[]{#multitrait-resid-cor}

```{r}
#| label: multitrait-resid-cor
mixed <- simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                            cor = 0.6, resid_cor = -0.3, seed = 32) |>
  additive(prop = 0.5, n_qtn = 50)

wide <- phenotypes_wide(mixed)
g <- genetic_values(mixed)[wide$id, ]
e <- as.matrix(wide[, c("Trait_1", "Trait_2")]) - g

tibble(
  correlation = c("Genetic", "Residual", "Phenotypic"),
  parameter = c(0.6, -0.3, NA),
  simulated = c(cor(g)[1, 2], cor(e)[1, 2], cor(wide$Trait_1, wide$Trait_2))
)
```

Here the genetic and residual correlations have opposite signs and the
phenotypic correlation sits between them. The theory section gives the
weighting.

### Correlations that decay with distance: repeated measurements

`cor_ar1()` builds a first-order autoregressive correlation matrix,
$R_{ij} = \rho^{|i-j|}$, for traits that are repeated measurements of one
character over time or environments. Matrices are accepted both for `cor` and
for `resid_cor`, and the one-call form sets everything at once.

[]{#multitrait-ar1}

```{r}
#| label: multitrait-ar1
cor_ar1(4, 0.8)

time_points <- simulate_phenotype(
  geno, n_traits = 4, architecture = "pleiotropy",
  cor = cor_ar1(4, 0.8), resid_cor = cor_ar1(4, 0.4),
  h2 = 0.5, n_qtn = 40, seed = 33
)

round(genetic_cor(time_points), 2)
round(coheritability(time_points), 2)
```

Adjacent time points are more strongly correlated than distant ones, as
requested, and the realized values scatter around the targets.

### Linkage without pleiotropy

`architecture = "ld"` simulates two traits whose causal loci are distinct
markers in linkage disequilibrium. For each locus of trait 1 the package
finds a partner on the same chromosome whose squared correlation with it lies
in the window `[r2_min, r2_max]`, and that partner becomes a causal locus of
trait 2. `qtn_table()` reports each pair and its squared correlation.

[]{#multitrait-ld}

```{r}
#| label: multitrait-ld
linked <- simulate_phenotype(geno, n_traits = 2, architecture = "ld",
                             r2_min = 0.4, r2_max = 0.9, seed = 34) |>
  additive(prop = 0.5, n_qtn = 10)
print(linked)

qtn_table(linked) |>
  filter(trait == "Trait_1") |>
  select(QTN_t1, QTN_t2, chr, ld_r2) |>
  head(5)
```

No marker is causal for both traits, yet the genetic values are strongly
correlated through linkage. Whether that correlation is positive or negative
depends on which alleles travel together. `ld_phase` reads the sign of each
pair's dosage correlation and sets trait 2's effect so that every pair pushes
the correlation one way.

[]{#multitrait-ld-phase}

```{r}
#| label: multitrait-ld-phase
ld_phase_cor <- function(phase) {
  simulate_phenotype(geno, n_traits = 2, architecture = "ld",
                     r2_min = 0.4, r2_max = 0.9, ld_phase = phase, seed = 34) |>
    additive(prop = 0.5, n_qtn = 10) |>
    genetic_cor() |>
    (\(m) m[1, 2])()
}

tibble(
  ld_phase = c("coded", "coupling", "repulsion"),
  genetic_correlation = sapply(c("coded", "coupling", "repulsion"), ld_phase_cor)
)
```

Under the default `"coded"` the sign follows the arbitrary marker coding;
`"coupling"` makes it positive and `"repulsion"` negative, with the same loci
and the same squared correlations.

### How variable is a realized correlation?

The same request repeated with different seeds gives a spread of realized
correlations. The spread is wider with few shared loci.

[]{#multitrait-spread}

```{r}
#| label: multitrait-spread
realized <- function(n_qtn, seeds = 1:20) {
  sapply(seeds, function(s) {
    simulate_phenotype(geno, n_traits = 2, architecture = "pleiotropy",
                       cor = 0.6, seed = s) |>
      additive(prop = 0.5, n_qtn = n_qtn) |>
      genetic_cor() |>
      (\(m) m[1, 2])()
  })
}

tibble(
  shared_loci = c(5, 50),
  parameter = 0.6,
  mean = c(mean(realized(5)), mean(realized(50))),
  sd = c(sd(realized(5)), sd(realized(50))),
  min = c(min(realized(5)), min(realized(50))),
  max = c(max(realized(5)), max(realized(50)))
)
```

[]{#multitrait-plot}

```{r}
#| label: multitrait-plot
#| fig-height: 6
#| fig-cap: "Diagnostic plots for the two-trait pleiotropy simulation, including the realized genetic correlation."
plot(pleio)
```

## Assumptions and theory

### The pleiotropic effect draw

Let $V_t$ be trait $t$'s layer share (`prop`), $\pi_t$ its pleiotropic share
(`pi`, one by default), and $r_{ts}$ the requested correlation. The shared
loci receive effect vectors drawn from a multivariate normal distribution
with covariance

$$
\Sigma_{tt} = \pi_t V_t, \qquad \Sigma_{ts} = r_{ts}\sqrt{V_t V_s},
$$ {#eq-pleio-covariance}

divided by the number of shared loci, and each trait-specific locus receives
an independent normal effect with variance $(1-\pi_t)V_t$ divided by the
number of specific loci (PleioArch, Prado et al., in preparation; the
reference implementation is bundled with the package as `simulateEffects.R`,
page not applicable; implemented in `.pleio_draw()`). Each effect is then
divided by $\sqrt{2\,\mathrm{MAF}(1-\mathrm{MAF})}$, the Hardy-Weinberg
standard deviation of the genotype code, which converts an effect per standard
deviation of the genotype code into an effect per allele.

With these moments, the expected genetic covariance between traits is
$r_{ts}\sqrt{V_t V_s}$ and each trait's expected genetic variance is $V_t$, so
`cor` is the ratio of expected moments. The realized correlation is the ratio
of sample moments and is therefore random. It converges to `cor` as the number
of shared loci and the number of individuals grow, provided the causal loci
are in approximate linkage equilibrium. With few shared loci it is attenuated
toward zero on average; the package documentation reports an average of about
0.82 times `cor` with two shared loci. Strong linkage disequilibrium among the
causal loci can prevent convergence altogether. These statements are package
documentation backed by numerical checks, not literature results.

The layer rescaling of @sec-scaling-derivation does not change a correlation,
because it multiplies each trait's values by a constant. It does fix each
trait's realized variance exactly at $V_t$, so the only randomness left in the
correlation is in the covariance.

### Attainable requests

The covariance matrix in @eq-pleio-covariance must be positive semi-definite.
For two traits this reduces to

$$
r^2 \le \pi_1 \pi_2 ,
$$ {#eq-pleio-feasible}

which is the constraint printed by the error in the worked example. For more
traits the package checks the matrix with diagonal $\pi_t$ and off-diagonal
$r_{ts}$ by its smallest eigenvalue (algebraic generalization; the two-trait
form follows the reference implementation). The rule has a plain reading: a
correlation is carried only by the shared loci, so it cannot exceed what the
shared variance can produce.

When the traits' layer shares are not proportional across layers, for example
mostly additive in one trait and mostly dominance in the other, the total
genetic correlation falls short of `cor`; the package derives the large-sample
target by covariance addition and warns when it differs from `cor` by more
than one percent (package derivation, `.pleio_total_cor_check()`).

### Phenotypic correlation from genetic and residual correlation

With genetic values $g$ and residuals $e$ for two traits and $y = g + e$,

$$
\operatorname{Cov}(y_1, y_2) = \operatorname{Cov}(g_1, g_2)
  + \operatorname{Cov}(e_1, e_2)
  + \operatorname{Cov}(g_1, e_2) + \operatorname{Cov}(e_1, g_2).
$$ {#eq-phenotypic-covariance}

In expectation the cross terms vanish, and dividing by the phenotypic
standard deviations gives the familiar
$r_P = r_G h_1 h_2 + r_E \sqrt{(1-h_1^2)(1-h_2^2)}$. The simulated values are
sample quantities, so the cross terms are not exactly zero and the realized
phenotypic correlation deviates slightly from this formula. The first term is
the off-diagonal element of `coheritability()`.

In fixed residual mode, `resid_cor` mixes independently drawn residuals
through the Cholesky factor of the requested matrix and re-standardizes each
trait's residual to its exact variance, so the residual variances are
unchanged and only their correlation is induced. In random mode the residuals
are a multivariate normal draw with the requested covariance as a parameter.
A `vqtl()` component is drawn independently per trait and dilutes the
residual correlation.

### Linkage disequilibrium between distinct loci

Under `"ld"` the squared correlation between two markers' genotype codes is
the package's measure of linkage disequilibrium, the genotypic or composite
$r^2$ (standard Pearson correlation on dosages; page unverified). The default
`partner = "strongest"` takes the in-window partner with the highest $r^2$,
so the realized values skew toward `r2_max`; `partner = "random"` draws
uniformly within the window. The resulting genetic correlation is a sum over
pairs of $b_1 b_2\, r$, where $r$ is the signed dosage correlation and $b_1$,
$b_2$ are the pair's effects, plus cross-pair terms between loci of different
pairs that happen to be in LD. `ld_phase` controls the sign pair by pair but
not the cross-pair terms, so on a chromosome with many nearby pairs the total
can still differ in sign from the pairs' common sign. The `"ld"` design
follows DECISION-014 of the project decision log; the linkage concept comes
from the original publication, with the v2 algorithm being a package design.

## Implementation and practical limits

The effect draws stay in R and are seeded per layer. The PleioArch draw uses
a symmetric eigenvalue square root of the covariance matrix rather than the
reference implementation's Cholesky factor, which gives the same distribution
but different realizations, and allows exactly singular feasible requests.

- `"ld"` requires `n_traits = 2`, one additive layer, and at most a dominance
  layer on the same linked loci. Epistasis and user-fixed loci are rejected
  under `"ld"` with random pairs; `qtn = list(trait1, trait2)` is accepted for
  direct pairs.
- Under `"pleiotropy"`, user-supplied `effect` is accepted only when no
  correlation is controlled; `qtn` fixes the shared loci but `pi < 1` cannot
  then be used.
- A single shared locus makes the realized correlation plus or minus one;
  the package warns.
- The realized correlation on a fixed sample of individuals cannot beat the
  sampling spread of a correlation over that many individuals, however many
  loci are shared.
- Most of the scatter of the realized correlation around its target comes from
  drawing a finite number of effect pairs, with a standard deviation of about
  $(1-r^2)/\sqrt{n_{qtn}-1}$. Linkage disequilibrium in the maize panel adds to
  it.

::: {.callout-note title="Partial pleiotropy by trait-specific layers"}
`pi` splits one layer into shared and specific loci. For loci that affect
only some traits with layer-specific settings, or to combine several
simulations into one trait, see `complex_phenotypes()` in the
[complex architectures chapter](complex.qmd).
:::
