# Dominance and epistasis {#sec-nonadditive}

Dominance makes heterozygotes deviate from the midpoint of the two
homozygotes; epistasis makes the effect of one locus depend on the genotype at
another. This chapter adds both to an additive trait on real human genotypes,
checks by hand how each component is built, and explains why the realized
genetic variance of a trait with additive and dominance layers on the same loci
is not the sum of the two requested shares. It ends with the orthogonal
genotypic model, which gives a Fisher-orthogonal partition when that is what a
study needs.

[]{#nonadditive-setup}

```{r}
#| label: nonadditive-setup
#| include: false
source("_setup.R")
source("_data.R")
```

## Worked example {#sec-nonadditive-example}

### Why human genotypes

Dominance is defined by heterozygotes. The maize panel of the earlier
chapters is inbred: fewer than half of its markers carry any heterozygote,
and those that do carry very few. The HapMap individuals used
here are outbred, so every common marker has heterozygotes. The
[genotype-input chapter](input.qmd) explains how this dataset is read; the
shared loader below repeats those steps.

[]{#nonadditive-library}

```{r}
#| label: nonadditive-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)

human <- book_human()
tibble(markers = nrow(human), individuals = ncol(human) - 5)
```

### Additive and dominance on the same loci

`dominance()` adds a dominance component. By default it reuses the additive
layer's markers (`same_as_add = TRUE`), so the trait has ten causal loci that
act both additively and through dominance.

[]{#nonadditive-dominance}

```{r}
#| label: nonadditive-dominance
ad <- simulate_phenotype(human, seed = 21) |>
  additive(prop = 0.4, n_qtn = 10) |>
  dominance(prop = 0.2)
print(ad)
```

The requested genetic share is 0.6 but the realized ratio printed by the
object is well below it. The printed note explains why: the additive and
dominance components of the same loci covary, and the covariance is negative
here. The `ad_report` element stores the same accounting as a table.

[]{#nonadditive-ad-report}

```{r}
#| label: nonadditive-ad-report
ad$ad_report |>
  select(trait, requested, realized, var_cA, var_cD, cov2_comp)
```

`var_cA` and `var_cD` are the realized variances of the scaled additive and
dominance components, as shares of the phenotypic variance; `cov2_comp` is
twice their covariance. The three add up to the realized genetic share.

### Rebuild the genetic value by hand

`qtn_table()` lists each layer's markers and effects. The additive component
multiplies the genotype code by the effect; the dominance component multiplies
a heterozygote indicator (one where the code is `0`, zero elsewhere) by the
effect. Each component is centered and scaled to its `prop`, then the two are
added. The code below repeats those steps for every individual and compares
the result with `genetic_values()`.

[]{#nonadditive-rebuild}

```{r}
#| label: nonadditive-rebuild
qt <- qtn_table(ad)
X <- as.matrix(human[, -(1:5)])
rownames(X) <- human$snp

add_rows <- filter(qt, layer == "additive")
dom_rows <- filter(qt, layer == "dominance")

score_a <- colSums(X[add_rows$snp, ] * add_rows$effect)
score_d <- colSums((X[dom_rows$snp, ] == 0) * dom_rows$effect)

scale_to <- function(score, prop) sqrt(prop) * (score - mean(score)) / sd(score)
g_a <- scale_to(score_a, 0.4)
g_d <- scale_to(score_d, 0.2)

truth <- genetic_values(ad)[, "Trait_1"]
tibble(
  id = names(truth),
  package = truth,
  by_hand = (g_a + g_d)[names(truth)]
) |>
  head()

tibble(
  quantity = c("Var(additive component)", "Var(dominance component)",
               "2 Cov(additive, dominance)", "Var(genetic value)"),
  parameter = c(0.4, 0.2, NA, 0.6),
  simulated = c(var(g_a), var(g_d), 2 * cov(g_a, g_d), var(g_a + g_d))
)
```

The two columns agree to rounding. The second table shows that each component
has exactly its requested variance and that the covariance term is what moves
the genetic variance away from 0.6. The sign of that term is explained in the
theory section.

### Dominance on separate loci, with chosen effects

`same_as_add = FALSE` draws new loci for the dominance layer, and `effect`
sets the dominance effects. Here the five dominance loci have no additive
effect, so the two components covary only by chance.

[]{#nonadditive-dominance-separate}

```{r}
#| label: nonadditive-dominance-separate
ad_separate <- simulate_phenotype(human, seed = 21) |>
  additive(prop = 0.4, n_qtn = 10) |>
  dominance(prop = 0.2, same_as_add = FALSE, n_qtn = 5,
            effect = c(0.5, 0.4, 0.3, 0.2, 0.1))
print(ad_separate)

qtn_table(ad_separate) |>
  filter(layer == "dominance") |>
  select(snp, maf, effect, var_explained)
```

### Epistasis: pairs of interacting loci

`epistasis()` adds interacting sets of markers. `n_pairs` is the number of
sets, `interaction` the number of markers per set, and `interaction_type`
says whether each position enters through its genotype code (`"a"`) or its
heterozygote indicator (`"d"`). The default is pairwise additive-by-additive.

[]{#nonadditive-epistasis}

```{r}
#| label: nonadditive-epistasis
ep <- simulate_phenotype(human, seed = 22) |>
  additive(prop = 0.3, n_qtn = 10) |>
  epistasis(prop = 0.2, n_pairs = 3)
print(ep)

qtn_table(ep) |>
  filter(layer == "epistasis") |>
  select(set, snp, maf, effect)
```

Each set has one effect, shared by its members, and `var_explained` is not
reported for epistatic members because a single marker has no marginal share
in a product term. The product is formed from centered codes, as the next
chunk verifies by rebuilding the epistatic component.

[]{#nonadditive-epistasis-rebuild}

```{r}
#| label: nonadditive-epistasis-rebuild
epi_rows <- qtn_table(ep) |> filter(layer == "epistasis")
X_centered <- X - rowMeans(X)

set_products <- lapply(split(epi_rows, epi_rows$set), function(rows) {
  rows$effect[1] * apply(X_centered[rows$snp, , drop = FALSE], 2, prod)
})
score_e <- Reduce(`+`, set_products)

add_rows <- qtn_table(ep) |> filter(layer == "additive")
score_a <- colSums(X[add_rows$snp, ] * add_rows$effect)
rebuilt <- scale_to(score_a, 0.3) + scale_to(score_e, 0.2)

truth <- genetic_values(ep)[, "Trait_1"]
all.equal(unname(truth), unname(rebuilt[names(truth)]))
```

Higher-order interactions and dominance-type positions follow the same
construction. The next simulation stacks an additive-by-dominance layer and a
dominance-by-dominance layer, and a three-way additive interaction.

[]{#nonadditive-epistasis-types}

```{r}
#| label: nonadditive-epistasis-types
ep_mixed <- simulate_phenotype(human, seed = 22) |>
  additive(prop = 0.3, n_qtn = 10) |>
  epistasis(prop = 0.1, n_pairs = 2, interaction_type = c("a", "d")) |>
  epistasis(prop = 0.1, n_pairs = 2, interaction_type = c("d", "d"))
print(ep_mixed)

ep_three <- simulate_phenotype(human, seed = 22) |>
  additive(prop = 0.3, n_qtn = 4) |>
  epistasis(prop = 0.2, n_pairs = 2, interaction = 3)

qtn_table(ep_three) |>
  filter(layer == "epistasis") |>
  select(set, snp, effect)
```

### The orthogonal genotypic model

In the layers above, the additive and dominance shares are set directly and
the components are not orthogonal. `additive(orthogonal = TRUE, a =, d =)`
instead assigns each locus a genotypic value of $-a$, $+d$, $+a$ for zero, one,
and two copies of the counted allele, scales the whole genotypic value to
`prop`, and lets the additive and dominance variances emerge from $a$, $d$, and
the allele frequencies.

[]{#nonadditive-orthogonal}

```{r}
#| label: nonadditive-orthogonal
orth <- simulate_phenotype(human, h2 = 0.6, seed = 23) |>
  additive(orthogonal = TRUE, a = rep(0.5, 10), d = rep(0.5, 10), n_qtn = 10)
print(orth)

qtn_table(orth) |>
  select(snp, maf, effect, d, var_explained) |>
  head(4)
```

The budget now shows an `additive` row, a `dominance` row, and an
`add_dom_cov` row. Vectors are passed for `a` and `d` because a single number
for `a` is the base of a geometric series, which would give each locus a
different degree of dominance. The degree of dominance is `d / abs(a)`, one
here, which is complete dominance at every locus. The next table compares no dominance,
complete dominance, and overdominance on the same loci.

[]{#nonadditive-degree}

```{r}
#| label: nonadditive-degree
degree_budget <- function(d) {
  budget <- simulate_phenotype(human, h2 = 0.6, seed = 23) |>
    additive(orthogonal = TRUE, a = rep(0.5, 10), d = rep(d, 10), n_qtn = 10) |>
    getElement("var_budget")
  tibble(
    degree = d / 0.5,
    additive = round(budget$prop[budget$component == "additive"], 3),
    dominance = round(budget$prop[budget$component == "dominance"], 3),
    add_dom_cov = round(budget$prop[budget$component == "add_dom_cov"], 3)
  )
}

bind_rows(degree_budget(0), degree_budget(0.5), degree_budget(1))
```

With `d = 0` the model is purely additive and the dominance row is zero.
With complete dominance a substantial part of the genotypic variance is
dominance variance, and with overdominance the dominance row grows further.
The three rows always add up to the requested 0.6; the additive row is not
simply what is left over, because the covariance row also changes. The `add_dom_cov` row is the
realized covariance between the additive and dominance parts of the genotypic
value; it is close to zero in a large random-mating sample and nonzero in a
structured sample such as this multi-population panel.

### The one-call model

`model = "AD"` splits `h2` equally between an additive and a dominance layer
on shared loci. It is a shortcut for the first simulation of this chapter, not
for the orthogonal model.

[]{#nonadditive-model-ad}

```{r}
#| label: nonadditive-model-ad
one_call <- simulate_phenotype(human, h2 = 0.6, n_qtn = 10, model = "AD", seed = 21)
one_call$var_budget
```

## Assumptions and theory

### Three design columns

For individual $i$ and locus $j$, let $x_{ij} \in \{-1, 0, 1\}$ be the genotype
code and $h_{ij} = \mathbb{1}[x_{ij} = 0]$ the heterozygote indicator. The
raw components are package definitions (`.component_raw()` and
`.epi_unit_column()`):

$$
c^{A}_i = \sum_j b_j\, x_{ij}, \qquad
c^{D}_i = \sum_j b_j\, h_{ij}, \qquad
c^{E}_i = \sum_{s} b_s \prod_{k \in s} \big(t_{ik} - \bar t_k\big),
$$ {#eq-nonadditive-components}

where $s$ indexes interacting sets and $t_{ik}$ is either $x_{ik}$ (position
type `"a"`) or $h_{ik}$ (type `"d"`), centered by its sample mean before the
product. Each component is centered and scaled to its `prop` exactly as in
@sec-scaling-derivation, and the genetic value is the sum of the scaled
components. The worked example rebuilt the first two and the third from these
definitions.

### Why shared loci make the components covary

Take one locus with counted-allele frequency $p$ in Hardy-Weinberg
proportions. The genotype code has mean $2p - 1$ and the heterozygote
indicator has mean $2p(1-p)$. Their covariance is

$$
\operatorname{Cov}(x, h) = \mathbb{E}[x h] - \mathbb{E}[x]\mathbb{E}[h]
 = 0 - (2p - 1)\, 2p(1-p),
$$ {#eq-dominance-covariance}

because $x h$ is zero for every genotype. The term is negative when the
counted allele is the major allele ($p > 0.5$) and positive when it is the
minor allele. This is an algebraic consequence of the two codings, also stated
in the package documentation of `dominance()`. The default orientation of `as_numeric()` codes the major allele `1`, so in
the HapMap data every locus contributes a negative cross term, and the sum of those terms is the negative covariance
seen in the worked example. The effect does not shrink with sample size: it is
a property of the codings and the frequencies, not sampling noise.

Consequently, for one additive and one dominance layer on shared loci,

$$
\operatorname{Var}(g) = \pi_A + \pi_D + 2\operatorname{Cov}(g^A, g^D),
$$ {#eq-ad-variance}

which the `ad_report` table reports as `var_cA + var_cD + cov2_comp`. The
requested shares are variances of components, not an orthogonal partition of
the genetic variance. `ad_report` also reports a Fisher-type split of the
block into a breeding-value part and a dominance-deviation part (`var_A`,
`var_D`, `cov2_AD`), which is the decomposition the orthogonal model targets.

Epistatic products are only partly orthogonal. Under linkage equilibrium the
centered product of two additive codes is uncorrelated with each locus's main
effect at any allele frequency. Linkage disequilibrium or population
structure makes it correlated. In addition, a centered dominance factor is not
orthogonal to the additive code at its own locus, so additive-by-dominance and
dominance-by-dominance products also carry additive-by-additive pieces.
The requested epistatic `prop` is therefore a simulation quantity, not an
orthogonal variance component in the sense of Cockerham or NOIA. The package
documents this limitation.

### The orthogonal genotypic model

Under `orthogonal = TRUE` the locus value is $a_j x_{ij} + d_j h_{ij}$, the
classical $-a$, $+d$, $+a$ genotypic values (Falconer and Mackay 1996,
genotypic values; page unverified). The additive part of the genetic value
uses the average effect of allele substitution

$$
\alpha_j = a_j + d_j\,(1 - 2p_j),
$$ {#eq-average-effect}

(Falconer 1985, *Genet. Res.* 46:337-347; Falconer and Mackay 1996, page
unverified; Fisher 1918), implemented in `.avg_effect()`, and the breeding
value is $A_i = \sum_j \alpha_j (n_{ij} - 2p_j)$ with $n_{ij}$ the number of
copies of the counted allele. The dominance deviation is the remainder
$D = g - A$. Under random mating the two are uncorrelated in expectation;
the package reports the realized `add_dom_cov` instead of assuming it is
zero. Linkage disequilibrium between loci does not break this orthogonality
in a no-epistasis model, but population structure or selection can.

The expected variances under Hardy-Weinberg proportions are
$V_A = \sum_j 2p_jq_j\alpha_j^2$ and $V_D = \sum_j (2p_jq_jd_j)^2$
(Falconer and Mackay 1996, page unverified). The worked example reports
realized rather than expected values, which is why the rows change with the
sample.

### Which model to use

- Use the variance-partition layers when the study needs a fixed share of
  dominance or epistatic variance in a simulated sample and treats the
  components as simulation quantities. Read `ad_report` before interpreting
  the realized ratio.
- Use the orthogonal model when the study needs breeding values, a defined
  degree of dominance, or an additive/dominance split that estimation methods
  will be compared against. The breeding value then feeds the selection and
  prediction chapters.

## Implementation and practical limits

- `dominance()` and dominance-type epistatic positions need heterozygotes.
  On the inbred maize panel, markers without heterozygotes are skipped with a
  warning and the remaining loci absorb the layer's share; `filter_geno(hets =
  "include")` restricts the candidate markers beforehand, and an F2 or an
  outbred population avoids the problem.
- The orthogonal model cannot be combined with a separate `dominance()`
  layer, with `vary_qtn`, or with the correlation-controlled `"pleiotropy"`
  and `"ld"` architectures; `qtn` alone is accepted.
- `var_explained` is reported per marker for additive and dominance layers
  only.
- Under `architecture = "ld"` a dominance layer must reuse the linked loci or
  supply disjoint linked pairs, and `epistasis()` is not available.
- The realized ratio printed by the object is a sample diagnostic. With
  shared loci it can differ from the requested share by a structural amount
  that depends on which allele is coded `1`; the genotype-input chapter shows
  how to control that orientation.

::: {.callout-note title="Allele orientation matters here"}
The sign of the additive-dominance covariance follows $1 - 2p$ for the
counted allele. Reading the same data with the minor allele coded `1` flips
the sign at every locus and can move the realized ratio to the other side of
the requested share. Record the orientation with your simulation.
:::
