# Contributions and cross usefulness {#sec-optimization}

Truncation selection keeps the best candidates and discards the rest. Over
cycles that erodes diversity and, with it, future gain. This chapter asks two
questions a breeder faces before mating: how much should each candidate
contribute to the next generation, and which crosses are worth making? The
first is answered by optimum contribution selection, the second by the
usefulness of a cross. Both are computed on simulated maize lines, so every
expectation can be checked against realized progeny.

[]{#optimization-setup}

```{r}
#| label: optimization-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-optimization-example}

### Load the package and candidates

[]{#optimization-library}

```{r}
#| label: optimization-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

The candidates are the first 40 inbred lines of the bundled maize panel
(see [A first simulation](first-simulation.qmd)), with an additive trait of 50
causal markers. `as_population()` turns the numeric genotype table into a
`Population`, the object the crossing and selection functions operate on.

[]{#optimization-candidates}

```{r}
#| label: optimization-candidates
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
pop <- as_population(geno, individuals = 1:40)
sim <- simulate_phenotype(pop, h2 = 0.5, seed = 11) |>
  additive(n_qtn = 50)
print(sim)
```

The realized genetic-to-phenotypic variance ratio differs from the requested
0.5 for the reasons explained in the [first chapter](first-simulation.qmd):
the sample covariance between the genetic and residual components is free, and
with only 40 individuals it can be large.

### Relatedness among candidates

`g_matrix()` builds the genomic relationship matrix from the marker dosages.
Its diagonal exceeds one by the genomic inbreeding of each line relative to
the current sample, and its off-diagonal entries measure how much two lines
share marker alleles beyond what the sample's allele frequencies predict.

[]{#optimization-g}

```{r}
#| label: optimization-g
G <- g_matrix(pop)
dim(G)
round(G[1:4, 1:4], 3)
summary(diag(G) - 1)
```

The lines are inbred, so their genomic inbreeding is close to one. Because
`G` is built from the current sample's allele frequencies, the diagonal is not
inbreeding relative to a founder population; the function's `base_freq`
argument fixes a reference base when one is needed.

### Optimum contributions

`optimum_contribution()` chooses a contribution vector `c`, one value per
candidate, that is non-negative and sums to one. It maximizes the expected
merit of the parents minus a penalty on their group coancestry. The merit is
the simulated breeding value by default. The penalty `lambda` sets the
trade-off.

[]{#optimization-ocs}

```{r}
#| label: optimization-ocs
oc <- optimum_contribution(sim, merit = "bv", lambda = 2)
print(oc)
```

The printed summary reports the expected merit `c'g`, the group coancestry
`0.5 c'Gc`, and the largest contributions. We confirm the two defining
properties of the solution directly: the contributions sum to one and the
reported coancestry equals the quadratic form computed from `G`.

[]{#optimization-ocs-check}

```{r}
#| label: optimization-ocs-check
cc <- oc$contributions
bv <- genetic_values(sim)[, 1]
tibble(
  quantity = c("Sum of contributions", "Expected merit", "Group coancestry"),
  reported = c(sum(cc), oc$merit, oc$coancestry),
  recomputed = c(
    sum(cc),
    sum(cc * bv[names(cc)]),
    0.5 * drop(crossprod(cc, G %*% cc))
  )
)
```

### The gain-diversity frontier

Repeating the optimization over a range of penalties traces the frontier
between expected merit and group coancestry. At `lambda = 0` nothing
penalizes relatedness, so all weight goes to the single best candidate. Larger
penalties spread the contributions over more parents at a cost in expected
merit.

[]{#optimization-frontier}

```{r}
#| label: optimization-frontier
frontier <- lapply(c(0, 0.5, 2, 10, 50), function(lambda) {
  o <- optimum_contribution(sim, lambda = lambda)
  tibble(
    lambda = lambda, n_parents = o$n_parents,
    merit = o$merit, coancestry = o$coancestry
  )
}) |>
  bind_rows()
frontier
```

Breeders more often think in terms of a coancestry they are willing to
accept than in terms of a penalty. `target_coancestry` finds the penalty whose
solution has exactly that group coancestry, by bisection on `lambda`.

[]{#optimization-target}

```{r}
#| label: optimization-target
tg <- optimum_contribution(sim, target_coancestry = 0.3)
print(tg)
```

### From contributions to parents

Contributions are fractions; a mating plan needs individuals. `sample_parents()`
fills `n` parent slots. The default `"allocate"` method gives each candidate
the whole-number part of `n` times its contribution, then assigns the
leftover slots to the largest remainders. The `"multinomial"` method draws
slots at random with replacement, weighted by the contributions.

[]{#optimization-parents}

```{r}
#| label: optimization-parents
parents <- sample_parents(oc, pop, n = 20, seed = 12)
print(parents)

slots <- attr(parents, "source") |>
  count(id, name = "slots") |>
  mutate(
    realized = slots / 20,
    optimized = oc$contributions[id]
  ) |>
  arrange(desc(optimized))
slots
```

The `source` attribute records which candidate fills each slot. Every
realized contribution is within one slot of its optimized value. The group
coancestry of the realized set is not guaranteed to equal the optimum, because
coancestry is quadratic in the contributions. The comparison below computes it
for both methods; the multinomial draw usually lands further from the optimum
because its counts carry sampling noise.

[]{#optimization-parents-coancestry}

```{r}
#| label: optimization-parents-coancestry
multinomial <- sample_parents(oc, pop, n = 20, seed = 12, method = "multinomial")

realized_coancestry <- function(parents) {
  counts <- table(attr(parents, "source")$id)
  c_realized <- setNames(rep(0, nrow(G)), rownames(G))
  c_realized[names(counts)] <- as.numeric(counts) / sum(counts)
  0.5 * drop(crossprod(c_realized, G %*% c_realized))
}

tibble(
  set = c("Optimized contributions", "allocate, n = 20", "multinomial, n = 20"),
  coancestry = c(
    oc$coancestry,
    realized_coancestry(parents),
    realized_coancestry(multinomial)
  )
)
```

The returned `Population` can be mated with `cross()` or
[`mating_design()`](mating.qmd); repeated copies of a parent receive
distinct identifiers because the crossing functions require unique names.

### Which crosses to make

Optimum contributions decide who should be a parent. Usefulness decides which
pairs to cross. `cross_usefulness()` simulates a progeny family for each
candidate pair, scores the progeny on the trait's additive effects, and reports
the family mean, its genetic standard deviation, and the usefulness
$U = \mu + i\,\sigma$, where $i$ is the selection intensity for the best
`select_top` fraction.

For the doubled-haploid and selfing schemes the family is derived from a
single F1 of the cross, which is only well defined for inbred parents. The
maize lines carry a little residual heterozygosity, so we take the first
eight lines and remove the few markers at which any of them is heterozygous.
This is a teaching convenience, not a recommendation for real data; it also
keeps the candidates diverse, which matters because the scoring effects are
standardized over the candidates themselves.

[]{#optimization-inbred}

```{r}
#| label: optimization-inbred
lines <- names(geno)[6:13]
inbred_geno <- filter_geno(
  geno[, c(names(geno)[1:5], lines)],
  hets = "remove", verbose = FALSE
)
inbred <- as_population(inbred_geno)
print(inbred)
```

The usefulness is computed on the scoring model of a simulation built on
these eight lines: the simulation supplies the causal markers and their fixed
additive effects.

[]{#optimization-usefulness}

```{r}
#| label: optimization-usefulness
sim8 <- simulate_phenotype(inbred, h2 = 0.5, seed = 11) |>
  additive(n_qtn = 50)
u <- cross_usefulness(sim8, scheme = "dh", n_progeny = 100, seed = 13)
head(u)
attr(u, "intensity")
```

All 28 pairs of the eight lines were evaluated, ranked best first. The
`intensity` attribute is the normal-theory value for the top ten percent. A
parent-mean ranking would use only the `midparent` column of the comparison below;
usefulness adds the family's genetic spread, which is why crosses with a lower
mean can rank above crosses with a higher one.

[]{#optimization-midparent}

```{r}
#| label: optimization-midparent
template <- template_effects(sim8)
parent_value <- additive_value(inbred, template$qtn, template$a)

u |>
  mutate(midparent = (parent_value[parent1] + parent_value[parent2]) / 2) |>
  select(parent1, parent2, midparent, mean, sd, usefulness) |>
  head(8)
```

### Check the usefulness against realized progeny

We simulate the top-ranked cross ourselves: make the F1, derive 100 doubled
haploids, score them with the same additive effects, and compare the
usefulness to the mean of the best ten progeny.

[]{#optimization-usefulness-check}

```{r}
#| label: optimization-usefulness-check
best <- u[1, ]
f1 <- cross(inbred[best$parent1], inbred[best$parent2], n = 1, seed = 14)
dh <- double_haploid(f1, n = 100, seed = 15)
progeny <- additive_value(dh, template$qtn, template$a)
top10 <- mean(sort(progeny, decreasing = TRUE)[1:10])

tibble(
  quantity = c(
    "Family mean", "Family genetic SD",
    "Usefulness (normal theory)", "Mean of the best 10 of 100 progeny"
  ),
  predicted = c(best$mean, best$sd, best$usefulness, NA),
  simulated = c(mean(progeny), sd(progeny), NA, top10)
)
```

The two family simulations use different seeds, so the means and standard
deviations differ by Monte Carlo error. The usefulness and the realized
top-fraction mean differ for a second reason explained below.

The `"selfcross"` scheme derives near-inbred lines by repeated selfing
instead of doubling the F1 gametes; its results are usually close to the
doubled-haploid ranking for an additive trait.

[]{#optimization-selfcross}

```{r}
#| label: optimization-selfcross
cross_usefulness(
  sim8, pairs = u[1:3, c("parent1", "parent2")],
  scheme = "selfcross", n_progeny = 100, generations = 5, seed = 13
)
```

## Assumptions and theory

### The optimum contribution problem

Optimum contribution selection maximizes

$$
c'g - \frac{\lambda}{2}\, c'Gc
\qquad\text{subject to}\qquad c \ge 0,\; \mathbf{1}'c = 1,
$$ {#eq-ocs-objective}

where $c$ is the vector of contributions, $g$ the vector of merits, $G$ the
genomic relationship matrix, and $\lambda$ the penalty. The group coancestry
of the chosen parents is $\tfrac12 c'Gc$. The equation map records the
reference as Meuwissen 1997 *J Anim Sci* 75:934–940 (verified; page
unverified). The implementation in `optimum_contribution()` and the internal
optimizer `.frank_wolfe()` is a package definition: an away-step Frank-Wolfe
method on the simplex, with the penalty tuned by bisection in `.tune_lambda()`
when a target coancestry is requested. With `direction = "low"` the merit is
negated.

Because the contributions sum to one, adding a constant to every merit does
not change the solution. The merit scale does matter relative to `lambda`:
doubling all merits halves the effective penalty, which is why the frontier
should be read as a curve over `lambda`, not as a set of universal values.

### The genomic relationship matrix

`g_matrix()` computes

$$
G = \frac{ZZ'}{2\sum_j p_j(1-p_j)}, \qquad Z = M - 2p,
$$ {#eq-vanraden}

with $M$ the individuals-by-markers gene-content matrix (0, 1, 2) and $p_j$
the frequency of the counted allele at marker $j$. This is VanRaden 2008
*J Dairy Sci* 91:4414–4423, method 1 (paper verified; page unverified in the
equation map). Markers monomorphic in the current set are dropped. The
optional `ridge` argument blends $G$ with the identity; that blend is a
package extension for numerical positive definiteness, not VanRaden's blend
with a pedigree matrix.

### Usefulness

The usefulness criterion is

$$
U = \mu + i\,\sigma,
$$ {#eq-usefulness}

with $\mu$ the family mean and $\sigma$ the family genetic standard deviation
of the additive value, both realized by the crossing engine from `n_progeny`
simulated progeny, and $i = \varphi(\Phi^{-1}(1-p))/p$ the standard
normal-theory selection intensity for the fraction $p$ in `select_top`. The
equation map cites Zhong & Jannink 2007 *Genetics* 177:567–576 and Lehermeier
et al. 2017 *Genetics* 207:1651–1661 (verified; pages unverified) for the
criterion and Falconer & Mackay 1996 (author-year, no page asserted) for the
intensity.

Two facts explain why the usefulness need not equal the realized
top-fraction mean. First, $i$ is the infinite-population value; the mean of
the top 10 of 100 ordered normal values is slightly smaller. Second, the
family's genetic values are a sum over a few dozen loci with geometric
effects, so they are not exactly normal; a family dominated by a few large
effects is discrete and skewed, and $U$ can then exceed the family maximum.
`U` is therefore a ranking criterion under a normality assumption, not an
exact prediction.

### Scores are on a fixed scale

The progeny scores in `cross_usefulness()` and the `additive_value()` checks
above use the additive effects frozen from the template simulation,
$g_i = \sum_j x_{ij} e_j$ with $e_j$ already multiplied by the layer's
$\sqrt{\pi}/\mathrm{sd}$ factor (`.additive_model()`, a package definition
mirroring `.genetic_matrix()`). Scoring a family by re-running
`simulate_phenotype()` on it would re-standardize the genetic component within
each family and destroy the comparability of means and standard deviations
between crosses.

## Implementation and practical limits

The optimizer, the relationship matrix, and the usefulness calculations are
deterministic R code; the random steps are the simulated meioses of
`cross_usefulness()` and the slot ordering or draws of `sample_parents()`,
all seeded through R.

- `optimum_contribution()` takes exactly one of `lambda`, `target_coancestry`
  or `max_coancestry`. A target above the coancestry of the unconstrained
  optimum cannot be reached by a non-negative penalty; the function warns and
  uses `lambda = 0`.
- The default merit `"bv"` is the transmitting average effect
  $\alpha_j = a_j + d_j(q_j - p_j)$, computed from the simulation's own
  effects. It is refused for models with an epistasis layer; pass a numeric
  `merit` there, for example estimated breeding values from
  [`predict_ebv()`](prediction.qmd).
- `g_matrix()` refuses missing genotypes and any coding other than `-1`, `0`,
  `1`; impute first with `as_numeric(impute = )`.
- `sample_parents()` with `"allocate"` can give a small contributor no slot
  when `n` is small. Increase `n` if every contributor must appear.
- `cross_usefulness()` evaluates all pairs by default; the count grows with
  the square of the number of candidates. Pass `pairs` for large panels.
- Only the additive component is scored; dominance and epistasis layers are
  ignored, and the function warns when a parent is heterozygous at a scoring
  locus under the `"dh"` or `"selfcross"` schemes.

::: {.callout-tip title="Combining the two tools"}
A common pipeline is to compute contributions with
`optimum_contribution()`, fill the parent slots with `sample_parents()`,
and then rank the pairs among those parents with `cross_usefulness()`
before calling [`mate()`](mating.qmd). The
[breeding schemes](programs.qmd) chapter shows how such a cycle is repeated.
:::
