# Inspecting and exporting results {#sec-output}

A simulation is only useful once its phenotypes, genetic values, and causal
markers reach the analysis that needs them. This chapter shows how to read a
`phenotype_sim` object in R, how its tables are shaped when there are several
traits and replications, what the diagnostic plot panels mean, and how to write
the results as delimited text or JSON, including the marker data split into
causal and non-causal files. It also checks that what is written can be read
back unchanged.

[]{#output-setup}

```{r}
#| label: output-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-output-example}

### A two-trait, three-replication simulation

The maize example from [the first chapter](first-simulation.qmd) is extended to
two traits and three simulation replications, with an additive layer and a
dominance layer on the same loci. The inbred panel has very few heterozygotes,
so a warning explains that dominance rests on fewer loci than requested; the
[dominance and epistasis chapter](nonadditive.qmd) discusses that choice.
Here the purpose is only to produce an object with enough structure to show
every output shape.

[]{#output-library}

```{r}
#| label: output-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
library(readr)
```

[]{#output-model}

```{r}
#| label: output-model
#| warning: false
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)

sim <- simulate_phenotype(geno, n_traits = 2, n_reps = 3, seed = 42) |>
  additive(prop = 0.4, n_qtn = 10) |>
  dominance(prop = 0.1, same_as_add = TRUE)
```

### Read the printed summary

Printing the object is the first inspection. The method reports the genotype
source, the number of traits, the architecture, the seed, and the variance
partition that was requested for each layer, followed by the realized
genetic-to-phenotypic variance ratio for each trait.

[]{#output-print}

```{r}
#| label: output-print
print(sim)
```

When additive and dominance layers share loci, the summary also decomposes the
realized share into the two component variances and their covariance. That
block is a diagnostic of this sample, in the sense explained in the
[first chapter](first-simulation.qmd#sec-first-example): the requested
proportions are what the layers were scaled to, and the realized values
include the covariance between the additive and dominance components. As the
printout says, that covariance is structural: it depends on allele frequencies
and on which allele is coded +1, not on sampling noise.

### Long and wide phenotype tables

`phenotypes_long()` returns one row per individual, trait, and replication.
`phenotypes_wide()` returns one row per individual and replication, with one
column per trait. Both are package definitions of the output layout, not
transformations the reader has to perform.

[]{#output-long-wide}

```{r}
#| label: output-long-wide
long <- phenotypes_long(sim)
wide <- phenotypes_wide(sim)

tibble(
  layout = c("long", "wide"),
  rows = c(nrow(long), nrow(wide)),
  columns = c(ncol(long), ncol(wide))
)

head(long)
head(wide)
```

The long table has 280 individuals times 2 traits times 3 replications rows.
The wide table collapses the trait dimension into columns, so it has half as
many rows and the same four columns, with `trait` and `value` replaced by
`Trait_1` and `Trait_2`. Count rows per trait and replication to
confirm that every combination is present once:

[]{#output-counts}

```{r}
#| label: output-counts
long |>
  count(trait, rep)
```

The long layout is what mixed-model software and plotting packages usually
expect. The wide layout makes between-trait comparisons direct, for example a
phenotypic correlation within a replication:

[]{#output-wide-cor}

```{r}
#| label: output-wide-cor
wide |>
  group_by(rep) |>
  summarise(phenotypic_correlation = cor(Trait_1, Trait_2))
```

### Genetic values and the QTN table

`genetic_values()` returns an individuals-by-traits matrix of the genetic
component before the residual was added. `qtn_table()` lists every causal
marker with the effect it was given, one row per marker, trait, and layer.

[]{#output-truth}

```{r}
#| label: output-truth
truth <- genetic_values(sim)
head(truth)

qtn_table(sim) |>
  select(trait, layer, snp, chr, pos, maf, effect, var_explained) |>
  head()
```

The column meanings are package definitions:

| Column | Meaning |
|---|---|
| `trait`, `layer` | The trait and the layer (`additive`, `dominance`, `epistasis`, `vqtl`, `transcriptome`) that uses the marker |
| `set` | For epistatic layers, the interacting set a marker belongs to; `NA` otherwise |
| `snp`, `chr`, `pos` | Marker identifier and map position copied from the genotype input; for a transcriptome layer `snp` holds a gene name |
| `maf` | Minor allele frequency in the simulated individuals |
| `effect` | The effect from the layer's series (the additive effect `a` for an orthogonal layer) |
| `d` | The per-locus dominance deviation of an orthogonal layer; `NA` otherwise |
| `var_explained` | The marker's marginal contribution to the trait variance |
| `QTN_t1`, `QTN_t2`, `ld_r2` | Always present; filled only under `architecture = "ld"` (the linked pair and its squared correlation) and `NA` otherwise |

Both functions take a `rep` argument. In the simulation above, the replications
share one genetic architecture, so `rep` makes no difference; only the residuals
differ. With `vary_qtn = TRUE` each replication samples its own causal markers
and effects, and `rep` selects which architecture to read:

[]{#output-vary-qtn}

```{r}
#| label: output-vary-qtn
vary <- simulate_phenotype(geno, n_reps = 3, vary_qtn = TRUE, seed = 7) |>
  additive(prop = 0.5, n_qtn = 4)

tibble(
  rep_1 = qtn_table(vary, rep = 1)$snp,
  rep_2 = qtn_table(vary, rep = 2)$snp,
  rep_3 = qtn_table(vary, rep = 3)$snp
)

tibble(
  id = rownames(genetic_values(vary, rep = 1)),
  rep_1 = genetic_values(vary, rep = 1)[, 1],
  rep_2 = genetic_values(vary, rep = 2)[, 1]
) |>
  head()
```

The three columns of marker names differ, and so do the genetic values of the
same individual across replications. The default `rep = 1` would silently
describe only the first architecture; reading another replication's phenotypes
against the first replication's QTN table would mismatch truth and data.

### The diagnostic plot

`plot()` draws up to four panels. With two or more traits the fourth panel is
the genetic-value scatter between the first two traits; with one trait it
compares the requested and realized genetic share instead.

[]{#output-plot}

```{r}
#| label: output-plot
#| fig-height: 6
#| fig-cap: "Diagnostic panels for the two-trait maize simulation: variance partition per trait, phenotype distribution of the first trait, effect sizes from the QTN table, and the genetic-value scatter between the two traits."
plot(sim)
```

| Panel | What it shows | What it does not show |
|---|---|---|
| `variance` | The requested proportion of each layer and the residual, stacked per trait | Realized sample variances or covariances between layers |
| `hist` | The realized phenotype distribution of the first trait, pooled over all replications | Other traits, or a single replication |
| `effects` | The `effect` column of `qtn_table()` for trait 1, replication 1 | Other traits or replications, and variance explained, which also depends on allele frequency |
| `cor` | Genetic values of trait 1 against trait 2 when `n_traits > 1`; otherwise realized versus requested genetic share | Phenotypic correlation, which includes residual correlation |

Select panels with `which`, for example `plot(sim, which = "hist")`, when a
report needs a single figure.

### Write phenotypes as text or JSON

`write_phenotypes()` writes the long table by default. The output directory
must already exist. The format, separator, and file type are arguments; the
examples below write three versions of the same phenotypes.

[]{#output-write}

```{r}
#| label: output-write
dir.create("results/export", showWarnings = FALSE, recursive = TRUE)

write_phenotypes(sim, file = "results/export/phenotypes.tsv")
write_phenotypes(sim, file = "results/export/phenotypes_wide.csv",
                 format = "wide", sep = ",")
write_phenotypes(sim, file = "results/export/phenotypes.json",
                 file_type = "json")

list.files("results/export")
```

Read them back to check the round trip. `read_tsv()` and `read_csv()` from
readr parse the text files; `jsonlite::read_json()` with
`simplifyVector = TRUE` turns the JSON array of row objects into a data frame.

[]{#output-roundtrip}

```{r}
#| label: output-roundtrip
from_tsv <- read_tsv("results/export/phenotypes.tsv", show_col_types = FALSE)
from_csv <- read_csv("results/export/phenotypes_wide.csv", show_col_types = FALSE)
from_json <- jsonlite::read_json("results/export/phenotypes.json",
                                 simplifyVector = TRUE)

tibble(
  file = c("phenotypes.tsv", "phenotypes_wide.csv", "phenotypes.json"),
  rows = c(nrow(from_tsv), nrow(from_csv), nrow(from_json)),
  max_abs_difference = c(
    max(abs(from_tsv$value - long$value)),
    max(abs(from_csv$Trait_1 - wide$Trait_1)),
    max(abs(from_json$value - long$value))
  )
)
```

The text files are written with up to 15 significant digits, so values read
back agree with the object to roughly machine precision rather than exactly.
JSON values are written with 17 significant digits and read back exactly. Both
statements are package definitions of the writers; the table shows the
consequence for this simulation.

### Write the QTN table

`write_qtn_table()` writes every column of `qtn_table()`. For a `vary_qtn`
simulation, `rep = "all"` stacks the replications with a leading `rep` column
so that the architectures stay distinguishable.

[]{#output-write-qtn}

```{r}
#| label: output-write-qtn
write_qtn_table(sim, file = "results/export/qtn_table.tsv")
write_qtn_table(vary, file = "results/export/qtn_table_all_reps.tsv",
                rep = "all")
write_qtn_table(sim, file = "results/export/qtn_table.json",
                file_type = "json")

read_tsv("results/export/qtn_table_all_reps.tsv", show_col_types = FALSE) |>
  select(rep, trait, layer, snp, effect, var_explained)

jsonlite::read_json("results/export/qtn_table.json", simplifyVector = TRUE) |>
  select(trait, layer, snp, effect) |>
  head(4)
```

### Split the markers into causal and non-causal files

A downstream analysis often needs the genotypes without the causal markers,
for example to study what a genome-wide association scan finds when the true
loci are absent, with the truth kept separately. `split_markers = TRUE` writes
the phenotype file together with three companions: the QTN table, the causal
markers, and every non-causal marker. Default companion names are derived from
`file`; `markers_files` overrides the two marker paths. The return value names
every path written.

[]{#output-split}

```{r}
#| label: output-split
files <- write_phenotypes(vary, file = "results/export/vary.tsv",
                          split_markers = TRUE, rep = "all")
files

list.files("results/export", pattern = "^vary")
```

The causal file uses the package's numeric layout, five metadata columns and
one column per individual, so it can be read back by `as_numeric()` or passed
to `simulate_phenotype()` again. With `rep = "all"` the causal set is the union
over replications: a marker causal in any selected replication is in the causal
file and never in the non-causal file.

[]{#output-split-read}

```{r}
#| label: output-split-read
causal <- read_tsv(files[["causal"]], show_col_types = FALSE)
noncausal <- read_tsv(files[["noncausal"]], show_col_types = FALSE)

tibble(
  file = c("causal", "noncausal", "input"),
  markers = c(nrow(causal), nrow(noncausal), nrow(geno)),
  columns = c(ncol(causal), ncol(noncausal), ncol(geno))
)

causal |>
  select(snp, chr, pos, cm, 6:8)
```

The two marker counts add up to the number of input markers, and the
causal markers are the union of the loci drawn in the three replications. The
JSON layout differs: one object with `individuals`, `markers`, and, in the
causal file, the `qtn_table` rows and a `causal_for` list on each marker that
names the layers using it.

[]{#output-split-json}

```{r}
#| label: output-split-json
json_files <- write_phenotypes(
  sim, file = "results/export/split.json", file_type = "json",
  split_markers = TRUE,
  markers_files = c(causal = "results/export/causal_markers.json",
                    noncausal = "results/export/noncausal_markers.json")
)

causal_json <- jsonlite::read_json(json_files[["causal"]],
                                   simplifyVector = TRUE)
names(causal_json)
length(causal_json$individuals)
causal_json$markers |>
  select(snp, chr, pos, maf) |>
  head(3)
causal_json$markers$causal_for[[1]]
```

Each marker object carries its genotypes as an array in `individuals` order,
and `causal_for` records that the first marker is used by both the additive and
the dominance layer of the first trait.

## Assumptions and theory

This chapter has no new statistical model. Its content is the set of package
definitions that fix how simulation results are represented:

- **Table layouts.** The long table is the canonical output, with columns
  `id`, `trait`, `rep`, and `value`. The wide table pivots `trait` into
  columns. Both come from the same realized values; neither is recomputed.
- **Genetic values.** The matrix from `genetic_values()` is the sum of the
  scaled additive, dominance, and epistatic layers, and the genetically
  mediated part of a genome-derived transcriptome layer. A `vqtl()` layer
  modulates the residual spread and contributes no genetic value, so it is
  excluded. The [complex architectures chapter](complex.qmd) returns to vQTLs.
- **Replications.** `n_reps` requests simulation replications. Without
  `vary_qtn`, replications share the architecture and differ in their
  residual draws; with `vary_qtn = TRUE` they are independent architectures.
  The `reps` argument of the entry-mean model is a different quantity, as the
  [conventions chapter](conventions.qmd) explains.
- **Causal set.** A marker is causal when any layer of any trait uses it in any
  selected replication. Gene rows of transcriptome layers are not markers and
  never appear in the marker files.

The `var_explained` column is a marginal quantity for each marker. Covariance
among marker contributions, which arises from linkage disequilibrium in
observed genotypes, means that the column need not sum to the layer's
proportion. The [additive models chapter](additive.qmd) examines this.

## Implementation and practical limits

Writers live in R; no compiled code is involved. Delimited text is written by
`data.table::fwrite()`, JSON by jsonlite. Before writing companion files, the
package checks that the directory exists and is writable, that no two paths
resolve to the same file, and then moves the set into place as a group, so a
failed write does not leave a partial set behind. These are package definitions
documented in `?write_phenotypes`.

- Output directories must exist; `write_phenotypes()` does not create them.
- Text output is not bit-for-bit; use JSON, or keep the R object, when exact
  values matter.
- Default companion names are derived from `file`; passing a short `file` name
  keeps the companion names short as well.
- The marker files use the genotypes of the simulated individuals. A phenotype
  built from expression alone has no markers, and `split_markers = TRUE` is an
  error in that case.
- `write_qtn_table()` writes one replication by default. For a `vary_qtn`
  simulation pass `rep = "all"` so the file describes every replication of the
  phenotype file.
- Specialized exporters of the frozen legacy interface, such as GEMMA `.fam`
  output, are not part of the grammar; the [legacy chapter](legacy.qmd)
  describes that interface.
