# Relationship matrices and prediction {#sec-prediction}

Suppose you have genotyped and phenotyped animals, and a new generation of
candidates with genotypes but no records. You want to predict the candidates'
breeding values from their relatives' records, compare a pedigree-based with a
genomic relationship matrix, and check how good the predictions are. Because
this is a simulation, the true breeding values are known, so the accuracy of
each predictor can be measured rather than assumed.

[]{#prediction-setup}

```{r}
#| label: prediction-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-prediction-example}

### Load the package and the mouse data

[]{#prediction-library}

```{r}
#| label: prediction-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

This chapter uses the heterogeneous-stock mouse genotypes distributed with the
BGLR package (1,814 mice and 10,346 SNPs with physical positions; see
`?BGLR::mice` for the provenance, which credits Valdar et al. 2006). The
[genotype-input chapter](input.qmd) shows how the genotypes are converted; here
the shared loader is used. Real phenotype records accompany the genotypes, and
the first rows show that they exist, but every trait below is simulated on the
real genotypes so that true breeding values are known.

[]{#prediction-data}

```{r}
#| label: prediction-data
source("_data.R")
mice <- book_mice()
records <- book_mice_records()

records |>
  select(SUBJECT.NAME, GENDER, cage, Litter, Obesity.BMI) |>
  head()
```

To keep the pedigree simulation fast, the example keeps four chromosomes and
200 founders. The mouse map has physical positions only, so a synthetic
genetic map is built from them; see the [founders chapter](founders.qmd) for
what `synthetic_map()` assumes.

[]{#prediction-founders}

```{r}
#| label: prediction-founders
mice <- mice[mice$chr %in% c("1", "2", "3", "4"), ]
mice$cm <- synthetic_map(mice$chr, mice$pos)
founder_ids <- colnames(mice)[-(1:5)][1:200]
founders <- as_population(mice, individuals = founder_ids)
founders
```

### The genomic relationship matrix

`g_matrix()` builds the additive genomic relationship matrix from marker
dosages. Its diagonal is about one plus the genomic inbreeding coefficient
relative to the base allele frequencies, so it averages near one in an outbred
sample in Hardy-Weinberg proportions and near two in inbred lines. Its
off-diagonal elements measure realized relatedness; centering on the sample
frequencies makes their average close to zero.

[]{#prediction-gmatrix}

```{r}
#| label: prediction-gmatrix
G <- g_matrix(founders)

tibble(
  quantity = c("Individuals", "Mean diagonal", "Mean off-diagonal",
               "Smallest off-diagonal", "Largest off-diagonal"),
  value = c(nrow(G), mean(diag(G)), mean(G[upper.tri(G)]),
            min(G[upper.tri(G)]), max(G[upper.tri(G)]))
)
```

The mean off-diagonal is close to zero because the matrix is centred on the
current sample's allele frequencies. Two options change the base. `ridge`
blends the matrix toward the identity when a positive-definite matrix is
required. `base_freq` fixes the allele frequencies used for centring, which
matters when the same base must be kept across generations.

[]{#prediction-gmatrix-options}

```{r}
#| label: prediction-gmatrix-options
G_ridge <- g_matrix(founders, ridge = 0.05)
p_base <- rowMeans(dosages(founders) + 1) / 2
G_fixed <- g_matrix(founders, base_freq = p_base)

tibble(
  matrix = c("Default", "ridge = 0.05", "base_freq = founder frequencies"),
  mean_diagonal = c(mean(diag(G)), mean(diag(G_ridge)), mean(diag(G_fixed))),
  smallest_eigenvalue = c(min(eigen(G, symmetric = TRUE)$values),
                          min(eigen(G_ridge, symmetric = TRUE)$values),
                          min(eigen(G_fixed, symmetric = TRUE)$values))
)
```

`dosages()` returns the marker-by-individual dosage matrix in `-1`, `0`, `1`
coding; adding one and halving gives the frequency of the counted allele. With
the founders' own frequencies, the fixed-base matrix equals the default one.
The ridge raises the smallest eigenvalue, which is what makes the matrix
positive definite.

### A two-generation pedigree

`a_matrix()` needs a recorded pedigree. `mating_design()` writes a mating plan
and `mate()` executes it through the meiosis simulator, recording each
progeny's parents. The plan below mates 100 mothers with 100 fathers at
random, then mates the first generation among itself.

[]{#prediction-pedigree}

```{r}
#| label: prediction-pedigree
plan1 <- mating_design(founders[1:100], founders[101:200],
                       design = "random", n_crosses = 50,
                       progeny_per_cross = 4, seed = 11)
gen1 <- mate(plan1, founders, seed = 12, prefix = "G1")

plan2 <- mating_design(gen1, design = "random", n_crosses = 50,
                       progeny_per_cross = 4, seed = 13)
gen2 <- mate(plan2, gen1, seed = 14, prefix = "G2")

everyone <- c(founders, gen1, gen2)
parentage(gen2) |>
  select(id, mother, father, generation) |>
  head()
```

`c()` pools the three generations into one population while keeping the
pedigree. `parentage()` lists each individual's recorded parents.

[]{#prediction-amatrix}

```{r}
#| label: prediction-amatrix
A <- a_matrix(everyone)
G_all <- g_matrix(everyone)

sibs <- parentage(gen2) |>
  group_by(mother, father) |>
  summarise(first = id[1], second = id[2], .groups = "drop")

tibble(
  relationship = c("Full sibs, pedigree A", "Full sibs, genomic G",
                   "Founder diagonal, A", "Founder diagonal, G"),
  mean = c(mean(A[cbind(sibs$first, sibs$second)]),
           mean(G_all[cbind(sibs$first, sibs$second)]),
           mean(diag(A)[founder_ids]), mean(diag(G_all)[founder_ids])),
  sd = c(sd(A[cbind(sibs$first, sibs$second)]),
         sd(G_all[cbind(sibs$first, sibs$second)]),
         sd(diag(A)[founder_ids]), sd(diag(G_all)[founder_ids]))
)
```

The pedigree gives full sibs of unrelated parents a relationship of one half;
the small spread in the first row arises because some first-generation mates
share a founder parent, which raises their progeny's relationship. The genomic
matrix measures the realized sharing of each pair, which varies much more
because of Mendelian sampling. The founders are unrelated and non-inbred in the
pedigree by definition, whereas the genomic diagonal spreads around its mean.

### Simulate a trait with known breeding values

`phenotype_value()` scores every individual on a fixed trait: chosen causal
loci, fixed additive effects, and a residual variance derived once from the
founders' additive variance at the requested heritability. The attribute
`genetic_value` holds each individual's additive value, which for a purely
additive trait is its breeding value on the trait's scale.

[]{#prediction-trait}

```{r}
#| label: prediction-trait
set.seed(21)
qtn <- sample(mice$snp, 100)
effects <- rnorm(100)

y <- phenotype_value(everyone, qtn, effects, h2 = 0.4, ref = founders, seed = 22)
true_bv <- additive_value(everyone, qtn, effects)

gen1_ids <- parentage(gen1)$id
gen2_ids <- parentage(gen2)$id

tibble(
  quantity = c("Residual variance parameter", "Additive variance, founders",
               "Additive variance, generation 2"),
  value = c(attr(y, "var_e"), var(true_bv[founder_ids]), var(true_bv[gen2_ids]))
)
```

`additive_value()` returns the same fixed additive values without a residual,
so it serves as the truth against which predictions are judged.

### Predict the candidates

The founders and first generation are the training set with records. The
second generation are candidates: they are in the population, so they are
predicted, but their records are withheld. The `ref` argument keeps the
reported variance components on the founder scale.

[]{#prediction-ebv}

```{r}
#| label: prediction-ebv
training <- y[c(founder_ids, gen1_ids)]

ebv_g <- predict_ebv(everyone, training, method = "gblup",
                     h2 = 0.4, ref = y[founder_ids])
ebv_a <- predict_ebv(everyone, training, method = "pedigree",
                     h2 = 0.4, ref = y[founder_ids])

tibble(
  id = head(gen2_ids),
  true_bv = true_bv[head(gen2_ids)],
  gblup = ebv_g[head(gen2_ids)],
  pedigree = ebv_a[head(gen2_ids)],
  reliability_gblup = attr(ebv_g, "reliability")[head(gen2_ids)]
)
```

The result is one predicted breeding value per individual with a reliability
attribute. The pedigree predictor gives full sibs identical values because
nothing in the pedigree distinguishes them. Individual predictions can be far
from the truth even when the predictor is good, so the predictor is judged on
all candidates at once. `prediction_accuracy()` correlates predictions with the known
breeding values and reports the slope of truth on prediction, which is one for
an unbiased predictor.

[]{#prediction-accuracy}

```{r}
#| label: prediction-accuracy
bind_rows(
  gblup = prediction_accuracy(ebv_g[gen2_ids], true_bv[gen2_ids]),
  pedigree = prediction_accuracy(ebv_a[gen2_ids], true_bv[gen2_ids]),
  .id = "method"
)
```

Both predictors are close to unbiased. The genomic predictor is more accurate
for these unphenotyped candidates because it uses each candidate's realized
marker sharing with the recorded individuals, whereas the pedigree predictor
can only assign full sibs the same prediction.

### Accuracy depends on heritability and training size

The same simulation makes it easy to vary the design. The table repeats the
genomic prediction for three heritabilities and three training sizes, drawing
new phenotypes for each heritability so that the residual variance matches.
Training records are drawn in one fixed random order from the founders and
the candidates' parents, so the larger sets contain the smaller ones and every
size mixes close and distant relatives of the candidates.

[]{#prediction-design}

```{r}
#| label: prediction-design
design <- expand.grid(h2 = c(0.2, 0.4, 0.8), n_train = c(100, 200, 400))
set.seed(29)
training_pool <- sample(c(founder_ids, gen1_ids))

design$accuracy <- mapply(function(h2, n_train) {
  y_h <- phenotype_value(everyone, qtn, effects, h2 = h2,
                         ref = founders, seed = 23)
  train <- y_h[training_pool[seq_len(n_train)]]
  ebv <- predict_ebv(everyone, train, method = "gblup",
                     h2 = h2, ref = y_h[founder_ids])
  prediction_accuracy(ebv[gen2_ids], true_bv[gen2_ids])[["accuracy"]]
}, design$h2, design$n_train)

design |>
  mutate(accuracy = round(accuracy, 3)) |>
  tidyr::pivot_wider(names_from = n_train, values_from = accuracy,
                     names_prefix = "n_train_")
```

Accuracy rises with heritability and with the number of records. Which
individuals are recorded matters as well: records on the candidates' parents
are worth more than the same number of records on distant founders, so a
training set's relatedness to the candidates can matter as much as its size. The exact
values depend on the random residuals and on the candidates, so a study would
replicate this design over seeds rather than report one run.

### Multi-trait prediction

Records on a correlated trait add information. The second trait below shares
the causal loci and has effects correlated with the first. Because the
simulation knows the additive values, the genetic covariance matrix of the two
traits in the founders can be computed directly, and the residual covariance is
set from the heritabilities with no residual correlation.

[]{#prediction-multitrait}

```{r}
#| label: prediction-multitrait
set.seed(24)
effects2 <- 0.7 * effects + rnorm(100, sd = 0.5)
y2 <- phenotype_value(everyone, qtn, effects2, h2 = 0.4,
                      ref = founders, seed = 25)
true_bv2 <- additive_value(everyone, qtn, effects2)

G0 <- cov(cbind(t1 = true_bv[founder_ids], t2 = true_bv2[founder_ids]))
R0 <- diag(c(attr(y, "var_e"), attr(y2, "var_e")))
dimnames(R0) <- dimnames(G0)
G0

Y <- cbind(t1 = y, t2 = y2)
Y[gen2_ids, "t1"] <- NA

ebv_mt <- predict_ebv(everyone, Y, method = "gblup", var_a = G0, var_e = R0)
ebv_mt_a <- predict_ebv(everyone, Y, method = "pedigree", var_a = G0, var_e = R0)

bind_rows(
  `gblup, single trait` = prediction_accuracy(ebv_g[gen2_ids], true_bv[gen2_ids]),
  `gblup, two traits` = prediction_accuracy(ebv_mt[gen2_ids, "t1"], true_bv[gen2_ids]),
  `pedigree, single trait` = prediction_accuracy(ebv_a[gen2_ids], true_bv[gen2_ids]),
  `pedigree, two traits` = prediction_accuracy(ebv_mt_a[gen2_ids, "t1"], true_bv[gen2_ids]),
  .id = "model"
)
```

The candidates have no record on the first trait but do have a record on the
second. The multi-trait model uses that record through the genetic covariance,
so accuracy for the first trait increases relative to the single-trait model.
The gain depends on the genetic correlation and on the second trait's
heritability.

### Supplying your own relationship matrix

A relationship matrix built elsewhere, or modified, can be passed through `K`.
Passing the matrix that `predict_ebv()` would have built reproduces the same
predictions, which confirms that `K` replaces the internal construction.

[]{#prediction-custom-k}

```{r}
#| label: prediction-custom-k
ebv_k <- predict_ebv(everyone, training, K = G_all, h2 = 0.4,
                     ref = y[founder_ids])
tibble(
  quantity = "Largest absolute difference from the gblup predictions",
  value = max(abs(ebv_k - ebv_g))
)
```

## Assumptions and theory

The predictions above use one trait model throughout: additive genetic values,
one record per individual, an overall mean as the only fixed effect, and known
variance components. There is no dominance, no repeated record, no
contemporary-group effect, no genotype-by-environment term, and no variance
component estimation.

### The mixed model being solved

Every predictor in this chapter solves the same known-variance mixed model.
Writing $y$ for the vector of records, $\mu$ for the mean, $Z$ for the matrix
that assigns each record to its individual, $u$ for the breeding values, and
$e$ for the residuals,

$$
y = \mathbf 1\mu + Zu + e, \qquad
u \sim N(0, K\sigma^2_A), \qquad
e \sim N(0, I\sigma^2_e).
$$ {#eq-blup-model}

$K$ is the relationship matrix: $A$ for the pedigree method and $G$ for the
genomic method. $\sigma^2_A$ is the additive variance and $\sigma^2_e$ the
residual variance. The package's roxygen documentation attributes this model
to Henderson (1975), and the equation map records that reference with its page
unverified. This is the model the package implements; it is not a claim that
the data must have arisen this way.

### Variance components from a heritability

With `h2`, the package sets the two variances from the phenotypic variance of
`ref` (by default the phenotyped subset): $\sigma^2_A = h^2\sigma^2_P$ and
$\sigma^2_e = (1 - h^2)\sigma^2_P$. The only quantity that enters the
predictions is their ratio $\lambda = \sigma^2_e / \sigma^2_A = (1 - h^2)/h^2$,
so the choice of `ref` changes the reported absolute variances but not the
predicted values. This is a package definition, implemented by
`.blup_variances()`.

### The generalized least-squares form

Rather than building the mixed-model equations with $K^{-1}$, the package
solves the equivalent form over the phenotyped rows $r$,

$$
V = K_{rr} + \lambda I, \qquad
\hat\mu = (\mathbf 1'V^{-1}\mathbf 1)^{-1}\mathbf 1'V^{-1}y, \qquad
\hat u = K_{\cdot r}V^{-1}(y - \mathbf 1\hat\mu).
$$ {#eq-blup-gls}

$V$ is the covariance of the records on the $\sigma^2_A$ scale. The first
expression is the generalized least-squares estimate of the mean. The second
regresses each individual's breeding value, phenotyped or not, on the adjusted
records through its covariances with them. No inverse of $K$ is needed, which
is why a singular genomic matrix is accepted. The equation map records that
this form equals the mixed-model solution numerically (package check). The
implementing function is `predict_ebv()`.

### Reliability

The reliability attribute is one minus the prediction error variance scaled by
the individual's prior variance,

$$
r^2_i = 1 - \frac{\mathrm{PEV}_i}{K_{ii}\sigma^2_A}, \qquad
\frac{\mathrm{PEV}_i}{\sigma^2_A} = K_{ii} - K_{ir}\,P\,K_{ri},
$$ {#eq-blup-reliability}

where $P$ is the projection $V^{-1} - V^{-1}\mathbf 1(\mathbf 1'V^{-1}\mathbf 1)^{-1}\mathbf 1'V^{-1}$,
which removes the estimated mean. This is the package definition in
`predict_ebv()`. It is a model-based quantity: it describes expected precision
under the assumed model and variances, not the realized accuracy measured
below.

### The pedigree relationship matrix

`a_matrix()` applies the tabular method. Processing individuals parents-first,
with $m(k)$ and $f(k)$ the recorded mother and father of individual $k$,

$$
A_{kj} = \tfrac12\left(A_{m(k),j} + A_{f(k),j}\right), \qquad
A_{kk} = 1 + F_k, \qquad F_k = \tfrac12 A_{m(k)f(k)}.
$$ {#eq-a-matrix}

An unknown parent contributes zero. Founders are unrelated and, with the
default `founder_f = 0`, non-inbred. The equation map attributes the method to
Emik & Terrill 1949 *J. Hered.* 40:51–55 (page unverified) and the coefficients
to Wright 1922 *Am. Nat.* 56:330–338 (page unverified). For inbred founders,
such as the maize lines used elsewhere in this book, `founder_f` must be set;
the mice here are an outbred stock, so the default was kept.

### The genomic relationship matrix

`g_matrix()` implements VanRaden's first method,

$$
G = \frac{ZZ'}{2\sum_j p_j(1-p_j)}, \qquad Z = M - 2p,
$$ {#eq-g-matrix}

where $M$ is the individuals-by-markers gene-content matrix with entries
0, 1, 2, and $p_j$ is the frequency of the counted allele at marker $j$.
Markers with $p_j$ equal to 0 or 1 are dropped. The optional ridge forms
$G^* = (1 - \text{ridge})\,G + \text{ridge}\,I$. The equation map records the
reference as VanRaden 2008 *J. Dairy Sci.* 91:4414–4423 (page unverified).
The denominator is the expected variance of the gene content summed over
markers under Hardy-Weinberg proportions at the frequencies $p$. The
diagonal is therefore about $1 + F_i$ relative to that base; it averaged close
to one above because the mice are an outbred stock, while inbred lines average
close to two.

### Accuracy and bias

`prediction_accuracy()` reports the correlation between predicted and true
breeding values over the individuals scored, and the slope of the regression
of truth on prediction,

$$
\text{accuracy} = \operatorname{cor}(\hat u, u), \qquad
\text{slope} = \frac{\operatorname{Cov}(u, \hat u)}{\operatorname{Var}(\hat u)}.
$$ {#eq-accuracy}

These are sample statistics of one simulation. The package documentation
notes that a slope of one is the property of best linear unbiased prediction
that the conditional mean of the truth given the prediction equals the
prediction, and that a slope below one indicates over-dispersed predictions.
This is a package definition; the equation map lists it as such.

## Implementation and practical limits

`predict_ebv()`, `a_matrix()`, `g_matrix()`, and `prediction_accuracy()` are
R functions in the selection module. The internal helpers named in the
[developer index](developer-index.qmd) include `.blup_variances()` for the
variance components, `.check_relationship()` for validating a supplied `K`, and
`.gblup_marker_effects()` for back-solving marker effects, which the
`marker_effects` attribute exposes when the genomic matrix was built internally
with `ridge = 0`.

- The variance components are supplied, not estimated. Accuracy obtained with
  the true simulated heritability is an upper reference for what an analysis
  with estimated components can achieve.
- `h2` with the default `ref` uses the phenotyped subset's variance. After
  selection that subset has a truncated variance, so pass the base population's
  records or `var_a` and `var_e` directly.
- Across cycles of genomic selection, keep `base_freq` at the base population's
  frequencies. A moving base re-expresses relatedness and inbreeding relative to
each new generation, which hides accumulated inbreeding and drops markers
fixed by selection; the package documentation also warns that it can inflate
apparent accuracy.
- The pedigree matrix treats founders as unrelated and non-inbred unless told
  otherwise. Real founders may be related, which the genomic matrix captures
  and the pedigree cannot.
- Missing phenotypes are refused rather than dropped in the single-trait model;
  the multi-trait model accepts `NA` for an unrecorded trait.
- Repeated records, contemporary groups, single-step models, and dominance
  relationship matrices are outside the scope of these functions.

::: {.callout-tip title="Where to go next"}
The predictions are the natural criterion for selection: pass them to
`select_ind()` through `on` as the [selection chapter](selection.qmd) shows,
or use them within the optimal-contribution methods of the
[contributions chapter](optimization.qmd).
:::
