# Combining ability and breeding schemes {#sec-programs}

A hybrid program wants to know which inbreds combine well in crosses; a
line-development program wants inbred lines from a segregating cross; a
population-improvement program wants gain cycle after cycle. This chapter
runs each of these on maize lines: a diallel scored for combining ability
both by expectation and by simulated progeny, a progeny test, three ways of
deriving inbred lines from one cross, and recurrent selection with its
response per cycle. Each simulated result is placed next to the quantity it
is supposed to estimate.

[]{#programs-setup}

```{r}
#| label: programs-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-programs-example}

### A trait with dominance

[]{#programs-library}

```{r}
#| label: programs-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

Combining ability is only interesting when crosses can differ from the
average of their parents, which needs dominance. The maize lines are inbred
and carry almost no heterozygotes, so we define the trait on an F2 of two
lines, where every segregating marker has heterozygotes, and freeze its
effects.

[]{#programs-trait}

```{r}
#| label: programs-trait
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(
  SNP55K_maize282_maf04,
  to_r = TRUE, to_file = FALSE, verbose = FALSE
)
pop <- as_population(geno, individuals = 1:30)
f1 <- cross(pop[1], pop[2], n = 1, seed = 1)
f2 <- selfcross(f1, n = 100, seed = 2)

sim <- simulate_phenotype(f2, h2 = 0.6, seed = 3) |>
  additive(prop = 0.4, n_qtn = 20) |>
  dominance(prop = 0.2)
template <- template_effects(sim)
head(template)
```

`template_effects()` returns each causal locus with its additive effect `a`
and dominance deviation `d` on the scale of the realized genetic values. The
functions below take these frozen effects and apply them to any population on
the same marker map, so values are comparable across populations. The check
confirms that the frozen effects reproduce the simulation's own genetic
values up to a constant.

[]{#programs-template-check}

```{r}
#| label: programs-template-check
frozen <- genotypic_value(f2, template$qtn, template$a, template$d)
realized <- genetic_values(sim)[, 1]
tibble(
  correlation = cor(frozen, realized[names(frozen)]),
  sd_of_difference = sd(frozen - realized[names(frozen)])
)
```

### A diallel among six inbreds

`combining_ability()` with `design = "diallel"` evaluates every unordered
pair of the candidates. With `method = "expected"` it computes the expected
progeny value of each cross from the parents' genotypes without simulating
anything; general combining ability (GCA) is each parent's average deviation
across its crosses and specific combining ability (SCA) is the remaining
cross-specific deviation.

[]{#programs-diallel}

```{r}
#| label: programs-diallel
lines <- pop[1:6]
expected <- combining_ability(
  lines, qtn = template$qtn, a = template$a, d = template$d,
  design = "diallel", method = "expected"
)
print(expected)
round(expected$gca, 3)
round(expected$sca, 3)
```

The SCA matrix is symmetric with an empty diagonal, because selfs are not
part of the design. GCAs sum to zero and each row of SCAs sums to zero; these
are properties of the decomposition, checked below.

[]{#programs-diallel-check}

```{r}
#| label: programs-diallel-check
tibble(
  quantity = c("Sum of GCAs", "Largest row sum of SCAs"),
  value = c(sum(expected$gca), max(abs(rowSums(expected$sca, na.rm = TRUE))))
)
```

`method = "simulated"` realizes progeny for every cross with the crossing
engine, scores them, and averages. With no residual and fully inbred parents,
every progeny of a cross is genetically identical, so the simulated and
expected results agree exactly. With a residual, each cross mean carries
environmental sampling error that shrinks with `n_progeny`.

[]{#programs-diallel-sim}

```{r}
#| label: programs-diallel-sim
simulated <- combining_ability(
  lines, qtn = template$qtn, a = template$a, d = template$d,
  design = "diallel", method = "simulated", n_progeny = 30, h2 = 0.5, seed = 4
)
print(simulated)

tibble(
  candidate = names(expected$gca),
  expected_gca = expected$gca,
  simulated_gca = simulated$gca[names(expected$gca)]
)
```

The parents' residual heterozygosity is tiny but not zero, so the
"identical progeny" statement holds only approximately for this panel; the
`h2 = 0.5` residual is the main source of the differences in the table.

### Testers and factorial designs

A topcross crosses every candidate to a common set of testers. GCA then
depends on the testers: the expected GCA of a candidate is proportional to
the average effect of its alleles **in the testers' allele frequencies**. A
factorial reports GCA for both sides.

[]{#programs-topcross}

```{r}
#| label: programs-topcross
testers <- pop[7:8]
topcross <- combining_ability(
  lines, testers, qtn = template$qtn, a = template$a, d = template$d,
  design = "topcross"
)
round(topcross$gca, 3)
round(topcross$sca, 3)

factorial <- combining_ability(
  lines, testers, qtn = template$qtn, a = template$a, d = template$d,
  design = "factorial"
)
round(factorial$gca_testers, 3)
```

With only two testers each candidate's two SCAs are mirror images, because
they must sum to zero.

### A progeny test

`progeny_test()` mates each parent to randomly drawn distinct mates and
reports the mean of its half-sib progeny. We test ten F2 plants against the
remaining ninety, with twenty progeny each and a residual such that the
heritability of the total genotypic value among the progeny is 0.3. The
progeny mean is compared with the parents' additive value on the frozen
effects.

[]{#programs-progeny}

```{r}
#| label: programs-progeny
test <- progeny_test(
  f2[1:10], f2[11:100],
  qtn = template$qtn, a = template$a, d = template$d,
  n_progeny = 20, h2 = 0.3, seed = 5
)
test

parent_additive <- additive_value(f2[1:10], template$qtn, template$a)
tibble(
  correlation = cor(test$progeny_mean, parent_additive[test$id]),
  own_performance_accuracy = sqrt(0.3),
  progeny_test_accuracy = sqrt(20 * 0.3 / (4 + 19 * 0.3))
)
```

The two reference accuracies are the classical values for an additive trait
described in the theory section; the observed correlation over ten parents
is one noisy draw, and the trait includes dominance, so the comparison is
illustrative rather than a test of the formula.

### Three ways to derive inbred lines

Starting from the same 100 F2 plants, single seed descent selfs each plant
one seed per generation without selection; bulk advance pools the selfed
progeny each generation and draws the next generation at random, losing line
identity; pedigree selection phenotypes every generation, keeps the best
fraction, and selfs those. The pedigree scheme needs a phenotyping rule,
written as a function of the current population.

[]{#programs-schemes}

```{r}
#| label: programs-schemes
phenotype <- function(p) {
  simulate_phenotype(p, h2 = 0.5, seed = 7) |> additive(n_qtn = 50)
}

ssd <- single_seed_descent(f2, generations = 5, seed = 9)
bulked <- bulk(f2, generations = 5, n = 100, seed = 9)
selected <- pedigree(
  f2, phenotype, generations = 5, prop = 0.2, pop_size = 100, seed = 9
)
attr(selected, "history")
```

The pedigree history records the number selected, the selection differential
and the standardized intensity in each generation, measured on the phenotype
each generation was ranked on. To compare the three schemes on one scale we
freeze the effects of the trait the pedigree scheme selected on and score
every final population with them.

[]{#programs-schemes-compare}

```{r}
#| label: programs-schemes-compare
frozen_trait <- template_effects(phenotype(f2))
summarize_lines <- function(p, scheme) {
  value <- additive_value(p, frozen_trait$qtn, frozen_trait$a)
  tibble(
    scheme = scheme, lines = n_individuals(p),
    heterozygosity = mean(dosages(p) == 0),
    mean_value = mean(value), sd_value = sd(value)
  )
}
bind_rows(
  summarize_lines(f2, "F2 (start)"),
  summarize_lines(ssd, "Single seed descent"),
  summarize_lines(bulked, "Bulk"),
  summarize_lines(selected, "Pedigree selection")
)
```

Five selfing generations drive the fraction of heterozygous marker calls
from its F2 value toward zero in all three schemes. Single seed descent
keeps the F2's mean, apart from sampling, and increases the genetic spread
among lines, as inbreeding converts within-line heterozygosity into
between-line variance. The bulk's mean drifts because seeds are drawn at
random each generation. Pedigree selection raises the mean and narrows the
spread, because it repeatedly keeps the top fifth. The
population returned by the pedigree scheme remembers the trait it was
selected on:

[]{#programs-population-trait}

```{r}
#| label: programs-population-trait
population_trait(selected)
```

### Pooling populations

Breeding programs combine material from several crosses. `c()` pools
`Population` objects that share a marker map and keeps identifiers unique.
Here lines from two single seed descent families are pooled.

[]{#programs-pool}

```{r}
#| label: programs-pool
other_f1 <- cross(pop[3], pop[4], n = 1, seed = 1)
other_f2 <- selfcross(other_f1, n = 50, seed = 2)
other_ssd <- single_seed_descent(other_f2, generations = 5, seed = 9)

pooled <- c(ssd[1:10], other_ssd[1:10])
print(pooled)
parentage(pooled) |>
  select(id, mother, father, generation) |>
  head(3)
```

`parentage()` reads the recorded pedigree, so the origin of each pooled line
remains traceable; see [Meiosis and crossing](cross.qmd).

### Recurrent selection

`recurrent_selection()` phenotypes the population, selects the best parents,
intercrosses them at random, and repeats. We run three cycles on thirty
founders, selecting ten parents per cycle with ten progeny per cross, and
follow the mean additive value on the frozen scale of the founders' trait.

[]{#programs-recurrent}

```{r}
#| label: programs-recurrent
founders <- pop
founder_trait <- template_effects(phenotype(founders))
cycle_mean <- function(p) mean(additive_value(p, founder_trait$qtn, founder_trait$a))

improved <- recurrent_selection(
  founders, phenotype, cycles = 3, n_parents = 10, progeny_per_cross = 10, seed = 8
)
attr(improved, "history")
tibble(
  population = c("Founders", "After 3 cycles"),
  mean_additive_value = c(cycle_mean(founders), cycle_mean(improved))
)
```

To see the response per cycle, run one cycle at a time, feeding each
result back in. The founders' trait is carried by the selected individuals
and their progeny, so later cycles are phenotyped on the same genetic scale
with the same residual variance.

[]{#programs-recurrent-cycles}

```{r}
#| label: programs-recurrent-cycles
current <- founders
response <- tibble(cycle = 0, mean_additive_value = cycle_mean(current))
for (k in 1:4) {
  current <- recurrent_selection(
    current, phenotype, cycles = 1, n_parents = 10, progeny_per_cross = 10,
    seed = 100 + k
  )
  response <- bind_rows(
    response, tibble(cycle = k, mean_additive_value = cycle_mean(current))
  )
}
response
```

The response per cycle is not constant: the genetic variance among the
candidates changes as selection and drift act, and the heritability on the
original residual scale falls with it.

## Assumptions and theory

### Expected cross means

For one locus with genotypic values $-a$, $d$, $+a$ at gene content 0, 1, 2,
a parent with gene content $x$ transmits the counted allele with probability
$g = x/2$. The expected progeny value of the cross between parents $i$ and
$k$ is

$$
E[G_{ik}] = g_i g_k\, a + \left[g_i(1-g_k) + (1-g_i) g_k\right] d - (1-g_i)(1-g_k)\, a,
$$ {#eq-expected-cross}

summed over loci (`.expected_cross_means()`). The equation map records this
as a package derivation from the genotypic values of Falconer & Mackay 1996
(page unverified). The expectation depends only on gamete frequencies, so
linkage does not enter it; linkage enters the variance of a realized family.

### GCA and SCA

For a topcross or factorial with cross means $Y_{ik}$ and grand mean $\mu$,

$$
g_i = \bar Y_{i\cdot} - \mu, \qquad h_k = \bar Y_{\cdot k} - \mu, \qquad
s_{ik} = Y_{ik} - \mu - g_i - h_k,
$$ {#eq-gca-factorial}

the two-way model the equation map attributes to Sprague & Tatum 1942
*J. Am. Soc. Agron.* 34:923–932 (page unverified). For a diallel of $p$
candidates without reciprocals or selfs, each candidate appears in $p-1$
crosses, and the least-squares GCA is

$$
g_i = (m_i - \mu)\,\frac{p-1}{p-2}, \qquad s_{ij} = Y_{ij} - \mu - g_i - g_j,
$$ {#eq-gca-diallel}

with $m_i$ the mean of candidate $i$'s crosses. The equation map records
this as algebraically identical to Griffing 1956 *Aust. J. Biol. Sci.*
9:463–493, method 4 (page unverified); the implementation is
`.decompose_ca()`. With $d = 0$ every expected SCA is exactly zero; with
dominance, the GCA of a candidate toward a set of testers is proportional to
the tester-referenced average effect $\alpha_T = a + d(1 - 2p_T)$, with
$p_T$ the testers' gamete frequency (Falconer & Mackay 1996, page
unverified).

### Progeny-test accuracy

Under the classical half-sib assumptions, the mates a large random-mating
population, each progeny by an independent mate, the package derives for an
additive trait the accuracy of a progeny test on $n$ records as

$$
r = \sqrt{\frac{n h^2}{4 + (n-1) h^2}},
$$ {#eq-progeny-accuracy}

which exceeds own-performance accuracy $\sqrt{h^2}$ once
$n > (4 - h^2)/(1 - h^2)$. The equation map lists this as a package
derivation of a classical result, with no page asserted. The worked example
violates the assumptions in two ways, dominance in the trait and a mate pool
of only ninety related F2 plants, which is why it is presented as an
illustration.

### What the schemes do

The schemes are package definitions implemented in `single_seed_descent()`,
`bulk()`, `pedigree()` and `recurrent_selection()`, following the
descriptions in Bernardo 2020 *Breeding for Quantitative Traits in Plants*
(cited by the roxygen documentation, no page given). Bulk advance draws the
next generation uniformly from the pooled selfed progeny, so each plant's
number of advanced seeds is multinomial; line identity is not tracked.
Pedigree selection and recurrent selection use `select_ind()` for the
selection step and record its differential and intensity per generation.
From the second generation on, the phenotype callback reuses the trait the
population carries: genetic values stay on the base scale and the residual
variance is fixed, so heritability changes as the genetic variance changes.
Pass `refit = TRUE` inside the callback to re-standardize every generation
instead.

## Implementation and practical limits

Expected combining ability is deterministic linear algebra on gamete
frequencies. The simulated method, the progeny test and all schemes draw
their meioses through the crossing engine, with every random quantity drawn
in R and passed to the Rust recombination kernel.

- `combining_ability()` and `progeny_test()` take frozen `qtn`, `a`, `d`
  effects; `template_effects()` refuses models with an epistasis layer or an
  expression-mediated layer carrying variance, because those have no
  per-locus `a`/`d` form.
- Only a diallel without reciprocals is modelled (Griffing's method 4);
  reciprocal and maternal effects are outside the model, and at least three
  candidates are required.
- A candidate appearing twice, for example after `c(pop, pop)`, is an error:
  it would be one cross scored twice.
- `h2` in the simulated method and in `progeny_test()` is the heritability of
  the **total** genotypic value in the reference population, a broad-sense
  value when `d` is not zero.
- `pedigree()` and `recurrent_selection()` require the callback to build its
  simulation from the population it is handed; a closure over the base
  population is detected and refused.
- `recurrent_selection()` draws crosses with replacement among the selected
  parents, so with few crosses some selected parents may not be used.
- Pooling populations that carry different traits drops the trait from the
  pool with a warning; the next phenotype is then fitted afresh.
