# Individual, family, and multivariate selection {#sec-selection}

Suppose you have a segregating population organized in families, with a
phenotype for each individual, and you want to choose the parents of the next
generation. This chapter selects on individual records, on family information,
and on several traits at once, and checks the classical prediction of response
against what the simulation actually delivers. The simulation knows every
individual's genetic value, so the genetic consequence of each selection
decision can be measured directly.

[]{#selection-setup}

```{r}
#| label: selection-setup
#| include: false
source("_setup.R")
```

## Worked example {#sec-selection-example}

### Load the package and build families

[]{#selection-library}

```{r}
#| label: selection-library
library(simplePHENOTYPES)
library(dplyr)
library(tibble)
```

The example starts from the bundled maize inbred lines, restricted to three
chromosomes to keep the meiosis simulation fast. Ten lines are crossed to ten
others in a factorial design, giving 100 F1 hybrids, and each F1 is selfed to
produce ten S1 progeny. The result is 1,000 individuals in 100 families, each
family being the selfed progeny of one F1. The [crossing chapter](cross.qmd)
explains the meiosis model; here only the resulting population matters.

[]{#selection-population}

```{r}
#| label: selection-population
data("SNP55K_maize282_maf04", package = "simplePHENOTYPES")
geno <- as_numeric(SNP55K_maize282_maf04, to_r = TRUE, to_file = FALSE,
                   verbose = FALSE)
geno <- geno[geno$chr %in% c("1", "2", "3"), ]

parents <- as_population(geno, individuals = 1:20)
plan_f1 <- mating_design(parents[1:10], parents[11:20], design = "factorial")
f1 <- mate(plan_f1, parents, seed = 1, prefix = "F1")

plan_s1 <- data.frame(mother = parentage(f1)$id, father = parentage(f1)$id, n = 10)
s1 <- mate(plan_s1, f1, seed = 2, prefix = "S1")
family <- families(s1, by = "selfed")

s1
table(family) |> head()
```

`mating_design()` writes the mother-father plan and `mate()` executes it. A
plan row with the same individual as mother and father is a self. `families()`
labels each S1 individual by the F1 it descends from, which is the grouping
the family methods need.

### Simulate the phenotype

A single additive trait with 50 causal loci and a genetic share of 0.3 is
simulated on the S1 population. The [additive chapter](additive.qmd) explains
the model; the selection functions only require a realized simulation object.

[]{#selection-trait}

```{r}
#| label: selection-trait
sim <- simulate_phenotype(s1, h2 = 0.3, n_qtn = 50, seed = 3)
sim

pheno <- phenotypes_wide(sim)
gv <- genetic_values(sim)[, "Trait_1"]
```

`phenotypes_wide()` gives one row per individual, and `genetic_values()`
returns the simulated genetic values that selection cannot see but the book
can use to judge each method.

### Mass selection and the selection intensity

`select_ind()` ranks individuals on a criterion and keeps the best. The
criterion `on = "pheno"` is the observed phenotype, which is what a breeder
without markers can select on. Exactly one of `n`, `prop`, or `intensity`
sets how many are kept.

[]{#selection-mass}

```{r}
#| label: selection-mass
top <- select_ind(sim, prop = 0.1, on = "pheno", method = "mass")
top

tibble(
  attribute = c("Individuals kept", "Selection differential",
                "Realized intensity", "Method"),
  value = c(n_individuals(top), attr(top, "differential"),
            attr(top, "intensity"), attr(top, "method"))
)
```

The result is a `Population` of the selected individuals, ready to be crossed,
carrying the selection differential $S$ (mean criterion of the selected minus
mean of all candidates) and the realized intensity $i = S / s_P$, where $s_P$
is the sample standard deviation of the criterion. The table compares the
realized intensity for several proportions with the large-population value
$i(p) = \phi(z_p)/p$ used when `intensity` is given instead of `prop`.

[]{#selection-intensity}

```{r}
#| label: selection-intensity
proportions <- c(0.5, 0.2, 0.1, 0.05, 0.02)

intensity <- tibble(
  prop = proportions,
  kept = sapply(proportions, function(p) n_individuals(select_ind(sim, prop = p))),
  realized_i = sapply(proportions, function(p) attr(select_ind(sim, prop = p), "intensity")),
  expected_i = dnorm(qnorm(1 - proportions)) / proportions
)
intensity

by_intensity <- select_ind(sim, intensity = 1.75)
n_individuals(by_intensity)
```

The realized intensities lie close to the large-population values. With
`intensity = 1.75`, the function keeps the count whose large-population
intensity is nearest to the request.

### The breeder's equation

For mass selection on a purely additive trait, the classical prediction is
that the genetic gain equals the heritability times the selection
differential, $R = h^2 S$. The simulation lets us compare that prediction with
the genetic selection differential actually achieved: the mean genetic value
of the selected individuals minus the mean of all candidates.

[]{#selection-response}

```{r}
#| label: selection-response
genetic_gain <- function(selected) {
  mean(gv[attr(selected, "selected")]) - mean(gv)
}

response <- tibble(
  prop = proportions,
  S = sapply(proportions, function(p) attr(select_ind(sim, prop = p), "differential")),
  predicted = 0.3 * S,
  simulated = sapply(proportions, function(p) genetic_gain(select_ind(sim, prop = p)))
)
response
```

The simulated gains track the prediction, with scatter from the particular
residuals drawn. The prediction assumes that the regression of genetic value
on phenotype is linear with slope $h^2$; the theory section below says when
that holds.

Selection on the true genetic value (`on = "gv"`) is the idealized upper
bound. For this additive model the breeding value (`on = "bv"`) is the same
quantity. A `direction = "low"` selection keeps the smallest values instead.

[]{#selection-criteria}

```{r}
#| label: selection-criteria
tibble(
  criterion = c("pheno", "gv", "bv", "pheno, direction = low"),
  genetic_gain = c(
    genetic_gain(select_ind(sim, prop = 0.1, on = "pheno")),
    genetic_gain(select_ind(sim, prop = 0.1, on = "gv")),
    genetic_gain(select_ind(sim, prop = 0.1, on = "bv")),
    genetic_gain(select_ind(sim, prop = 0.1, on = "pheno", direction = "low"))
  )
)
```

Any externally computed score, such as the genomic predictions of the
[prediction chapter](prediction.qmd), can be passed to `on` as a named numeric
vector or as a function of the simulation object.

### The random control

`method = "random"` draws individuals at random. Its reported differential and
intensity are the realized values on the criterion for the individuals drawn,
so they scatter around zero. It is the drift control against which a selection
method's gain is judged.

[]{#selection-random}

```{r}
#| label: selection-random
controls <- sapply(1:5, function(k) {
  set.seed(k)
  genetic_gain(select_ind(sim, prop = 0.1, method = "random"))
})
tibble(replicate = 1:5, genetic_gain = controls)
```

### Response in the next generation

A selected `Population` remembers the trait it was selected on, so progeny
can be phenotyped on the same scale. The selected individuals are intermated
at random, as are the individuals of a random control, and the progeny are
scored with the carried trait. `population_trait()` shows what is carried.

[]{#selection-progeny}

```{r}
#| label: selection-progeny
population_trait(top)

control <- select_ind(sim, prop = 0.1, method = "random")

progeny_of <- function(selected, seed) {
  plan <- mating_design(selected, design = "random", n_crosses = 50,
                        progeny_per_cross = 10, seed = seed)
  mate(plan, selected, seed = seed + 1)
}

next_sel <- simulate_phenotype(progeny_of(top, 10), seed = 12)
next_ctrl <- simulate_phenotype(progeny_of(control, 20), seed = 22)

tibble(
  population = c("All candidates", "Selected parents", "Random-control parents",
                 "Progeny of selected", "Progeny of random control"),
  mean_genetic_value = c(mean(gv), mean(gv[attr(top, "selected")]),
                         mean(gv[attr(control, "selected")]),
                         mean(genetic_values(next_sel)[, "Trait_1"]),
                         mean(genetic_values(next_ctrl)[, "Trait_1"])),
  mean_phenotype = c(mean(pheno$Trait_1), NA, NA,
                     mean(phenotypes_wide(next_sel)$Trait_1),
                     mean(phenotypes_wide(next_ctrl)$Trait_1))
)
```

The progeny of the selected parents have a higher mean than the progeny of
the random control, and each progeny mean is close to its parents' mean
genetic value, as expected for an additive trait; the remaining gap reflects
which parents the random plan happened to use and how often. Because the carried trait keeps the base population's loci, effects,
and residual variance, these means are comparable with the candidates' means.
The progeny phenotype means equal the genetic means because the default fixed
residual mode centres the residuals; see the [first chapter](first-simulation.qmd).
The [trait chapter](trait.qmd) explains the carried trait in detail.

### Family selection

When individuals are grouped in families, their family mean carries
information about their breeding values. Three methods use it.
`within_family` keeps the best individuals inside each family,
`among_family` keeps whole families with the best means, and `combined`
ranks on an index that weights the individual's own deviation and its family's
deviation. The combined index needs the trait's heritability and the
correlation of breeding values among family members, which for S1 sibs of a
non-inbred F1 is two thirds.

[]{#selection-family}

```{r}
#| label: selection-family
within <- select_ind(sim, n = 100, method = "within_family", family = family)
among <- select_ind(sim, n = 100, method = "among_family", family = family)
combined <- select_ind(sim, n = 100, method = "combined", family = family,
                       h2 = 0.3, family_relationship = 2 / 3)
mass <- select_ind(sim, n = 100, method = "mass")

tibble(
  method = c("mass", "within_family", "among_family", "combined"),
  kept = c(n_individuals(mass), n_individuals(within),
           n_individuals(among), n_individuals(combined)),
  families_represented = c(
    length(unique(family[attr(mass, "selected")])),
    length(unique(family[attr(within, "selected")])),
    length(unique(family[attr(among, "selected")])),
    length(unique(family[attr(combined, "selected")]))
  ),
  genetic_gain = c(genetic_gain(mass), genetic_gain(within),
                   genetic_gain(among), genetic_gain(combined))
)
```

Within-family selection keeps every family represented, which conserves
diversity at the cost of gain. Among-family selection keeps whole families, so
it returns at least `n` individuals and concentrates the selection in few
families. The combined index uses both sources of information. Which method
gains most depends on the heritability and the family structure; at this
moderate heritability the methods that use the family mean are competitive
with mass selection.

`n_per_family` keeps a stated number from every family instead of allocating
a total.

[]{#selection-per-family}

```{r}
#| label: selection-per-family
one_each <- select_ind(sim, method = "within_family", family = family,
                       n_per_family = 1)
n_individuals(one_each)
head(attr(one_each, "n_per_family"))
```

### Selection on several traits

Multi-trait selection needs a multi-trait simulation. Two pleiotropic traits
with a requested genetic correlation of 0.3 are simulated on the same
population; the [multi-trait chapter](multitrait.qmd) explains the
architecture. The index weights use the simulation's true breeding-value
covariance; the Smith-Hazel index then ranks phenotypes, while the quadratic
index ranks the true breeding values.

[]{#selection-multitrait}

```{r}
#| label: selection-multitrait
sim2 <- simulate_phenotype(s1, architecture = "pleiotropy", n_traits = 2,
                           h2 = c(0.4, 0.4), n_qtn = 50,
                           cor = matrix(c(1, 0.3, 0.3, 1), 2), seed = 4)
gv2 <- genetic_values(sim2)

gain2 <- function(selected) {
  colMeans(gv2[attr(selected, "selected"), , drop = FALSE]) - colMeans(gv2)
}

index <- select_ind(sim2, n = 100, method = "index", weights = c(1, 1))
index_t1 <- select_ind(sim2, n = 100, method = "index", weights = c(1, 0))
linear_bv <- select_ind(sim2, n = 100, method = "quadratic_index",
                        weights = c(1, 1))
quadratic <- select_ind(sim2, n = 100, method = "quadratic_index",
                        weights = c(1, 1),
                        quad_weights = matrix(c(-1, 0, 0, 0), 2))

spread_t1 <- function(selected) sd(gv2[attr(selected, "selected"), "Trait_1"])

bind_rows(
  `index, equal weights` = gain2(index),
  `index, trait 1 only` = gain2(index_t1),
  `quadratic index, linear part only` = gain2(linear_bv),
  `quadratic index, penalty on trait 1` = gain2(quadratic),
  .id = "method"
) |>
  mutate(sd_trait1_selected = c(spread_t1(index), spread_t1(index_t1),
                                spread_t1(linear_bv), spread_t1(quadratic)))
```

`method = "index"` is the Smith-Hazel index: economic weights, one per trait,
are converted to index weights through the phenotypic and genetic covariance
matrices, and individuals are ranked on the weighted phenotypes. Putting all
weight on the first trait still moves the second because the traits are
genetically correlated. `method = "quadratic_index"` scores the true breeding
values directly, as the package documentation states, so its gains are an
idealized reference and are not comparable with the phenotype-based index.
Its quadratic term, here a penalty on the square of the first trait, pulls the
selected set toward intermediate values of that trait: the gain on trait 1
drops and so does the spread of the selected individuals, which is the form
for traits with an intermediate optimum.

Independent culling levels keep an individual only if it passes a threshold on
every trait. `sequential = TRUE` culls the traits in order, each among the
survivors of the previous ones.

[]{#selection-culling}

```{r}
#| label: selection-culling
culled <- select_ind(sim2, method = "culling", culling = c(0.4, 0.4))
culled_seq <- select_ind(sim2, method = "culling", culling = c(0.4, 0.4),
                         sequential = TRUE)

bind_rows(
  `simultaneous culling` = c(kept = n_individuals(culled), gain2(culled)),
  `sequential culling` = c(kept = n_individuals(culled_seq), gain2(culled_seq)),
  .id = "method"
)
```

With culling the number kept follows from the proportions, so `n` is not
given. Simultaneous culling at 0.4 on each of two traits keeps fewer than
40 percent, because an individual must pass both thresholds. Sequential
culling at the same proportions keeps exactly 0.4 of 0.4.

### Selection with a relatedness penalty

`method = "bqp"` chooses exactly `n` individuals that maximize the summed
standardized merit minus a penalty on the relatedness among those chosen. With
`lambda = 0` the penalty vanishes and the method reduces to truncation on the
merit. The relatedness is measured on the method's own genomic relationship
matrix, so the comparison below summarizes each selected set by its mean
genomic relationship from `g_matrix()` instead, which is an independent
measure described in the [prediction chapter](prediction.qmd).

[]{#selection-bqp}

```{r}
#| label: selection-bqp
mass20 <- select_ind(sim2, n = 20, method = "mass")
bqp0 <- select_ind(sim2, n = 20, method = "bqp", lambda = 0)
bqp1 <- select_ind(sim2, n = 20, method = "bqp", lambda = 1)
bqp5 <- select_ind(sim2, n = 20, method = "bqp", lambda = 5)

G_s1 <- g_matrix(s1)
mean_relationship <- function(selected) {
  ids <- attr(selected, "selected")
  block <- G_s1[ids, ids]
  mean(block[upper.tri(block)])
}

bind_rows(
  `mass, trait 1` = gain2(mass20),
  `bqp, lambda = 0` = gain2(bqp0),
  `bqp, lambda = 1` = gain2(bqp1),
  `bqp, lambda = 5` = gain2(bqp5),
  .id = "method"
) |>
  mutate(
    same_set_as_mass = c(TRUE,
                         setequal(attr(bqp0, "selected"), attr(mass20, "selected")),
                         setequal(attr(bqp1, "selected"), attr(mass20, "selected")),
                         setequal(attr(bqp5, "selected"), attr(mass20, "selected"))),
    mean_relationship = c(mean_relationship(mass20), mean_relationship(bqp0),
                          mean_relationship(bqp1), mean_relationship(bqp5)),
    solver = c(NA, attr(bqp0, "bqp")$solver, attr(bqp1, "bqp")$solver,
               attr(bqp5, "bqp")$solver)
  )
```

Without `weights`, the method uses the single criterion given by `on` and
`trait`, here the phenotype of the first trait, so the unpenalized solution
coincides with mass selection on that trait. Increasing the penalty lowers
the mean relationship among the selected individuals and, with it, the gain;
the penalty weight is a choice about that trade-off. With 1,000 candidates the
exact enumeration is infeasible, so a greedy construction followed by local
search is used; the reported solver says which, and it carries no optimality
guarantee.

### The method manifest

`selection_methods()` lists every selection and scoring operator the package
exports, with the function that implements it. Front ends build their
interfaces from this table.

[]{#selection-manifest}

```{r}
#| label: selection-manifest
selection_methods() |>
  select(id, fn)
```

## Assumptions and theory

The worked example uses an additive trait, a single record per individual,
families of one type, and candidates from one generation. There is no
dominance, no shared family environment, and no genotype-by-environment
interaction.

### Differential, intensity, and the count kept

With `prop`, the number kept is $k = \max(1, \operatorname{round}(pN))$ for
$N$ candidates, a package definition in `.resolve_keep()`. The selection
differential and realized intensity reported as attributes are

$$
S = \bar y_{\text{sel}} - \bar y, \qquad i = \frac{S}{s_y},
$$ {#eq-selection-differential}

with $s_y$ the sample standard deviation of the criterion over all candidates.
With `intensity`, the count is the one whose large-population intensity
$i(p) = \phi(z_p)/p$ is nearest to the request, where $z_p$ is the standard
normal quantile cutting off the upper fraction $p$ and $\phi$ is the standard
normal density. The equation map attributes these definitions to Falconer &
Mackay 1996 (page unverified). The large-population formula overstates the
expected realized intensity in small samples.

### Response to mass selection

The package documentation states the classical response $R = i\,h^2\sigma_P =
h^2 S$ for a purely additive trait, and the more general
$R = i\,\mathrm{Cov}(A, P)/\sigma_P$, the regression of breeding value $A$
on the criterion $P$. The equation map attributes the response to Falconer &
Mackay 1996 (page unverified) and lists the general index form under Smith
1936 *Ann. Eugen.* 7:240–250 and Hazel 1943 *Genetics* 28:476–490 (page
unverified). Two conditions are needed for the simple form. The regression of
$A$ on $P$ must be linear, which holds under joint normality but not for an
arbitrary joint distribution. And $\mathrm{Cov}(A, P)$ must equal the additive
variance, which requires every non-additive component of the phenotype to be
uncorrelated with the breeding value. With an additive model and independent
residuals, as in this chapter, both hold up to sampling. With dominance or
epistasis under linkage disequilibrium, or in inbred and selected populations,
the covariance need not equal the additive variance; the package documentation
for `select_ind()` gives counterexamples.

The chapter's check compared the genetic selection differential of the
selected parents with $h^2 S$. The progeny mean then follows from the parents'
breeding values under random mating of the selected set, which the progeny
table illustrated.

### The combined index

`method = "combined"` scores each individual on the Lush index of its own
deviation and its family's deviation from the overall mean. With $h^2$ the
heritability, $r$ the correlation of breeding values among family members,
$n$ the family size, and $t = r h^2$ the phenotypic intraclass correlation
implied by an additive model, the weights are

$$
b_1 = \frac{h^2(1 - r)}{1 - r h^2}, \qquad
b_2 = \frac{h^2\, n\, r\,(1 - h^2)}{[1 + (n - 1) r h^2](1 - r h^2)}, \qquad
I = b_1 (y - \bar y) + b_2 (\bar y_f - \bar y),
$$ {#eq-combined-index}

the solution $b = V^{-1}c$ of the selection-index equations. The equation map
records the reference Lush 1947 81:241–261, 362–379 (verified). The weights are
optimal under an additive model in which family members covary only through
breeding values; dominance, epistasis, or a shared environment add to the
covariance of full-sib or selfed families, so the index is then reasonable but
not optimal. The value $r = 2/3$ for S1 sibs of a non-inbred parent is the
package's derivation from the tabular relationship matrix; the equation map
notes that this attribution has no primary page.

### The Smith-Hazel index and culling

`method = "index"` computes $b = P^{-1} G a$ and ranks on $I = b'y$, with $P$
the sample phenotypic covariance matrix, $G$ the sample covariance matrix of
the true breeding values, and $a$ the economic weights. The equation map cites
Smith 1936 7:240–250 and Hazel 1943 28:476–490. When $P$ is singular a
minimum-norm solution is used, a package extension without a genetic source.
Hazel's $G$ is the covariance of phenotypes with breeding values; the two
coincide only when breeding values and dominance deviations are uncorrelated,
which holds under random mating but not in general in selfed or selected
populations.

`method = "culling"` is independent culling levels, attributed in the equation
map to Hazel & Lush 1942 33(11):393–399 (verified). Under their idealized
conditions of equal, uncorrelated traits, the package derivation gives
aggregate responses proportional to $i(p)\sqrt{T}$ for the index,
$T\, i(p^{1/T})$ for simultaneous culling, and $i(p)$ for tandem selection,
so index selection is at least as efficient as culling, which is at least as
efficient as tandem selection.

### The relatedness-penalized set

`method = "bqp"` maximizes
$\sum_i\sum_j w_j s_{ij} x_i - \lambda \sum_i\sum_k G_{ik} x_i x_k$ over
indicator variables $x_i \in \{0, 1\}$ with exactly $n$ ones, where $s_{ij}$
is the standardized criterion of trait $j$, $w_j$ its weight, and $G$ the
paper's genomic relationship matrix built from the column-standardized marker
matrix. The package documentation attributes the method to Montesinos-Lopez et
al. (2025) and solves it by exact enumeration for small problems and by a
greedy start with local search otherwise, which carries no optimality
guarantee. The quadratic index of `method = "quadratic_index"` follows
Ceron-Rojas et al. (2026) as cited in the package documentation; the equation
map marks its article number unverified.

## Implementation and practical limits

`select_ind()` is an R function. The internal helpers named in the
[developer index](developer-index.qmd) include `.resolve_keep()` for the
count, `.combined_score()` for the Lush index, `.index_weights()` for the
Smith-Hazel weights, `.quadratic_index_score()`, `.select_culling()`, and the
breeding-value construction `.breeding_value_matrix()` that the `"bv"`
criterion and the index methods use.

- The Smith-Hazel index builds its weights from the simulation's true
  breeding-value covariance and ranks phenotypes; the quadratic index ranks
  the true breeding values themselves. Neither uses estimates. To select on predictions, compute the index from
  the predicted values and pass it through `on` with `method = "mass"`.
- `"bv"` is refused for models with an epistasis layer, because an epistatic
  term has no per-locus average effect to reconstruct.
- `among_family` returns at least `n` individuals, and the reported intensity
  describes the whole returned set.
- The combined index uses one heritability and one family relationship for all
  families, so it assumes families of one type from comparable parents.
- The random control's differential is a realized quantity and can take either
  sign; replicate it over seeds before reading anything into one draw.
- Breeding scheme wrappers such as `recurrent_selection()` apply mass
  selection in each cycle; see the [breeding schemes chapter](programs.qmd).

::: {.callout-tip title="Carrying the selection forward"}
Every result of `select_ind()` on a Population-backed simulation is itself a
`Population` with the trait attached. Crossing it, as the progeny example did,
is the first step of a breeding cycle; the [contributions chapter](optimization.qmd)
shows how to choose parental contributions rather than a fixed number of
parents.
:::
