---
title: "Simulating breeding populations"
author: "simplePHENOTYPES"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulating breeding populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
```

```{r setup}
library(simplePHENOTYPES)
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04
```

simplePHENOTYPES can simulate meiosis directly, which means you can build a
pedigree from real founder genotypes: cross two lines, self the offspring,
produce doubled haploids, and then simulate phenotypes on the result.

These functions implement the algorithms from the `isqg` package, reimplemented
in Rust. Recombination follows the Karlin & Liberman count-location model: the
number of crossovers on a chromosome is Poisson with mean equal to its length in
Morgans, crossover positions are uniform along it, and chromosomes assort
independently.

There are four functions to learn:

| Function | What it does |
|---|---|
| `as_population()` | turns genotype data into phased founders |
| `cross()` | mates two individuals |
| `selfcross()` | self-pollinates one individual |
| `double_haploid()` | makes fully homozygous progeny in one step |

# Before you start: you need a genetic map

Meiosis is driven by *genetic* distance, in centiMorgans, not physical distance
in base pairs. Two markers 1 Mb apart may recombine freely or almost never,
depending on where they sit relative to the centromere.

The genetic map lives in the `cm` column of your genotype data:

```{r map-check}
aggregate(cm ~ chr, data = geno, FUN = max)
```

If yours is empty, `as_population()` will stop and tell you so. Build one with
`synthetic_map()`:

```{r map-build, eval = FALSE}
geno$cm <- synthetic_map(geno$chr, geno$pos)
```

The map bundled with this data set is synthetic — a model, not a measured maize
linkage map. See `vignette("genetic-maps")` for how it is built, how to tune it,
and how to substitute a real map.

# Founders

`as_population()` converts genotype data into a `Population`: the phased
representation meiosis needs. Pick your founders with `individuals`.

```{r founders}
pop <- as_population(geno, individuals = c("33-16", "38-11"))
pop
```

A dosage matrix says an individual is heterozygous but not *which* homologue
carries which allele, so heterozygotes are phased arbitrarily. For an inbred
panel like this one that is nearly lossless, because heterozygotes are rare. For
outbred material it is a real assumption — see `?as_population`.

## Working with a Population

```{r pop-basics}
n_individuals(pop)
pop[1]                       # subset individuals
dim(dosages(pop))            # markers x individuals, coded -1/0/1
dosages(pop)[1:4, ]
```

`dosages()` collapses the phased haplotypes back to the familiar -1 / 0 / 1
coding: `-1` and `1` are the two homozygotes, `0` is a heterozygote.

# Your first cross

`cross()` mates two individuals. Each parent must be a *single* individual, so
subset with `[` if your population holds more.

```{r f1}
f1 <- cross(pop[1], pop[2], n = 1, seed = 11)
f1
```

The markers worth watching are the ones that can tell the founders apart: both
founders homozygous, and for *different* alleles. At those the F1 must be
heterozygous, because it received one homologue from each parent.

```{r f1-het}
d_pop <- dosages(pop)
informative <- which(d_pop[, 1] != d_pop[, 2] & d_pop[, 1] != 0 & d_pop[, 2] != 0)
length(informative)

all(dosages(f1)[informative, 1] == 0)
```

Requiring both founders to be homozygous matters. A locus where one founder is
heterozygous also "differs" between them, but the F1 there is heterozygous only
half the time, and including such loci quietly blurs every ratio that follows.

Ask for more progeny with `n`. Each is an independent draw, so full sibs differ:

```{r f1-many}
sibs <- cross(pop[1], pop[2], n = 5, seed = 12)
sibs
```

# Selfing

`selfcross()` pollinates an individual with itself. Both gametes come from
independent meioses of the same plant, so an F2 segregates:

```{r f2}
f2 <- selfcross(f1, n = 200, seed = 13)
f2
```

At the informative markers the classic 1:2:1 ratio appears:

```{r f2-segregation}
round(prop.table(table(dosages(f2)[informative, ])), 3)
```

# Doubled haploids

`double_haploid()` takes a single gamete and doubles it, producing a completely
homozygous individual in one generation. This is the fast route to a fixed line:

```{r dh}
dh <- double_haploid(f1, n = 200, seed = 14)

any(dosages(dh) == 0)                                    # no heterozygotes at all
round(prop.table(table(dosages(dh)[informative, ])), 3)  # 1:1
```

# Backcrossing

A backcross is just a cross back to one of the parents — there is no separate
function:

```{r backcross}
bc1 <- cross(f1, pop[1], n = 100, seed = 15)
```

Every BC1 gets one gamete from the recurrent parent, so at an informative locus
it is either homozygous for that parent or still heterozygous — never homozygous
for the donor. Counting how many of the two alleles came from the recurrent
parent:

```{r backcross-alleles}
recurrent <- dosages(pop[1])[informative, 1]
d_bc <- dosages(bc1)[informative, ]

n_recurrent <- ifelse(d_bc == recurrent, 2, ifelse(d_bc == 0, 1, 0))
round(prop.table(table(n_recurrent)), 3)

mean(n_recurrent) / 2   # fraction of the genome from the recurrent parent
```

Half the loci are fixed for the recurrent parent and half are heterozygous, which
works out to 75% of alleles — the familiar backcross arithmetic (50% from its own
gamete, plus half of the F1's 50%).

# Recurrent selfing: from an F1 to inbred lines

Repeated selfing halves heterozygosity each generation. Single-seed descent is a
loop — advance each line by selfing one plant per generation:

```{r ssd}
het_of <- function(x) mean(dosages(x)[informative, 1] == 0)

set.seed(7)
lines <- vector("list", 20)                              # 20 independent F2 lines
for (i in seq_along(lines)) lines[[i]] <- selfcross(f1, n = 1)

het <- numeric(6)
het_all <- function(pop_list) {
  out <- numeric(length(pop_list))
  for (i in seq_along(pop_list)) out[i] <- het_of(pop_list[[i]])
  out
}

het[1] <- mean(het_all(lines))
for (g in 2:6) {
  for (i in seq_along(lines)) lines[[i]] <- selfcross(lines[[i]], n = 1)
  het[g] <- mean(het_all(lines))
}

data.frame(
  generation = paste0("F", 2:7),
  heterozygosity = round(het, 3),
  expected = round(0.5^(1:6), 3)
)
```

Averaged across lines, heterozygosity halves each generation, as theory says it
should. After six generations the lines are essentially fixed — the same endpoint
`double_haploid()` reaches in one step, which is why doubled haploids are
attractive when the technology exists for your crop.

Note the averaging. Any *single* line is a poor estimate of the expectation,
because heterozygosity is not scattered locus by locus — it survives in long
chromosomal blocks, so a genome carries far fewer independent pieces of
information than it has markers:

```{r ssd-spread}
round(range(het_all(lines)), 3)
```

At F7 individual lines range from essentially fixed to several percent
heterozygous. That spread is real biology, not simulation noise, and it is why
breeding programmes carry many lines rather than one.

# Checking the simulation is doing what you think

Segregation ratios confirm the alleles are being transmitted correctly, but they
say nothing about *linkage*. To check recombination you have to ask which parent
each locus came from.

Doubled haploids make this easy: they are homozygous, so their allele names the
founder it came from directly.

```{r linkage}
on_chr1 <- geno$chr == 1
p1 <- dosages(pop[1])[on_chr1, 1]
p2 <- dosages(pop[2])[on_chr1, 1]

# Informative loci: both founders homozygous, for different alleles.
inf1 <- which(p1 != p2 & p1 != 0 & p2 != 0)
cm1 <- geno$cm[on_chr1][inf1]

from_p1 <- dosages(dh)[on_chr1, ][inf1, ] == p1[inf1]

pick <- c(5, 20, 100, 400)
data.frame(
  distance_cM   = round(cm1[pick] - cm1[1], 1),
  recombination = round(rowMeans(from_p1[pick, ] != from_p1[rep(1, 4), ]), 3),
  haldane       = round((1 - exp(-2 * (cm1[pick] - cm1[1]) / 100)) / 2, 3)
)
```

The observed recombination fractions track Haldane's map function,
$(1 - e^{-2d})/2$ at distance $d$ Morgans. Nearby loci travel together, distant
ones approach one half — and never exceed it, because a recombination fraction
above 0.5 is impossible.

This check is worth remembering: it is sensitive to bugs that segregation ratios
cannot see. An earlier version of this package reconstructed progeny phase from
dosages, which silently produced haplotypes the founders never had. F2 and DH
ratios still looked perfect; the recombination fractions came out at 0.99, which
is what exposed it.

# Reproducibility

All randomness is drawn by R, so a pedigree is exactly reproducible. Either pass
`seed`, or set one beforehand:

```{r seeds}
a <- selfcross(f1, n = 5, seed = 99)
b <- selfcross(f1, n = 5, seed = 99)
identical(dosages(a), dosages(b))

set.seed(99)
c1 <- selfcross(f1, n = 5)
set.seed(99)
c2 <- selfcross(f1, n = 5)
identical(dosages(c1), dosages(c2))
```

# Phenotyping a pedigree

A `Population` is accepted anywhere `simulate_phenotype()` takes genotype data,
so you can go straight from pedigree to phenotype:

```{r phenotype}
ph <- simulate_phenotype(f2, seed = 21) |>
  additive(prop = 0.6, n_qtn = 5)
ph

head(phenotypes_long(ph))
```

This is the natural way to simulate a mapping population: build the pedigree,
then lay a genetic architecture over it. See `vignette("simplePHENOTYPES-v2")`
for the full grammar.

# A complete example

Putting it together — a biparental doubled-haploid mapping population with a
simulated trait:

```{r complete}
founders <- as_population(geno, individuals = c("33-16", "38-11"))
f1_line  <- cross(founders[1], founders[2], n = 1, seed = 1)
dh_pop   <- double_haploid(f1_line, n = 150, seed = 2)

trait <- simulate_phenotype(dh_pop, seed = 3) |>
  additive(prop = 0.5, n_qtn = 10)

n_individuals(dh_pop)
head(phenotypes_wide(trait))
```

# Common problems

**"`mother` must contain exactly one individual"** — the mating functions take
one individual at a time. Subset with `[`: `cross(pop[1], pop[2])`, not
`cross(pop, pop)`.

**"The genetic map (`cm`) is all NA"** — meiosis needs recombination distances.
Build a map with `synthetic_map()` first.

**"`cm` must be non-decreasing within each chromosome"** — sort your markers by
chromosome and position before converting.

**Progeny look unexpectedly similar to one parent** — check how many markers are
actually informative. Two founders that share an allele at a locus produce
progeny that are uninformative there no matter how meiosis went.

# Session information

```{r session}
sessionInfo()
```
