---
title: "Complete reference: every option in one place"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete reference: every option in one place}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette exercises **every option added in version 2**, end to end, in one
runnable document. It is deliberately exhaustive rather than gentle: if you want
a guided introduction, start with `vignette("simplePHENOTYPES-v2")`, then
`vignette("breeding-populations")` and `vignette("genetic-maps")`.

Everything below actually runs when this vignette is built, so the printed
output is real, not illustrative.

```{r library}
library(simplePHENOTYPES)
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04
dim(geno)
```

The first five columns are marker metadata (`snp`, `allele`, `chr`, `pos`,
`cm`); the rest are individuals.

```{r peek}
geno[1:5, 1:8]
```

# 1. Genotype input

## Converting from other formats

`as_numeric()` reads HapMap, VCF, GDS and PLINK bed/ped and returns the numeric
format used throughout.

```{r as-numeric}
hmp <- data.frame(
  `rs#`   = c("snp1", "snp2", "snp3"),
  alleles = c("C/T", "A/G", "G/T"),
  chrom   = c(1, 1, 2),
  pos     = c(100, 500, 200),
  strand  = "+", `assembly#` = NA, center = NA,
  protLSID = NA, assayLSID = NA, panelLSID = NA, QCcode = NA,
  L1 = c("C", "A", "G"), L2 = c("T", "G", "T"),
  L3 = c("C", "A", "T"), L4 = c("T", "G", "G"),
  check.names = FALSE, stringsAsFactors = FALSE
)

as_numeric(hmp, to_r = TRUE, verbose = FALSE)
```

Coding schemes and allele orientation:

```{r as-numeric-options}
# 0 / 1 / 2 instead of -1 / 0 / 1
as_numeric(hmp, to_r = TRUE, code_as = "012", verbose = FALSE)[, 1:8]

# Orient against a supplied reference allele rather than by frequency
as_numeric(hmp, to_r = TRUE, verbose = FALSE,
           method = "reference", ref_allele = c("C", "A", "G"))[, 1:8]
```

Reading from files (not run here, since it needs your data):

```{r read-files, eval = FALSE}
as_numeric("my_genotypes.hmp.txt", to_r = TRUE)
as_numeric("my_genotypes.vcf",     to_r = TRUE)
as_numeric("my_genotypes.bed",     to_r = TRUE)   # needs .bim and .fam alongside
```

## Filtering markers with `filter_geno()`

`filter_geno()` is one place to trim a genotype before simulating: it drops
monomorphic markers, filters by minor-allele frequency and by heterozygosity,
and prunes by linkage disequilibrium. A marker must pass every requested filter.

```{r filter-basic, message = FALSE}
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04

# MAF window + only markers that carry heterozygotes (e.g. before dominance)
g <- filter_geno(geno, maf_below = 0.48, hets = "include", verbose = FALSE)
nrow(g)
```

Heterozygosity matters for dominance: a dominance deviation acts on
heterozygotes, so on a near-inbred panel `dominance()` errors unless the loci
have heterozygotes. Pre-filtering fixes that:

```{r filter-dominance}
simulate_phenotype(g, seed = 60) |>
  additive(prop = 0.4, n_qtn = 5) |>
  dominance(prop = 0.1)
```

**LD pruning and haplotype blocks.** The three sliding-window pruners take a
`c(window, step, threshold)` specification; `blocks` groups markers into
Gabriel *et al.* (2002) haplotype blocks and keeps one tag per block. (Shown on a
single chromosome to keep the vignette quick.)

```{r filter-ld, message = FALSE}
chr1 <- geno[geno$chr == geno$chr[1], ][1:400, ]

# pairwise composite-r2 pruning (r2 > 0.2 in 50-marker windows)
nrow(filter_geno(chr1, indep_pairwise = c(50, 5, 0.2), verbose = FALSE))

# phased (EM haplotype) r2 pruning
nrow(filter_geno(chr1, indep_pairphase = c(50, 5, 0.2), verbose = FALSE))

# variance-inflation-factor pruning (PLINK's --indep)
nrow(filter_geno(chr1, indep = c(50, 5, 2), verbose = FALSE))

# Gabriel haplotype blocks: keep one tag marker per block
nrow(filter_geno(chr1, blocks = TRUE, block_max_kb = 2000, verbose = FALSE))
```

Using the VIF pruner or the block definition prints a one-per-session citation
for the method it implements (PLINK's VIF procedure, or Gabriel *et al.* for the
blocks); the pairwise pruners are generic r^2 operations and print nothing.

# 2. The genetic map

Meiosis needs distances in centiMorgans. `SNP55K_maize282_maf04` ships with a
**synthetic** map in `cm` — modelled from the physical positions, *not* a
published maize linkage map, and not to be quoted as measured recombination
distance.

```{r map-shipped}
chrom_length <- aggregate(cm ~ chr, data = geno, FUN = max)
chrom_length

round(sum(chrom_length$cm))             # total genetic length, cM
```

`synthetic_map()` builds such a map for your own data. Every argument:

```{r map-options}
# Default: pericentromeric suppression, length from cm_per_mb
cm_default <- synthetic_map(geno$chr, geno$pos)

# Uniform map: cM strictly proportional to bp
cm_uniform <- synthetic_map(geno$chr, geno$pos, suppression = 0)

# Fixed total length per chromosome
cm_fixed <- synthetic_map(geno$chr, geno$pos,
                          total_cm = c(200, 180, 175, 170, 160,
                                       130, 130, 125, 115, 110))

# Explicit centromere positions, and a wider/stronger suppressed region
chr1 <- geno$pos[geno$chr == 1]
mid <- aggregate(pos ~ chr, data = geno, FUN = median)
cm_custom <- synthetic_map(geno$chr, geno$pos,
                           centromere = mid$pos,
                           suppression = 0.95, width = 0.25)

round(c(default = max(cm_default[geno$chr == 1]),
        uniform = max(cm_uniform[geno$chr == 1]),
        fixed   = max(cm_fixed[geno$chr == 1]),
        custom  = max(cm_custom[geno$chr == 1])), 1)
```

The suppression model concentrates recombination on the chromosome arms:

```{r map-landscape}
along_chr1 <- data.frame(
  pos = chr1,
  cm  = cm_default[geno$chr == 1],
  bin = cut(chr1, breaks = 8, labels = FALSE)
)

span <- aggregate(cbind(cm, pos) ~ bin, data = along_chr1,
                  FUN = function(x) max(x) - min(x))
span$cM_per_Mb <- round(span$cm / (span$pos / 1e6), 2)

span[, c("bin", "cM_per_Mb")]     # low near the centromere, high on the arms
```

To use your own real map, just assign it:

```{r map-real, eval = FALSE}
geno$cm <- my_real_map_in_cM
```

# 3. The simulation grammar

A simulation starts with `simulate_phenotype()` and gains variance components as
layers. Each `prop` is a proportion of total phenotypic variance, so
heritability is the sum of the genetic proportions.

```{r grammar-basic}
ph <- simulate_phenotype(geno, seed = 1) |>
  additive(prop = 0.5, n_qtn = 3)
ph
```

## Every layer

```{r grammar-layers}
full <- simulate_phenotype(geno, seed = 2) |>
  additive(prop = 0.3, n_qtn = 5) |>
  dominance(prop = 0.1) |>
  epistasis(prop = 0.1, n_pairs = 2, interaction = 2) |>
  vqtl(prop = 0.1)
full
```

## Layer options

```{r grammar-layer-options}
# Geometric effects with an explicit base (0.5, 0.25, 0.125, ...)
simulate_phenotype(geno, seed = 3) |>
  additive(prop = 0.5, n_qtn = 3, effect = 0.5)

# Explicit effect series, one per QTN
simulate_phenotype(geno, seed = 3) |>
  additive(prop = 0.5, n_qtn = 3, effect = c(0.6, 0.3, 0.1))

# Dominance and vQTL on their own loci rather than the additive ones
simulate_phenotype(geno, seed = 4) |>
  additive(prop = 0.3, n_qtn = 5) |>
  dominance(prop = 0.1, same_as_add = FALSE, n_qtn = 3) |>
  vqtl(prop = 0.1, same_as_add = FALSE, n_qtn = 2)

# Three-way epistasis
simulate_phenotype(geno, seed = 5) |>
  additive(prop = 0.3, n_qtn = 4) |>
  epistasis(prop = 0.2, n_pairs = 2, interaction = 3)
```

## Replications and per-trait proportions

```{r grammar-reps}
reps <- simulate_phenotype(geno, n_traits = 2, n_reps = 3, seed = 6) |>
  additive(prop = c(0.6, 0.3), n_qtn = 4)     # different h2 per trait
table(phenotypes_long(reps)$rep)
head(phenotypes_wide(reps))
```

## One call instead of a pipe

```{r grammar-one-call}
simulate_phenotype(geno, h2 = 0.5, n_qtn = 3, seed = 7)              # model "A"
simulate_phenotype(geno, h2 = 0.6, n_qtn = 4, model = "AD", seed = 7)
```

## Fine control: fixed loci, phase, means, subsets

Set the causal loci for one layer while the rest are drawn at random. Give
marker names or column indices:

```{r fine-qtn}
simulate_phenotype(geno, h2 = 0.5, seed = 1) |>
  additive(qtn = c("ss196442916", "ss196439337", "ss196480535"))
```

`phase = "repulsion"` alternates the effect signs so linked increasing and
decreasing alleles oppose one another (the realized variance is still `prop`;
only the sign structure changes):

```{r fine-phase}
ph_rep <- simulate_phenotype(geno, h2 = 0.5, seed = 1) |>
  additive(n_qtn = 6, phase = "repulsion")
qtn_table(ph_rep)$effect
```

A per-trait intercept with `mean`, and a subset of individuals with
`individuals`:

```{r fine-mean}
ph_m <- simulate_phenotype(geno, n_traits = 2, h2 = 0.5, mean = c(10, 20),
                           seed = 1) |>
  additive(n_qtn = 5)
colMeans(phenotypes_wide(ph_m)[, c("Trait_1", "Trait_2")])
```

```{r fine-subset}
ph_s <- simulate_phenotype(geno, h2 = 0.5,
                           individuals = c("33-16", "38-11", "A188", "B73"),
                           seed = 1) |>
  additive(n_qtn = 3)
n_individuals_sim <- ph_s$n_ind
n_individuals_sim
```

`vary_qtn = TRUE` gives each replication its own QTNs, and `distinct_chr = TRUE`
puts each trait's QTNs on separate chromosomes:

```{r fine-vary}
ph_v <- simulate_phenotype(geno, h2 = 0.5, n_reps = 3, vary_qtn = TRUE,
                           seed = 1) |>
  additive(n_qtn = 4)
# a different QTN set per replication
lengths(ph_v$layers[[1]]$qtn_reps)

ph_d <- simulate_phenotype(geno, n_traits = 2, h2 = 0.4, distinct_chr = TRUE,
                           seed = 1) |>
  additive(n_qtn = 3)
lapply(ph_d$layers[[1]]$qtn, function(i) unique(geno$chr[i]))
```

# 4. Genetic architectures

## Independent

Trait-specific QTNs, no enforced sharing.

```{r arch-independent}
ind <- simulate_phenotype(geno, n_traits = 2, h2 = 0.4, seed = 8) |>
  additive(n_qtn = 5)

qtn <- ind$layers[[1]]$qtn
head(qtn[[1]], 3)      # trait 1
head(qtn[[2]], 3)      # trait 2: different loci
```

## Pleiotropy with a target genetic correlation

Shared QTNs whose effects are drawn from a multivariate normal, so the genetic
correlation is the target in expectation. Works for **any number of traits**.

```{r arch-pleiotropy}
ph2 <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 2,
                          cor = 0.6, seed = 9) |>
  additive(prop = 0.5, n_qtn = 200)

g <- genetic_values(ph2)
round(cor(g[, 1], g[, 2]), 3)
```

Any number of traits, with a full target matrix including negative
correlations:

```{r arch-pleiotropy-matrix}
R <- matrix(c( 1.0,  0.8, -0.4,
               0.8,  1.0, -0.2,
              -0.4, -0.2,  1.0), 3, 3)

# Average over 20 seeds, since any single simulation scatters around the target
total <- matrix(0, 3, 3)
for (s in 1:20) {
  sim <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 3,
                            cor = R, h2 = 0.5, seed = s) |>
    additive(n_qtn = 300)
  total <- total + cor(genetic_values(sim))
}

round(total / 20, 2)      # compare with R above
```

Partial pleiotropy — only some of each trait's genetic variance is shared —
via `pi`:

```{r arch-pleiotropy-pi}
part <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 3,
                           cor = 0.5, pi = c(1, 0.8, 0.6), seed = 10) |>
  additive(prop = 0.5, n_qtn = 100)

# Every trait keeps n_qtn loci; a shared core carries the covariance
qtn_pi <- part$layers[[1]]$qtn
lengths(qtn_pi)

shared <- intersect(intersect(qtn_pi[[1]], qtn_pi[[2]]), qtn_pi[[3]])
length(shared)
```

Concentrating variance in major pleiotropic loci. This is available but costs
precision in the realized correlation, so both arguments default to zero:

```{r arch-pleiotropy-major}
simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 2,
                   cor = 0.6, n_pleio_major = 5, prop_var_major = 0.5,
                   seed = 11) |>
  additive(prop = 0.5, n_qtn = 100)
```

Impossible requests are an error rather than a silent approximation:

```{r arch-pleiotropy-error, error = TRUE}
simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 2,
                   cor = 0.9, pi = 0.5, seed = 12) |>
  additive(prop = 0.5, n_qtn = 20)
```

## Linkage disequilibrium

The causal variant sits near, but not on, the marker a study would test.

```{r arch-ld}
simulate_phenotype(geno, architecture = "ld", n_traits = 2,
                   ld_type = "indirect", seed = 13) |>
  additive(prop = 0.4, n_qtn = 3)

simulate_phenotype(geno, architecture = "ld", n_traits = 2,
                   ld_type = "direct", r2_max = 0.9, r2_min = 0.4, seed = 14) |>
  additive(prop = 0.4, n_qtn = 3)
```

## Combining architectures

```{r arch-complex}
pleio <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 2,
                            cor = 0.5, seed = 15) |>
  additive(prop = 0.4, n_qtn = 5)
indep <- simulate_phenotype(geno, n_traits = 2, seed = 15) |>
  additive(prop = 0.3, n_qtn = 5)

complex_phenotypes(pleio, indep, h2 = 0.5)
```

# 5. Breeding populations

Meiosis is simulated directly, so a pedigree can be built from real founders and
phenotyped without leaving the package.

```{r pop-founders}
pop <- as_population(geno, individuals = c("33-16", "38-11", "A188", "B73"))
pop
n_individuals(pop)
dim(dosages(pop))
```

## Crossing, selfing, doubled haploids

```{r pop-crosses}
f1  <- cross(pop[1], pop[2], n = 1, seed = 21)
f2  <- selfcross(f1, n = 200, seed = 22)
dh  <- double_haploid(f1, n = 200, seed = 23)
bc1 <- cross(f1, pop[1], n = 100, seed = 24)      # backcross to founder 1

c(F1 = n_individuals(f1), F2 = n_individuals(f2),
  DH = n_individuals(dh), BC1 = n_individuals(bc1))
```

Checking the genetics came out right:

```{r pop-checks}
d <- dosages(pop)
inf <- which(d[, 1] != d[, 2] & d[, 1] != 0 & d[, 2] != 0)  # informative loci

# F1 is heterozygous everywhere the founders differ
all(dosages(f1)[inf, 1] == 0)

# F2 segregates 1:2:1
round(prop.table(table(dosages(f2)[inf, ])), 3)

# Doubled haploids are fully homozygous, ~1:1
any(dosages(dh) == 0)
round(prop.table(table(dosages(dh)[inf, ])), 3)

# Backcross to founder 1: half heterozygous, half like the recurrent parent
round(prop.table(table(dosages(bc1)[inf, ])), 3)
```

## Recurrent selfing

Single-seed descent: each generation one plant is selfed, and the next
generation descends from one of its progeny.

```{r pop-recurrent}
gen <- f1
het <- numeric(5)
for (i in 1:5) {
  gen <- selfcross(gen[1], n = 100, seed = 30 + i)
  het[i] <- mean(dosages(gen)[inf, ] == 0)
}
round(het, 3)     # heterozygosity halves each generation
```

## Reproducibility

```{r pop-repro}
identical(dosages(selfcross(f1, n = 10, seed = 99)),
          dosages(selfcross(f1, n = 10, seed = 99)))
```

## Phenotyping a pedigree

A `Population` is accepted anywhere `simulate_phenotype()` takes genotypes.

```{r pop-phenotype}
simulate_phenotype(f2, seed = 40) |>
  additive(prop = 0.6, n_qtn = 5)
```

# 6. Inspecting a simulation

Nothing is written to disk unless you ask for it. Everything the original
`create_phenotypes()` wrote to `Log_Sim.txt` and its companion files lives in
the returned object instead.

`print()` shows the requested variance budget alongside what was actually
realized:

```{r inspect-print}
# dominance needs heterozygous loci; on this near-inbred panel filter to them
# first (otherwise dominance() errors -- see filter_geno() above)
gh <- filter_geno(geno, hets = "include", verbose = FALSE)
ph_i <- simulate_phenotype(gh, h2 = 0.5, seed = 60) |>
  additive(prop = 0.3, n_qtn = 5) |>
  dominance(prop = 0.2)
ph_i
```

`genetic_values()` returns the genetic component, before the residual — the
equivalent of the old `Genetic_values.txt`:

```{r inspect-gv}
gv <- genetic_values(ph_i)
head(gv)

# Realized heritability, computed by hand from the same pieces
round(var(gv[, 1]) / var(phenotypes_wide(ph_i)$Trait_1), 3)
```

The `var_explained` column of `qtn_table()` gives each additive/dominance QTN's
marginal share of phenotypic variance (`NA` for epistatic sets, where variance
is not per-locus). `qtn_table()` names the causal loci and their effects — the equivalent of
`Additive_QTNs.txt` and `QTN_effects_summary.txt`. Epistatic sets are expanded
one row per member, sharing a `set` number and effect:

```{r inspect-qtn}
qtn_table(ph_i)
```

Everything else the log recorded is a field on the object:

```{r inspect-fields}
ph_i$seed
ph_i$n_traits
ph_i$var_budget
```

Note that a `vqtl()` layer is an explicit residual-heterogeneity variance share,
not part of the broad-sense `h2` budget. It varies the conditional residual
variance rather than adding genetic value; `print()` says so when one is present:

```{r inspect-vqtl}
simulate_phenotype(geno, h2 = 0.5, seed = 61) |>
  additive(prop = 0.3, n_qtn = 4) |>
  vqtl(prop = 0.2)
```

`plot()` gives a four-panel diagnostic summary -- variance partition, phenotype
distribution, QTN effects, and (for two or more traits) the genetic-value
scatter:

```{r inspect-plot, fig.width = 8, fig.height = 6}
ph_plot <- simulate_phenotype(geno, n_traits = 2, h2 = 0.5, seed = 62) |>
  additive(prop = 0.4, n_qtn = 8) |>
  dominance(prop = 0.1)
plot(ph_plot)
```

# 7. Output

## In memory

```{r out-memory}
ph3 <- simulate_phenotype(geno, n_traits = 2, seed = 50) |>
  additive(prop = 0.5, n_qtn = 3)

head(phenotypes_long(ph3))
head(phenotypes_wide(ph3))
```

## Written to disk

Each call writes exactly the file you name — nothing else is created alongside
it.

```{r out-disk}
out_dir <- file.path(tempdir(), "sp_demo")
dir.create(out_dir, showWarnings = FALSE)

write_phenotypes(ph3, file = file.path(out_dir, "phenotypes_long.txt"))
write_phenotypes(ph3, file = file.path(out_dir, "phenotypes_wide.csv"),
                 format = "wide", sep = ",")

list.files(out_dir)
head(read.delim(file.path(out_dir, "phenotypes_long.txt")))
```

# 8. Original implementation of create_phenotypes

`create_phenotypes()` is the original interface, kept frozen so published
results stay reproducible. It receives bug fixes only; new work should use the
grammar above.

Its output always lands in a folder of its own, so a directory is never
littered with loose files:

```{r legacy}
legacy_home <- file.path(tempdir(), "sp_legacy")
dir.create(legacy_home, showWarnings = FALSE)

invisible(create_phenotypes(
  geno_obj = geno, add_QTN_num = 3, add_effect = 0.2,
  rep = 2, h2 = 0.5, model = "A",
  home_dir = legacy_home, verbose = FALSE
))

list.files(legacy_home)                                  # one folder
list.files(file.path(legacy_home, "simplePHENOTYPES_output"))
```

Naming the folder yourself, and re-running without overwriting:

```{r legacy-outdir}
invisible(create_phenotypes(
  geno_obj = geno, add_QTN_num = 3, add_effect = 0.2,
  rep = 1, h2 = 0.5, model = "A",
  home_dir = legacy_home, output_dir = "my_run", verbose = FALSE
))
invisible(create_phenotypes(
  geno_obj = geno, add_QTN_num = 3, add_effect = 0.2,
  rep = 1, h2 = 0.5, model = "A",
  home_dir = legacy_home, output_dir = "my_run", verbose = FALSE
))

list.files(legacy_home)      # my_run and my_run(1): nothing is overwritten
```

```{r cleanup, include = FALSE}
unlink(c(out_dir, legacy_home), recursive = TRUE)
```

# Session information

```{r session}
sessionInfo()
```
