---
title: "Genetic maps and recombination"
author: "simplePHENOTYPES"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Genetic maps and recombination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#", fig.width = 6, fig.height = 3.5
)
```

```{r setup}
library(simplePHENOTYPES)
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04
```

The crossing functions — `cross()`, `selfcross()` and `double_haploid()` — need
to know how often two markers recombine. This vignette explains what they need,
how `synthetic_map()` supplies it when you have nothing better, and how to
replace it with a real map.

# Why base pairs are not enough

Marker files almost always carry *physical* positions in base pairs. Meiosis
does not care about those. What matters is *genetic* distance in centiMorgans:
1 cM means a 1% chance of a crossover between two loci per generation.

The two are not proportional. Recombination is heavily suppressed near the
centromere and elevated toward the chromosome ends, so two markers 10 Mb apart
can be tightly linked in one region and freely recombining in another.

simplePHENOTYPES keeps the genetic map in the `cm` column, alongside `pos`:

```{r columns}
head(geno[, 1:5])
```

If `cm` is missing, `as_population()` stops rather than guessing:

```{r no-map, error = TRUE}
no_map <- geno
no_map$cm <- NA
as_population(no_map, individuals = c("33-16", "38-11"))
```

# `synthetic_map()`

`synthetic_map()` turns physical positions into a plausible centiMorgan map.

```{r basic}
cm <- synthetic_map(geno$chr, geno$pos)
aggregate(cm ~ chr, data = data.frame(chr = geno$chr, cm = cm), FUN = max)
```

**This is a model, not a measurement.** It is not derived from any published
linkage map, and its values should not be reported as estimated recombination
distances. Its purpose is to give a simulation a realistic recombination
landscape when no real map is at hand. If you have a real map, use it (see
below).

## What the model does

Each marker gets a local recombination rate that dips around the centromere:

$$r(p) = 1 - s \exp\left(-\tfrac{1}{2}\left(\frac{p - c}{wL}\right)^2\right)$$

with centromere `c`, chromosome span `L`, suppression strength `s` and relative
width `w`. That rate is integrated along the chromosome and rescaled to the
target length, so cM always increases with bp.

The consequence is the flat middle section of the curve — a large stretch of
physical genome that contributes very little genetic distance:

```{r shape}
on1 <- geno$chr == 1
plot(geno$pos[on1] / 1e6, geno$cm[on1], type = "l", lwd = 2,
     xlab = "Physical position, chromosome 1 (Mb)", ylab = "Genetic position (cM)")
lines(geno$pos[on1] / 1e6, synthetic_map(geno$chr, geno$pos, suppression = 0)[on1],
      lwd = 2, lty = 2, col = "grey50")
legend("topleft", c("centromere-suppressed", "uniform"),
       lwd = 2, lty = c(1, 2), col = c("black", "grey50"), bty = "n")
```

Local rate, in cM per Mb, makes the same point more directly:

```{r local-rate}
pos1 <- geno$pos[on1] / 1e6
cm1  <- geno$cm[on1]
bin  <- cut(pos1, breaks = seq(0, 310, by = 20), labels = FALSE)

along <- data.frame(pos = pos1, cm = cm1, bin = bin)
span  <- aggregate(cbind(cm, pos) ~ bin, data = along,
                   FUN = function(x) diff(range(x)))
span$cM_per_Mb <- round(span$cm / span$pos, 2)

span[, c("bin", "cM_per_Mb")]
```

Rates drop to roughly 0.2 cM/Mb in the middle against about 1 cM/Mb on the arms.

## Tuning it

| Argument | Effect |
|---|---|
| `cm_per_mb` | overall scale; default 0.73 gives ~1,500 cM for this genome |
| `total_cm` | set chromosome lengths directly instead of by rate |
| `suppression` | 0 = uniform map, closer to 1 = stronger centromeric suppression |
| `width` | how much of the chromosome the suppressed region covers |
| `centromere` | centromere positions in bp; defaults to each chromosome's midpoint |

```{r tuning}
# A uniform map: cM strictly proportional to bp
flat <- synthetic_map(geno$chr, geno$pos, suppression = 0)

# Fixed lengths, one per chromosome
fixed <- synthetic_map(geno$chr, geno$pos, total_cm = 150)
aggregate(cm ~ chr, data = data.frame(chr = geno$chr, cm = fixed), FUN = max)

# Real centromere positions, if you know them
limits <- aggregate(pos ~ chr, data = geno,
                    FUN = function(p) min(p) + 0.45 * diff(range(p)))
tuned <- synthetic_map(geno$chr, geno$pos, centromere = limits$pos)
```

# Using a real map

If you have measured genetic positions, use them — just put them in `cm`, in
centiMorgans, sorted within each chromosome:

```{r real-map, eval = FALSE}
geno$cm <- my_map$cM[match(geno$snp, my_map$marker)]
```

The map must satisfy three conditions, which `as_population()` checks:

1. no missing values,
2. non-decreasing within each chromosome,
3. markers ordered by position.

Co-located markers may share a cM value; that correctly means no recombination
between them.

```{r validate}
sorted <- aggregate(cm ~ chr, data = geno, FUN = function(x) !is.unsorted(x))
ok <- all(!is.na(geno$cm)) && all(sorted$cm)
ok
```

# How the map changes results

The map is not decoration — it determines how much recombination happens. The
number of crossovers on a chromosome is Poisson with mean equal to its length in
Morgans, so a 150 cM chromosome averages 1.5 crossovers per meiosis.

Compare linkage under the default map with linkage under a map ten times
shorter, where crossovers are rare and markers stay linked much further:

```{r compare}
long_map  <- geno
short_map <- geno
short_map$cm <- geno$cm / 10

linkage <- function(g, seed) {
  pop <- as_population(g, individuals = c("33-16", "38-11"))
  f1  <- cross(pop[1], pop[2], n = 1, seed = seed)
  dh  <- double_haploid(f1, n = 100, seed = seed + 1)

  on1 <- g$chr == 1
  p1 <- dosages(pop[1])[on1, 1]; p2 <- dosages(pop[2])[on1, 1]
  inf <- which(p1 != p2 & p1 != 0 & p2 != 0)
  from_p1 <- dosages(dh)[on1, ][inf, ] == p1[inf]
  # recombination between the first informative locus and one ~100 loci away
  mean(from_p1[100, ] != from_p1[1, ])
}

c(default = linkage(long_map, 1), ten_times_shorter = linkage(short_map, 1))
```

Shortening the map keeps the same markers linked far more tightly. If your
simulated population shows implausibly long haplotype blocks, or breaks up far
faster than your species should, the map is the first thing to check.

# Session information

```{r session}
sessionInfo()
```
