---
title: "The simplePHENOTYPES v2 grammar"
subtitle: Samuel B Fernandes and Alexander E Lipka
author: "simplePHENOTYPES (`r Sys.Date()`)"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: spacelab
    highlight: textmate

vignette: >
  %\VignetteIndexEntry{The simplePHENOTYPES v2 grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.width = 7,
  fig.height = 4
)
set.seed(1)
```

```{r setup}
library(simplePHENOTYPES)
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04
```

This vignette walks through every user-facing function in simplePHENOTYPES v2,
using the bundled maize 282 panel throughout. The v1 function
`create_phenotypes()` is covered at the end: it is still supported, but it is
frozen, and new work should use the grammar described here.

# The example data

`SNP55K_maize282_maf04` is 10,650 markers scored on 280 inbred lines, coded
-1 / 0 / 1. The first five columns are marker metadata.

```{r data}
dim(geno)
geno[1:5, 1:8]
```

# The genetic map

`cm` holds a genetic map in centiMorgans. Earlier versions of the package left
this column empty, which meant nothing in the package could simulate meiosis.

```{r map}
tapply(geno$cm, geno$chr, max)
```

**This map is synthetic.** It is generated from the physical positions by
`synthetic_map()` and is not taken from any published maize linkage map, so its
cM values should not be quoted as measured recombination distances. It exists so
the crossing functions have realistic distances to work with. If you have a real
map for your population, use that instead — just put it in `cm`.

Recombination is not spread evenly along a chromosome. In maize it is strongly
suppressed around the centromere, so a large share of the physical genome
contributes very little genetic distance. `synthetic_map()` models that, and it
is visible as the flat middle section of each curve:

```{r map-shape}
plot(geno$pos[geno$chr == 1] / 1e6, geno$cm[geno$chr == 1],
     type = "l", lwd = 2,
     xlab = "Physical position on chromosome 1 (Mb)",
     ylab = "Genetic position (cM)")
```

To build a map for your own data, or to compare against a uniform one:

```{r synthetic-map}
uniform <- synthetic_map(geno$chr, geno$pos, suppression = 0)
head(round(uniform, 2))
```

# Reading genotype data

`as_numeric()` converts a genotype file into the numeric format used above,
handling HapMap, VCF, GDS and PLINK bed/ped input.

```{r as-numeric, eval = FALSE}
num <- as_numeric("my_genotypes.hmp.txt")
```

`format_conversion()` converts between formats without simulating anything:

```{r format-conversion, eval = FALSE}
format_conversion(file = "my_genotypes.vcf", to = "numeric", to_r = TRUE)
```

# Simulating phenotypes: the grammar

A simulation starts with `simulate_phenotype()`, which fixes the genetic
architecture, and then adds variance components as layers. Each layer's `prop`
is a proportion of total phenotypic variance, so heritability is the sum of the
genetic proportions and the residual is whatever is left.

```{r single-trait}
ph <- simulate_phenotype(geno, seed = 1) |>
  additive(prop = 0.5, n_qtn = 3)
ph
```

The pipe runs eagerly — there is no terminal `simulate()` call, and every object
already carries realized phenotypes.

```{r long-output}
head(phenotypes_long(ph))
```

## Dominance, epistasis and variance QTL

The other three layers work the same way. By default `dominance()` and `vqtl()`
reuse the additive QTNs; set `same_as_add = FALSE` to draw fresh ones.

```{r layers}
ph_ade <- simulate_phenotype(geno, seed = 2) |>
  additive(prop = 0.4, n_qtn = 5) |>
  dominance(prop = 0.1, degree = 0.5) |>
  epistasis(prop = 0.1, n_pairs = 2)
ph_ade
```

```{r vqtl}
ph_v <- simulate_phenotype(geno, seed = 3) |>
  additive(prop = 0.3, n_qtn = 4) |>
  vqtl(prop = 0.2)
ph_v
```

## Multiple traits and genetic correlation

`architecture = "pleiotropy"` puts the same QTNs behind several traits. With two
traits, `cor` sets the **genetic** correlation through the PleioArch algorithm,
which draws the pleiotropic QTN effects from a bivariate normal.

```{r pleiotropy}
ph_pleio <- simulate_phenotype(geno, architecture = "pleiotropy",
                               n_traits = 2, cor = 0.6, seed = 4) |>
  additive(prop = 0.5, n_qtn = 100)
ph_pleio

wide <- phenotypes_wide(ph_pleio)
round(cor(wide$Trait_1, wide$Trait_2), 2)
```

Two things are worth knowing before you compare that number with `cor`.

*It is measured on the wrong scale.* `cor` is the genetic correlation, but what
you observe above is the phenotypic one, and the residuals are drawn
independently. With uncorrelated residuals the two are related by
$r_P \approx r_G \sqrt{h^2_1 h^2_2}$, so a genetic correlation of 0.6 between
traits with $h^2 = 0.5$ shows up as a phenotypic correlation near 0.3.

*It is realized, not fixed.* `cor` is achieved in expectation over repeated
simulations, not exactly in any one of them. The QTN effects are random draws,
so the realized correlation scatters around the target — and with few QTNs that
scatter is wide. Averaged over 12 seeds the realized genetic correlation for a
target of 0.6 comes out near 0.6 either way, but the spread across individual
runs has a standard deviation of roughly 0.4 at `n_qtn = 10`, against roughly
0.16 at `n_qtn = 200`. If you need a tight correlation in a particular data set,
use many QTNs, and check what you actually got rather than assuming.

With more than two traits the exact algorithm does not apply and the package
falls back to a Cholesky decomposition, warning that the phenotypes will be
correlated as requested but individual effect sizes are not guaranteed.

## Linked, not causal: the "ld" architecture

`architecture = "ld"` places the causal variant near, but not on, the marker
that a study would actually test — the spurious-pleiotropy scenario.

```{r ld}
ph_ld <- simulate_phenotype(geno, architecture = "ld", n_traits = 2,
                            ld_type = "indirect", seed = 5) |>
  additive(prop = 0.4, n_qtn = 3)
ph_ld
```

## Combining architectures

`complex_phenotypes()` merges complete models under a common heritability,
weighting them by their genetic variances. This is how partial pleiotropy is
built: combine a pleiotropic model with an independent one.

```{r complex}
pleio <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 2,
                            cor = 0.5, seed = 6) |>
  additive(prop = 0.4, n_qtn = 5)
indep <- simulate_phenotype(geno, n_traits = 2, seed = 6) |>
  additive(prop = 0.3, n_qtn = 5)

both <- complex_phenotypes(pleio, indep, h2 = 0.5)
both
```

## One call instead of a pipe

Supplying `h2` together with `n_qtn` realizes a complete model in a single call.

```{r one-call}
simulate_phenotype(geno, h2 = 0.6, n_qtn = 5, model = "AD", seed = 7)
```

# Writing results

Phenotypes come back as an object; the exporters write them to disk in the
format a downstream tool expects.

```{r exporters, eval = FALSE}
write_phenotypes(ph, format = "long", dir = tempdir())
write_phenotypes(ph, format = "gemma", dir = tempdir())
```

```{r wide}
head(phenotypes_wide(ph_pleio))
```

# Multi-generation crosses

The crossing functions simulate meiosis directly, so they can build a pedigree
from real founder genotypes. Start by phasing the founders into a `Population`:

```{r population}
pop <- as_population(geno, individuals = c("33-16", "38-11"))
pop
```

Heterozygotes have to be phased arbitrarily, which is nearly lossless for an
inbred panel like this one but a real assumption for outbred material — see
`?as_population`.

## Making an F1, an F2 and doubled haploids

```{r pedigree}
f1 <- cross(pop[1], pop[2], n = 1, seed = 11)
f2 <- selfcross(f1, n = 200, seed = 12)
dh <- double_haploid(f1, n = 400, seed = 13)
f2
```

The result is textbook genetics. At markers where the two founders differ, the
F1 is heterozygous, the F2 segregates 1:2:1, and the doubled haploids are
completely homozygous and split 1:1:

```{r segregation}
het_in_f1 <- which(dosages(f1)[, 1] == 0)

round(prop.table(table(dosages(f2)[het_in_f1, ])), 3)
round(prop.table(table(dosages(dh)[het_in_f1, ])), 3)
any(dosages(dh) == 0)
```

Recombination follows the genetic map. To see it, measure which *parent* each
doubled haploid inherited a locus from, rather than comparing genotypes
directly — the two founders happen to share an allele at many markers, and
those markers carry no information about recombination. Restricting to loci
where the F1 is heterozygous, and asking whether a locus came from the same
homologue as the first locus, recovers the recombination fraction:

```{r linkage}
on_chr1 <- geno$chr == 1
p1 <- dosages(pop[1])[on_chr1, 1]
p2 <- dosages(pop[2])[on_chr1, 1]

# Informative loci: both founders homozygous, and for different alleles.
inf <- which(p1 != p2 & p1 != 0 & p2 != 0)
cm1 <- geno$cm[on_chr1][inf]

# A doubled haploid is homozygous, so its allele names the founder it came from.
from_p1 <- dosages(dh)[on_chr1, ][inf, ] == p1[inf]

pick <- c(5, 20, 100, 400)
data.frame(
  distance_cM   = round(cm1[pick] - cm1[1], 1),
  recombination = round(rowMeans(from_p1[pick, ] != from_p1[rep(1, 4), ]), 3),
  haldane       = round((1 - exp(-2 * (cm1[pick] - cm1[1]) / 100)) / 2, 3)
)
```

The observed fractions track Haldane's map function, which is what the
count-location model implies: crossovers arrive as a Poisson process along the
chromosome with no interference, so loci recombine with probability
$(1 - e^{-2d})/2$ at distance $d$ Morgans, approaching one half — never
exceeding it — as they become unlinked.

All randomness is drawn by R, so `set.seed()` or the `seed =` argument makes any
pedigree exactly reproducible.

## Phenotyping a simulated pedigree

A `Population` is accepted anywhere `simulate_phenotype()` takes `geno`, so a
pedigree can be phenotyped without any conversion:

```{r phenotype-population}
ph_f2 <- simulate_phenotype(f2, seed = 14) |>
  additive(prop = 0.6, n_qtn = 5)
ph_f2
head(phenotypes_long(ph_f2))
```

# The v1 interface

`create_phenotypes()` is the original v1 function. It is **frozen**: it keeps
its original code paths and seed handling so that existing scripts keep
reproducing their published results, and it receives bug fixes only. New work
should use the grammar above, which is why it is documented as superseded.

```{r legacy, eval = FALSE}
create_phenotypes(
  geno_obj = geno,
  add_QTN_num = 3,
  add_effect = 0.2,
  rep = 10,
  h2 = 0.7,
  model = "A",
  to_r = TRUE,
  home_dir = tempdir()
)
```

# Session information

```{r session}
sessionInfo()
```
