---
title: "Migrating from v1 create_phenotypes() to the v2 grammar"
subtitle: Samuel B Fernandes and Alexander E Lipka
author: "simplePHENOTYPES (`r Sys.Date()`)"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: spacelab
    highlight: textmate

vignette: >
  %\VignetteIndexEntry{Migrating from v1 create_phenotypes() to the v2 grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(simplePHENOTYPES)
data("SNP55K_maize282_maf04")
geno <- SNP55K_maize282_maf04
```

This vignette re-creates every example from the original (v1) README / vignette
using the v2 grammar, so a v1 user can see the direct translation. The legacy
`create_phenotypes()` is **still shipped** (frozen, bug-fix only) if you need
bit-for-bit continuity with old scripts; everything below shows the *new* way.

Two structural differences to keep in mind:

- **Heritability emerges from variance proportions.** In v1 you set `h2` and
  effect sizes; in v2 each layer takes a `prop` (share of phenotypic variance)
  and the trait's h² is the sum of those proportions. You can still pass a target
  `h2` to `simulate_phenotype()` for the one-call shortcut.
- **`big_add_QTN_effect` is gone from the grammar** (it lives only in the frozen
  `create_phenotypes()`). Give a large-effect QTN an explicit `effect` instead.

# Single trait

**v1**

```{r v1-single, eval = FALSE}
create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  add_QTN_num = 3,
  add_effect = 0.2,
  big_add_QTN_effect = 0.9,
  rep = 10,
  h2 = 0.7,
  model = "A",
  home_dir = tempdir())
```

**v2** — an additive layer taking 70% of the variance over three QTNs. The
one-call form (`h2` + `n_qtn`) is the closest analogue:

```{r v2-single}
ph <- simulate_phenotype(geno, h2 = 0.7, n_qtn = 3, n_reps = 10, seed = 1)
ph
```

To reproduce the v1 "one big QTN plus a geometric tail", give the effects
explicitly instead of `big_add_QTN_effect`:

```{r v2-single-big}
simulate_phenotype(geno, seed = 1) |>
  additive(prop = 0.7, n_qtn = 3, effect = c(0.9, 0.2, 0.2^2))
```

# Pleiotropy

In v1, three pleiotropic traits shared QTNs and the genetic correlation was an
*uncontrolled* by-product of the per-trait effect sizes. **v2 lets you set the
genetic correlation directly** (the PleioArch engine), which is the main reason
to prefer the grammar here.

**v1**

```{r v1-pleio, eval = FALSE}
create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  add_QTN_num = 3, dom_QTN_num = 4,
  big_add_QTN_effect = c(0.3, 0.3, 0.3),
  h2 = c(0.2, 0.4, 0.4),
  add_effect = c(0.04, 0.2, 0.1),
  dom_effect = c(0.04, 0.2, 0.1),
  ntraits = 3, rep = 10, architecture = "pleiotropic",
  model = "AD", seed = 10, to_r = TRUE)
```

**v2** — additive + dominance on the same shared QTNs, three traits. Without
`cor` the correlation is emergent (as in v1); with `cor` it is realized on
target:

```{r v2-pleio}
ph <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 3,
                         cor = 0.5, seed = 10) |>
  additive(prop = 0.3, n_qtn = 3) |>
  dominance(prop = 0.1, same_as_add = TRUE)
round(cor(genetic_values(ph)), 2)
```

`cor` also accepts a full matrix for asymmetric targets:

```{r v2-pleio-matrix}
R <- matrix(c(1, 0.3, -0.5,
              0.3, 1, -0.2,
              -0.5, -0.2, 1), 3)
simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 3,
                   cor = R, seed = 10) |>
  additive(prop = 0.4, n_qtn = 100) |>
  genetic_values() |>
  cor() |>
  round(2)
```

# Partial pleiotropy

v1 had a dedicated `architecture = "partially"`. **v2 builds it by combining a
pleiotropic model (shared loci) with an independent model (trait-specific loci)
via `complex_phenotypes()`** — which is exactly what "partial pleiotropy" means:
some loci are shared, some are trait-specific.

**v1**

```{r v1-partial, eval = FALSE}
cor_matrix <- matrix(c(1, 0.3, -0.9,
                       0.3, 1, -0.5,
                       -0.9, -0.5, 1), 3)
create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  ntraits = 3, pleio_a = 3,
  trait_spec_a_QTN_num = c(4, 10, 1),
  h2 = c(0.2, 0.4, 0.8),
  cor = cor_matrix,
  architecture = "partially", model = "A", rep = 20, to_r = TRUE)
```

**v2** — the shared (pleiotropic) component carries the correlation; the
independent component adds trait-specific loci; `complex_phenotypes()` merges
them under a common heritability:

```{r v2-partial}
cor_matrix <- matrix(c(1, 0.3, -0.9,
                       0.3, 1, -0.5,
                       -0.9, -0.5, 1), 3)

shared <- simulate_phenotype(geno, architecture = "pleiotropy", n_traits = 3,
                             cor = cor_matrix, seed = 10) |>
  additive(prop = 0.5, n_qtn = 3)

specific <- simulate_phenotype(geno, architecture = "independent", n_traits = 3,
                               seed = 10) |>
  additive(prop = 0.5, n_qtn = 5)

partial <- complex_phenotypes(shared, specific, h2 = c(0.2, 0.4, 0.8))
partial
```

The realized correlation is now driven by the shared loci while each trait still
carries its own private QTNs — the definition of partial pleiotropy. (A grammar
layer uses one QTN count per call, so the v1 per-trait `trait_spec_a_QTN_num =
c(4, 10, 1)` becomes a single `n_qtn` here; if you need different trait-specific
counts, build one independent model per trait and combine them.)

# Spurious pleiotropy (LD)

v1's `architecture = "LD"` with `type_of_ld` maps directly onto v2's
`architecture = "ld"`. The two traits get **distinct** causal loci that sit in
linkage disequilibrium; `ld_type = "direct"` (the v2 default) links the two
traits' causal SNPs to each other, `"indirect"` links each to a shared hidden
cause-of-LD locus. `ld_max`/`ld_min` become `r2_max`/`r2_min`.

**v1**

```{r v1-ld, eval = FALSE}
create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  add_QTN_num = 3,
  h2 = c(0.2, 0.4),
  add_effect = c(0.02, 0.05),
  rep = 5, seed = 200, architecture = "LD",
  ld_max = 0.8, ld_min = 0.2, type_of_ld = "indirect", model = "A")
```

**v2**

```{r v2-ld}
ph_ld <- simulate_phenotype(geno, architecture = "ld", n_traits = 2,
                            ld_type = "indirect", r2_max = 0.8, r2_min = 0.2,
                            seed = 200) |>
  additive(prop = c(0.2, 0.4), n_qtn = 3)

# distinct causal loci per trait: QTN_t1 / QTN_t2 are the linked pair
qtn_table(ph_ld)[, c("trait", "snp", "QTN_t1", "QTN_t2", "ld_r2")]
```

# Choosing the QTNs yourself

v1 took a `QTN_list`; in v2 every layer accepts a `qtn` argument (marker names or
indices). The other layers stay random.

**v1**

```{r v1-qtnlist, eval = FALSE}
QTN_list <- list()
QTN_list$add[[1]] <- c("ss196523212")
QTN_list$dom[[1]] <- c("ss196510214", "ss196472187")
QTN_list$epi[[1]] <- c("ss196530605", "ss196475446")
create_phenotypes(
  geno_obj = SNP55K_maize282_maf04,
  add_QTN_num = 1, dom_QTN_num = 2, epi_QTN_num = 1, epi_interaction = 2,
  h2 = 0.9, QTN_list = QTN_list, ntraits = 1, model = "ADE")
```

**v2** — pass the loci to each layer; `epistasis(qtn=)` takes an
`n_pairs × interaction` matrix:

```{r v2-qtnlist}
simulate_phenotype(geno, seed = 1) |>
  additive(prop = 0.5, qtn = "ss196523212") |>
  dominance(prop = 0.2, qtn = c("ss196510214", "ss196472187")) |>
  epistasis(prop = 0.2,
            qtn = matrix(c("ss196530605", "ss196475446"), nrow = 1))
```

# Variance explained by each QTN

v1 used `QTN_variance = TRUE`. In v2 the per-QTN variance is always available
from `qtn_table()`:

```{r v2-varexp}
ph <- simulate_phenotype(geno, h2 = 0.6, seed = 1) |>
  additive(prop = 0.4, n_qtn = 5) |>
  dominance(prop = 0.2, same_as_add = TRUE)
qtn_table(ph)[, c("trait", "layer", "snp", "effect", "var_explained")]
```

# Output formats

v1 controlled output with `output_format` / `out_geno` / `to_r`. In v2 the object
is always returned, and the exporters are separate:

```{r v2-output, eval = FALSE}
ph <- simulate_phenotype(geno, h2 = 0.5, n_qtn = 3, n_traits = 2, seed = 1)

phenotypes_long(ph)                       # long (default), one row per id x trait x rep
phenotypes_wide(ph)                        # wide, one column per trait
write_phenotypes(ph, "pheno.txt")          # to disk (long)
write_phenotypes(ph, "pheno_wide.csv", format = "wide", sep = ",")
```

# Reading genotypes from files

v1 read files through `geno_obj` / `geno_file` / `geno_path`. In v2, convert to
the numeric object once with `as_numeric()` (HapMap / VCF / GDS / PLINK / numeric)
and pass it to `simulate_phenotype()`:

```{r v2-files, eval = FALSE}
num <- as_numeric("WGS_chrm_1.hmp.txt")    # or .vcf, .gds, .bed, ...
simulate_phenotype(num, h2 = 0.2, n_qtn = 3, seed = 200)
```

# Not yet in the grammar

A few v1 knobs do not have a grammar equivalent yet and remain only in the frozen
`create_phenotypes()`:

- **Residual correlation** (`cor_res`) — the grammar controls *genetic*
  correlation (`cor`); a residual-correlation argument is planned but not yet
  implemented.
- **`big_add_QTN_effect`** — intentionally dropped (use an explicit `effect`).
- **`architecture = "partially"`** as a single call — recreated with
  `complex_phenotypes()`, as shown above.
